---
title: "The Binomial Distribution"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

---
class: middle, center, inverse

# The Binomial Distribution

---

# Roadmap

- Motivation: Discrete probability distributions
- Bernoulli trials
- The Binomial distribution
  - Definition and parameters
  - Formula
  - Mean and variance
- Working with the Binomial in R
- Applications and examples

--

**By the end of this lecture, you should be able to:**

- Define a Bernoulli trial and identify when it applies
- State the conditions required for a Binomial distribution
- Calculate Binomial probabilities using the formula and R
- Compute the mean and variance of a Binomial random variable
- Apply the Binomial distribution to real-world problems

---

# Discrete Probability Distributions

- A **random variable** assigns a numerical value to each outcome of a random experiment
- A **discrete random variable** takes on a countable number of distinct values
- A **probability distribution** describes the probability of each possible value

--

- Examples of discrete random variables:
  - Number of heads in 10 coin flips
  - Number of defective items in a batch
  - Number of successes in a fixed number of trials

---

# Bernoulli Trials

A **Bernoulli trial** is a single experiment with exactly two possible outcomes:

- **Success** (with probability $p$)
- **Failure** (with probability $1 - p = q$)

--

**Examples:**
- Flipping a coin: heads = success, tails = failure
- Testing a component: works = success, fails = failure
- A free throw: makes it = success, misses = failure

--

**Key property:** Each trial is independent — the outcome of one trial does not affect the others

---

# The Bernoulli Distribution

If $X$ is the outcome of a single Bernoulli trial:

$$P(X = 1) = p \quad \text{(success)}$$
$$P(X = 0) = 1 - p \quad \text{(failure)}$$

--

**Mean:** $E(X) = p$

**Variance:** $Var(X) = p(1-p)$

---
class: middle, center, inverse

# The Binomial Distribution

---

# From Bernoulli to Binomial

- Suppose we repeat a Bernoulli trial $n$ times
- Let $X$ = the **total number of successes** in $n$ trials
- Then $X$ follows a **Binomial distribution**

--

$$X \sim \text{Binomial}(n, p)$$

- $n$ = number of trials (fixed)
- $p$ = probability of success on each trial

---

# Conditions for the Binomial Distribution

The Binomial distribution applies when:

1. There are a **fixed number** of trials, $n$
2. Each trial has exactly **two outcomes** (success or failure)
3. The probability of success $p$ is **constant** across all trials
4. The trials are **independent** of each other

--

**Memory aid:** FICS — Fixed, Independent, Constant, Success/Failure (two outcomes)

---

# The Binomial Formula

The probability of exactly $k$ successes in $n$ trials:

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$$

--

Where the **binomial coefficient** is:

$$\binom{n}{k} = \frac{n!}{k!(n-k)!}$$

- This counts the number of ways to choose $k$ successes from $n$ trials

---

# Understanding the Formula

$$P(X = k) = \underbrace{\binom{n}{k}}_{\text{# of arrangements}} \cdot \underbrace{p^k}_{\text{prob of }k\text{ successes}} \cdot \underbrace{(1-p)^{n-k}}_{\text{prob of }n-k\text{ failures}}$$

--

**Example:** Flip a fair coin 3 times. What is the probability of exactly 2 heads?

$$P(X = 2) = \binom{3}{2} (0.5)^2 (0.5)^1 = 3 \cdot 0.25 \cdot 0.5 = 0.375$$

---

# Worked Example: Coin Flips

**Problem:** Flip a fair coin 3 times. Find the full probability distribution of $X$ = number of heads.

--

| $k$ | $\binom{3}{k}$ | $p^k$ | $(1-p)^{3-k}$ | $P(X=k)$ |
|-----|----------------|--------|----------------|-----------|
| 0   | 1              | 1      | 0.125          | 0.125     |
| 1   | 3              | 0.5    | 0.25           | 0.375     |
| 2   | 3              | 0.25   | 0.5            | 0.375     |
| 3   | 1              | 0.125  | 1              | 0.125     |

--

All probabilities sum to 1: $0.125 + 0.375 + 0.375 + 0.125 = 1$ `r emo::ji("check")`

---

# Mean and Variance of the Binomial

For $X \sim \text{Binomial}(n, p)$:

$$E(X) = np$$

$$Var(X) = np(1-p)$$

$$SD(X) = \sqrt{np(1-p)}$$

--

**Intuition:** If you flip a fair coin 100 times, you expect $np = 100 \times 0.5 = 50$ heads.

---

# Mean and Variance: Example

**Problem:** You take a 20-question multiple choice test. Each question has 4 options and you guess randomly on every question.

- $n = 20$, $p = 0.25$

--

**Mean:** $E(X) = np = 20 \times 0.25 = 5$ correct answers

**Variance:** $Var(X) = np(1-p) = 20 \times 0.25 \times 0.75 = 3.75$

**SD:** $SD(X) = \sqrt{3.75} \approx 1.94$

---
class: middle, center, inverse

# The Binomial in R

---

# R Functions for the Binomial

R provides four functions for working with the Binomial distribution:

| Function | Description | Usage |
|----------|-------------|-------|
| `dbinom(k, n, p)` | $P(X = k)$ — exact probability | Point probability |
| `pbinom(k, n, p)` | $P(X \leq k)$ — cumulative probability | CDF |
| `qbinom(prob, n, p)` | Quantile (inverse CDF) | Finding cutoffs |
| `rbinom(m, n, p)` | Generate $m$ random values | Simulation |

---

# Using `dbinom()`

**Exact probability:** $P(X = k)$

```r
# P(X = 2) when n = 3, p = 0.5
dbinom(2, size = 3, prob = 0.5)
```
```
[1] 0.375
```

--

```r
# Full distribution: P(X = 0, 1, 2, 3)
dbinom(0:3, size = 3, prob = 0.5)
```
```
[1] 0.125 0.375 0.375 0.125
```

---

# Using `pbinom()`

**Cumulative probability:** $P(X \leq k)$

```r
# P(X <= 2) when n = 3, p = 0.5
pbinom(2, size = 3, prob = 0.5)
```
```
[1] 0.875
```

--

**Upper tail:** $P(X > k) = 1 - P(X \leq k)$

```r
# P(X > 2) when n = 3, p = 0.5
pbinom(2, size = 3, prob = 0.5, lower.tail = FALSE)
```
```
[1] 0.125
```

---

# Using `pbinom()`: Between Two Values

**Between two values:** $P(a \leq X \leq b) = P(X \leq b) - P(X \leq a - 1)$

```r
# P(1 <= X <= 3) when n = 5, p = 0.4
pbinom(3, size = 5, prob = 0.4) - pbinom(0, size = 5, prob = 0.4)
```

--

```r
# Or equivalently:
sum(dbinom(1:3, size = 5, prob = 0.4))
```

---

# Visualizing the Binomial Distribution

```{r binomial-plot, echo=FALSE, fig.height=4}
library(tidyverse)
tibble(
  k = 0:20,
  prob = dbinom(0:20, size = 20, prob = 0.4)
) %>%
  ggplot(aes(x = k, y = prob)) +
  geom_col(fill = "#003366", alpha = 0.8) +
  labs(
    title = "Binomial Distribution: n = 20, p = 0.4",
    x = "Number of Successes (k)",
    y = "P(X = k)"
  ) +
  theme_minimal()
```

---

# Shape of the Binomial Distribution

The shape depends on $p$:

- $p < 0.5$: right-skewed (skewed toward higher values)
- $p = 0.5$: symmetric
- $p > 0.5$: left-skewed (skewed toward lower values)

--

As $n$ increases, the Binomial distribution becomes more bell-shaped (approaches the Normal distribution).

<!-- TODO: add comparison plots showing skew for different p values -->

---
class: middle, center, inverse

# Applications

---

# Application 1: Quality Control

**Problem:** A manufacturing process produces defective items 3% of the time. A batch of 50 items is sampled.

- $X$ = number of defective items
- $X \sim \text{Binomial}(50, 0.03)$

--

**What is the probability of finding more than 3 defectives?**

```r
pbinom(3, size = 50, prob = 0.03, lower.tail = FALSE)
```

--

```
[1] 0.0573
```

About a 5.7% chance.

---

# Application 2: Drug Efficacy

**Problem:** A new drug is effective in 70% of patients. A clinical trial tests 15 patients.

- $X \sim \text{Binomial}(15, 0.70)$

--

**Expected number of patients who respond:**
$$E(X) = 15 \times 0.70 = 10.5$$

**Probability all 15 respond:**
```r
dbinom(15, size = 15, prob = 0.70)
```
```
[1] 0.00475
```

---

# Application 3: Free Throws

**Problem:** A basketball player makes 80% of free throws. She shoots 10 free throws.

- $X \sim \text{Binomial}(10, 0.80)$

--

.pull-left[
**P(exactly 8 made):**
```r
dbinom(8, size = 10, prob = 0.80)
```
```
[1] 0.302
```
]

.pull-right[
**P(at least 9 made):**
```r
pbinom(8, size = 10, prob = 0.80,
       lower.tail = FALSE)
```
```
[1] 0.376
```
]

---

# Simulation with `rbinom()`

We can simulate Binomial experiments in R:

```r
# Simulate 1000 repetitions of flipping a fair coin 10 times
set.seed(123)
sims <- rbinom(1000, size = 10, prob = 0.5)

# Summarize
mean(sims)   # should be close to np = 5
var(sims)    # should be close to np(1-p) = 2.5
```

--

```
[1] 5.002   # mean
[1] 2.524   # variance
```

The simulation results closely match the theoretical values. `r emo::ji("tada")`

---

# From Binomial to Normal

When $n$ is large and $p$ is not too close to 0 or 1:

$$X \sim \text{Binomial}(n, p) \approx \text{Normal}(np, \sqrt{np(1-p)})$$

--

**Rule of thumb:** The approximation works well when:
- $np \geq 10$
- $n(1-p) \geq 10$

--

This is the foundation for many statistical inference procedures (e.g., tests for proportions).

---

# Summary: Key Formulas

| Quantity | Formula |
|----------|---------|
| PMF | $P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$ |
| Mean | $E(X) = np$ |
| Variance | $Var(X) = np(1-p)$ |
| SD | $SD(X) = \sqrt{np(1-p)}$ |

--

**R functions:**
- `dbinom(k, n, p)` — exact probability
- `pbinom(k, n, p)` — cumulative probability
- `qbinom(prob, n, p)` — quantile
- `rbinom(m, n, p)` — random samples

---

# Common Mistakes to Avoid

- **Forgetting independence:** Binomial requires independent trials
- **Non-constant $p$:** If $p$ changes across trials, it's not Binomial
- **Off-by-one errors:** $P(X \geq k)$ vs. $P(X > k)$
  - $P(X \geq k) = 1 - P(X \leq k - 1)$ = `pbinom(k-1, n, p, lower.tail=FALSE)` ... actually = `pbinom(k-1, n, p, lower.tail=FALSE)`
  - $P(X > k) = 1 - P(X \leq k)$ = `pbinom(k, n, p, lower.tail=FALSE)`
- **Confusing `size` and `prob`** in R function arguments

---
class: middle, center, inverse

# Wrapping Up...

---

# What We Covered

- **Bernoulli trials**: single binary outcome experiments
- **Binomial distribution**: counts successes across $n$ independent Bernoulli trials
- **Formula**: $P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$
- **Mean**: $E(X) = np$; **Variance**: $Var(X) = np(1-p)$
- **R functions**: `dbinom()`, `pbinom()`, `qbinom()`, `rbinom()`
- **Applications**: quality control, medicine, sports, and more

--

The Binomial distribution is one of the most important discrete distributions — it underpins proportion-based inference throughout statistics.
