---
title: "R basics"
author: "S. Mason Garrison"
bibliography: library.bib
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
#       slideNumberFormat: ""  
library(tidyverse)
library(vembedr)
library(knitr)
if (!require("emo")) devtools::install_github("hadley/emo")
# Installs library if missing
if (!require("HistData")) install.packages("HistData") 
library(emo)
knitr::opts_chunk$set(echo = FALSE,out.width = "90%", fig.align = "center")

```

---

# More Accessible Resources

- [R Graph Catalog](http://shinyapps.stat.ubc.ca/r-graph-catalog/)
- [intRo](http://www.intro-stats.com/)
- .hand-pink[[Data Science for Psychologists](https://datascience4psych.github.io/DataScience4Psych/)]

---

class: center, middle

# Wrapping Up...

---

# R Basics

---

# R Basics

- R is a powerful language for statistical computing and graphics
- It's free, open-source, and has a large community of users and developers

---

# R Installation and Setup

- Download R from CRAN (Comprehensive R Archive Network): https://cran.r-project.org/
- Choose the version for your operating system
- RStudio is a popular IDE for R, available at: https://www.rstudio.com/products/rstudio/download/

---

# R Console

- After installation, you can start R or RStudio
- You'll see a console where you can enter commands
- Try a simple calculation:

```{r, echo=TRUE}
2 + 3

```

---

# Variable Assignment

- Use `<-` or `=` to assign values to variables

Variable names should be descriptive and use `under_scores` or `camelCase`

```{r, echo= TRUE}
x <- 10
y = 20
total_sum <- x + y
totalSum <- x + y
print(total_sum)

```

---

# Basic Data Types

- Numeric: 1.5, 2, 3.14
- Character: "hello", 'world'
- Logical: TRUE, FALSE

```{r}
num_var <- 1.5
char_var <- "hello"
log_var <- TRUE
```

---

# Vectors


- Create vectors using `c()` function
- Vectors can contain elements of the same type

```{r,echo=TRUE}

numeric_vector <- c(1, 2, 3, 4, 5)
character_vector <- c("a", "b", "c")
print(numeric_vector)

```

---

# Basic Plotting in R

.pull-left[
- R has built-in plotting functions
- These are useful for quick visualizations
]
.pull-right[
```{r, echo=TRUE}
x <- 1:10
y <- x^2
plot(x, y, main="Simple Plot", xlab="X axis", ylab="Y axis")
```
]


---

# Using Functions

.pull-left-narrow[
- R has many built-in functions
- You can also create your own functions
]
.pull-right-wide[
```{r, echo=TRUE}
# Built-in function
mean(c(1, 2, 3, 4, 5))

# Custom function
square <- function(x) {
  return(x^2)
}
square(4)

```
]

---

# Comments

- All text after the pound sign "#" within the same line is considered a comment.

> 1 + 1      # this is a comment 
[1] 2

---

# Getting Help

- R provides extensive documentation. 
- Use `?` or `help()` to get information about functions
- For example, entering `?c` or `help(c)` at the prompt gives documentation of the function `c` in R.
- Example:

```{r, echo=TRUE}
?mean
help(plot)
```


---

# Getting Help

> `help(c)`

If you are not sure about the name of the function you are looking for, you can perform a fuzzy search with the apropos function.

> `apropos("nova")`
[1] "anova"                "anova.glm" 
   ....



---

# Wrapping Up...






