---
title: "Review for Test 2"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Review for Test 2
### Data Science for Psychologists

---

## Roadmap

- Recap of Test 1 Topics
- New Material: Data Wrangling
- New Material: Data Visualization
- New Material: Probability & Distributions
- New Material: Inference & Hypothesis Testing
- Practice Problems

--

**By the end of this review, you should be able to:**

- Wrangle and reshape data using `dplyr` and `tidyr` verbs
- Create and interpret visualizations using `ggplot2`
- Understand probability distributions and calculate probabilities
- Apply hypothesis testing logic and interpret p-values
- Distinguish between Type I and Type II errors and understand statistical power

---

class: middle, center, inverse

# What's on the Test?

---

## Test 2 Coverage

.pull-left[
**Carried over from Test 1:**
- R basics & data types
- Reading in data
- Exploratory data analysis
- Data visualization basics
]

.pull-right[
**New material:**
- Data wrangling (`dplyr`, `tidyr`)
- Advanced `ggplot2`
- Probability & distributions
- Sampling distributions & CLT
- Hypothesis testing
- Confidence intervals
]

---

class: middle, center, inverse

# Data Wrangling

---

## Core `dplyr` Verbs

The six core `dplyr` verbs:

--

- `filter()` — keep rows that meet a condition

--

- `select()` — keep (or drop) columns by name

--

- `mutate()` — add or modify columns

--

- `arrange()` — sort rows

--

- `summarize()` — collapse rows to summary statistics

--

- `group_by()` — perform operations by group

---

## `filter()`

Keep rows where a condition is `TRUE`:

```{r echo=TRUE, eval=FALSE}
starwars %>%
  filter(species == "Human", height > 170)
```

--

Common operators:
- `==` equal to
- `!=` not equal to
- `>`, `<`, `>=`, `<=`
- `%in%` — value is in a vector
- `is.na()` — check for missing values
- `!` — negate

---

## `select()`

Keep or drop columns:

```{r echo=TRUE, eval=FALSE}
starwars %>%
  select(name, height, mass, species)

starwars %>%
  select(-films, -vehicles, -starships)
```

--

Helper functions:
- `starts_with("x")`
- `ends_with("x")`
- `contains("x")`
- `everything()`

---

## `mutate()`

Add or modify columns:

```{r echo=TRUE, eval=FALSE}
starwars %>%
  mutate(
    bmi = mass / (height / 100)^2,
    tall = height > 180
  )
```

--

- The new column is added to the right
- You can reference a column you just created in the same `mutate()` call
- Use `if_else()` or `case_when()` for conditional logic

---

## `case_when()`

A vectorized if-else for multiple conditions:

```{r echo=TRUE, eval=FALSE}
starwars %>%
  mutate(
    size = case_when(
      height < 100 ~ "small",
      height < 180 ~ "medium",
      height >= 180 ~ "large",
      TRUE ~ "unknown"
    )
  )
```

--

- Conditions are evaluated in order
- The last `TRUE ~ ...` is the "otherwise" catch-all

---

## `summarize()` and `group_by()`

```{r echo=TRUE, eval=FALSE}
starwars %>%
  group_by(species) %>%
  summarize(
    n = n(),
    mean_height = mean(height, na.rm = TRUE),
    mean_mass = mean(mass, na.rm = TRUE)
  )
```

--

- `group_by()` alone doesn't change the data, but affects subsequent operations
- `n()` counts rows in each group
- Always use `na.rm = TRUE` for summary stats when NAs may be present
- `ungroup()` to remove grouping

---

## The Pipe: `%>%`

Chain operations together:

```{r echo=TRUE, eval=FALSE}
starwars %>%
  filter(species == "Human") %>%
  select(name, height, mass) %>%
  mutate(bmi = mass / (height / 100)^2) %>%
  arrange(desc(bmi))
```

--

- Read `%>%` as "and then"
- The output of the left side becomes the first argument of the right side
- Makes code readable from top to bottom

---

## Joining Data

Combine two data frames by a shared key:

.pull-left[
**Mutating joins** (add columns):
- `left_join()` — keep all rows from left
- `right_join()` — keep all rows from right
- `inner_join()` — keep only matching rows
- `full_join()` — keep all rows from both
]

.pull-right[
**Filtering joins** (filter rows):
- `semi_join()` — keep rows in left that match right
- `anti_join()` — keep rows in left that don't match right
]

---

## Joining Example

```{r echo=TRUE, eval=FALSE}
# Two data frames with a shared "id" column
left_join(df1, df2, by = "id")

# Different column names
left_join(df1, df2, by = c("student_id" = "id"))
```

--

- The most common join is `left_join()` — it keeps all rows from the left data frame
- If there are multiple matches, rows are duplicated
- Watch out for duplicate rows after joining!

---

## Reshaping: `pivot_longer()` and `pivot_wider()`

.pull-left[
**Wide to long:**
```{r echo=TRUE, eval=FALSE}
df %>%
  pivot_longer(
    cols = c(y2018, y2019, y2020),
    names_to = "year",
    values_to = "value"
  )
```
]

.pull-right[
**Long to wide:**
```{r echo=TRUE, eval=FALSE}
df %>%
  pivot_wider(
    names_from = year,
    values_from = value
  )
```
]

--

- **Tidy data**: each variable in a column, each observation in a row, each value in a cell
- Most `ggplot2` and `dplyr` functions work best with tidy (long) data

---

class: middle, center, inverse

# Data Visualization

---

## `ggplot2` Grammar of Graphics

Every ggplot has three required components:

--

1. **Data** — the data frame

--

2. **Aesthetics** (`aes()`) — mappings from data to visual properties (x, y, color, size, shape...)

--

3. **Geometry** (`geom_*()`) — the type of plot

```{r echo=TRUE, eval=FALSE}
ggplot(data = starwars, aes(x = height, y = mass)) +
  geom_point()
```

---

## Common `geom_` functions

| Geom | Use |
|------|-----|
| `geom_point()` | scatterplot |
| `geom_line()` | line chart |
| `geom_bar()` | bar chart (counts) |
| `geom_col()` | bar chart (values) |
| `geom_histogram()` | histogram |
| `geom_boxplot()` | boxplot |
| `geom_density()` | density curve |
| `geom_smooth()` | smoothed trend line |
| `geom_text()` / `geom_label()` | text annotations |

---

## Aesthetics: `aes()`

Map data variables to visual properties:

```{r echo=TRUE, eval=FALSE}
ggplot(starwars, aes(x = height, y = mass,
                     color = gender, size = birth_year)) +
  geom_point(alpha = 0.6)
```

--

- **Inside `aes()`**: mapped to data (varies by observation)
- **Outside `aes()`**: fixed for all observations
- `alpha` controls transparency (0 = invisible, 1 = opaque)

---

## Faceting

Split a plot into subplots by a variable:

```{r echo=TRUE, eval=FALSE}
ggplot(starwars, aes(x = height)) +
  geom_histogram() +
  facet_wrap(~ gender)

ggplot(starwars, aes(x = height, y = mass)) +
  geom_point() +
  facet_grid(gender ~ species)
```

--

- `facet_wrap()` — wrap panels into a grid
- `facet_grid()` — explicitly specify rows and columns
- Use `scales = "free"` to allow different axis scales per panel

---

## Labels, Themes, and Scales

```{r echo=TRUE, eval=FALSE}
ggplot(starwars, aes(x = height, y = mass, color = gender)) +
  geom_point() +
  labs(
    title = "Height vs. Mass in Star Wars",
    x = "Height (cm)",
    y = "Mass (kg)",
    color = "Gender"
  ) +
  theme_minimal() +
  scale_color_viridis_d()
```

--

- `labs()` — set title, axis labels, legend titles
- `theme_*()` — overall look (minimal, bw, classic, etc.)
- `scale_*()` — control how data maps to visual properties

---

## Visualizing Distributions

.pull-left[
**One variable:**
- `geom_histogram()` — binned counts
- `geom_density()` — smoothed density
- `geom_boxplot()` — five-number summary
- `geom_violin()` — density + boxplot

]

.pull-right[
**Two variables:**
- `geom_point()` — scatterplot
- `geom_smooth()` — trend line
- `geom_jitter()` — jittered points

**Categorical:**
- `geom_bar()` — counts
- `geom_col()` — pre-computed values
]

---

class: middle, center, inverse

# Probability & Distributions

---

## Probability Basics

- **Probability**: a number between 0 and 1 representing the likelihood of an event
- **P(A)** = probability of event A
- **P(A') = 1 - P(A)** — complement
- **P(A or B) = P(A) + P(B) - P(A and B)** — addition rule
- **P(A and B) = P(A) × P(B|A)** — multiplication rule

--

**Independence:**
- Events A and B are independent if P(A and B) = P(A) × P(B)
- Knowing B occurred doesn't change the probability of A

---

## Conditional Probability

$$P(A|B) = \frac{P(A \text{ and } B)}{P(B)}$$

--

- Read as: "probability of A **given** B"
- The sample space is restricted to cases where B occurred

--

**Bayes' Theorem:**
$$P(A|B) = \frac{P(B|A) \cdot P(A)}{P(B)}$$

---

## Random Variables

- A **random variable** assigns a number to each outcome of a random process
- **Discrete**: countable values (e.g., number of heads in 10 flips)
- **Continuous**: any value in an interval (e.g., height, time)

--

**Expected value (mean):**
$$E(X) = \mu = \sum x \cdot P(X = x)$$

**Variance:**
$$Var(X) = \sigma^2 = \sum (x - \mu)^2 \cdot P(X = x)$$

---

## The Normal Distribution

$$X \sim N(\mu, \sigma^2)$$

- Symmetric, bell-shaped
- Defined by mean $\mu$ and standard deviation $\sigma$
- **68-95-99.7 rule**: 68% within 1 SD, 95% within 2 SD, 99.7% within 3 SD

--

**Standardizing (Z-score):**
$$Z = \frac{X - \mu}{\sigma}$$

The **standard normal** distribution: $Z \sim N(0, 1)$

---

## Working with Normal Distributions in R

```{r echo=TRUE, eval=FALSE}
# P(X < x)
pnorm(x, mean = mu, sd = sigma)

# P(X > x)
pnorm(x, mean = mu, sd = sigma, lower.tail = FALSE)

# Find x given P(X < x) = p
qnorm(p, mean = mu, sd = sigma)

# Generate random normal values
rnorm(n, mean = mu, sd = sigma)
```

--

- `p` functions: give probabilities (areas under the curve)
- `q` functions: give quantiles (the reverse)
- `d` functions: give density (height of the curve)
- `r` functions: generate random samples

---

## Other Important Distributions

.pull-left[
**Binomial**: count of successes in $n$ independent trials with probability $p$
$$X \sim B(n, p)$$
- `dbinom()`, `pbinom()`, `qbinom()`, `rbinom()`

**Poisson**: count of events in a fixed interval
$$X \sim Pois(\lambda)$$
- `dpois()`, `ppois()`, `qpois()`, `rpois()`
]

.pull-right[
**t-distribution**: like normal but heavier tails; used when $\sigma$ unknown
- `dt()`, `pt()`, `qt()`, `rt()`

**F-distribution**: ratio of variances; used in ANOVA
- `df()`, `pf()`, `qf()`, `rf()`

**Chi-square**: used for categorical data tests
- `dchisq()`, `pchisq()`, `qchisq()`, `rchisq()`
]

---

class: middle, center, inverse

# Sampling Distributions & Inference

---

## Sampling Distribution of $\bar{X}$

If we repeatedly sample $n$ observations from a population with mean $\mu$ and SD $\sigma$:

$$\bar{X} \sim N\left(\mu, \frac{\sigma^2}{n}\right)$$

--

- The **standard error** is $SE = \sigma / \sqrt{n}$
- As $n$ increases, the sampling distribution gets narrower
- This is the **Central Limit Theorem (CLT)**

---

## Central Limit Theorem (CLT)

For a sufficiently large sample size $n$, the sampling distribution of $\bar{X}$ is approximately normal, **regardless of the shape of the population distribution**.

--

- "Sufficiently large" is often $n \geq 30$
- The larger the sample, the better the approximation
- For normally distributed populations, any $n$ works

--

**Why does this matter?**
It lets us use normal-based methods (z-tests, t-tests, confidence intervals) even when we don't know the shape of the population.

---

## Confidence Intervals

A **confidence interval** gives a plausible range of values for a population parameter.

$$\bar{x} \pm z^* \cdot \frac{\sigma}{\sqrt{n}}$$

--

- $z^*$ is the critical value (e.g., 1.96 for 95% CI)
- A 95% CI means: if we repeated this procedure many times, 95% of the intervals would contain the true parameter
- **The interval either contains the parameter or it doesn't** — no probability statement about a single interval

---

## t-Confidence Intervals

When $\sigma$ is unknown (almost always), use the $t$-distribution:

$$\bar{x} \pm t^*_{df} \cdot \frac{s}{\sqrt{n}}$$

--

- $df = n - 1$ degrees of freedom
- Use `qt()` to find $t^*$: `qt(0.975, df = n - 1)`
- The t-distribution is wider than normal for small $n$, converges to normal as $n \to \infty$

---

class: middle, center, inverse

# Hypothesis Testing

---

## Hypothesis Testing Framework

1. **State hypotheses**: $H_0$ (null) and $H_a$ (alternative)
2. **Set significance level**: $\alpha$ (commonly 0.05)
3. **Collect data and compute test statistic**
4. **Calculate p-value**
5. **Make a decision**: reject or fail to reject $H_0$
6. **State conclusion in context**

---

## Hypotheses

- $H_0$: The null hypothesis — assumes no effect, no difference, status quo
- $H_a$: The alternative hypothesis — what we're trying to show

--

**Types of alternative hypotheses:**
- Two-sided: $H_a: \mu \neq \mu_0$
- One-sided (right): $H_a: \mu > \mu_0$
- One-sided (left): $H_a: \mu < \mu_0$

--

We **never** accept $H_0$ — we only **fail to reject** it. Absence of evidence is not evidence of absence.

---

## The p-value

The **p-value** is the probability of observing data as extreme as (or more extreme than) what we observed, **assuming $H_0$ is true**.

--

- Small p-value → data are unusual under $H_0$ → evidence against $H_0$
- Large p-value → data are consistent with $H_0$ → insufficient evidence against $H_0$

--

**Decision rule:**
- If $p < \alpha$: reject $H_0$ (result is "statistically significant")
- If $p \geq \alpha$: fail to reject $H_0$

---

## One-Sample t-test

Test whether a population mean equals a hypothesized value:

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

```{r echo=TRUE, eval=FALSE}
t.test(x, mu = mu_0)
# or
t.test(x, mu = mu_0, alternative = "greater")
```

--

- Under $H_0$, this statistic follows a $t$-distribution with $df = n - 1$
- The p-value is computed from that distribution

---

## Two-Sample t-test

Compare means from two independent groups:

$$t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}$$

```{r echo=TRUE, eval=FALSE}
t.test(x ~ group, data = df)
t.test(x1, x2)
```

--

- Assumes independence between groups
- Does **not** require equal variances (Welch's t-test is the default in R)
- Paired data: use `t.test(x1, x2, paired = TRUE)`

---

## Type I and Type II Errors

|  | $H_0$ is True | $H_0$ is False |
|--|---------------|----------------|
| **Reject $H_0$** | Type I Error ($\alpha$) | Correct! (Power) |
| **Fail to reject $H_0$** | Correct! | Type II Error ($\beta$) |

--

- **Type I error** ($\alpha$): false positive — rejecting a true $H_0$
- **Type II error** ($\beta$): false negative — failing to reject a false $H_0$
- **Power** = $1 - \beta$ = probability of correctly rejecting a false $H_0$

---

## Factors Affecting Power

Power increases when:

--

- Effect size is larger
--

- Sample size $n$ is larger
--

- $\alpha$ is larger (but this also increases Type I error rate)
--

- Population variance $\sigma^2$ is smaller

--

**There is always a trade-off between Type I and Type II error rates.**
Decreasing $\alpha$ (stricter test) increases $\beta$ (less power) for a fixed sample size.

---

## Effect Size

A standardized measure of the magnitude of an effect, independent of sample size.

**Cohen's d** (for comparing means):
$$d = \frac{\bar{x}_1 - \bar{x}_2}{s_{pooled}}$$

--

Conventional benchmarks (Cohen, 1988):
- Small: $d = 0.2$
- Medium: $d = 0.5$
- Large: $d = 0.8$

--

Statistical significance $\neq$ practical significance. A tiny effect can be statistically significant with a large enough sample.

---

class: middle, center, inverse

# Multiple Testing & Other Tests

---

## Multiple Comparisons Problem

If you run many tests at $\alpha = 0.05$, you expect 5% of true null hypotheses to be rejected by chance.

--

**Bonferroni correction**: adjust $\alpha$ downward:
$$\alpha_{adjusted} = \frac{\alpha}{m}$$
where $m$ is the number of tests.

--

- Conservative but simple
- Other methods: Benjamini-Hochberg (FDR), Holm

---

## Chi-Square Test of Independence

Test whether two categorical variables are associated:

$$\chi^2 = \sum \frac{(O - E)^2}{E}$$

```{r echo=TRUE, eval=FALSE}
chisq.test(table(df$var1, df$var2))
```

--

- $O$ = observed counts, $E$ = expected counts under independence
- $H_0$: the two variables are independent
- $df = (rows - 1)(cols - 1)$
- Requires expected cell counts $\geq 5$

---

## ANOVA

**Analysis of Variance**: compare means across 3+ groups.

$H_0$: all group means are equal
$H_a$: at least one mean differs

```{r echo=TRUE, eval=FALSE}
aov_result <- aov(y ~ group, data = df)
summary(aov_result)
```

--

- F-statistic = variance between groups / variance within groups
- Significant ANOVA → follow up with post-hoc tests (e.g., Tukey HSD)
- Assumptions: normality, equal variances, independence

---

## Linear Regression

Model the relationship between a response variable $Y$ and predictor(s) $X$:

$$Y = \beta_0 + \beta_1 X + \epsilon$$

```{r echo=TRUE, eval=FALSE}
lm_fit <- lm(y ~ x, data = df)
summary(lm_fit)
```

--

- $\beta_0$: intercept (expected $Y$ when $X = 0$)
- $\beta_1$: slope (expected change in $Y$ for one-unit increase in $X$)
- $R^2$: proportion of variance in $Y$ explained by the model
- Residuals: $e_i = y_i - \hat{y}_i$

---

## Regression Assumptions (LINE)

- **L**inearity: the relationship between $X$ and $Y$ is linear
- **I**ndependence: observations are independent
- **N**ormality: residuals are normally distributed
- **E**qual variance: residuals have constant variance (homoscedasticity)

--

Diagnose with residual plots:
```{r echo=TRUE, eval=FALSE}
plot(lm_fit)
```

<!-- TODO: add residual diagnostic plots image -->

---

class: middle, center, inverse

# Practice Problems

---

## Practice: Data Wrangling

Using the `starwars` dataset:

1. Filter to characters who are Human **or** Droid
2. Select `name`, `height`, `mass`, `species`
3. Create a new variable `bmi = mass / (height/100)^2`
4. Find the mean BMI by species
5. Arrange by mean BMI in descending order

--

```{r echo=TRUE, eval=FALSE}
starwars %>%
  filter(species %in% c("Human", "Droid")) %>%
  select(name, height, mass, species) %>%
  mutate(bmi = mass / (height / 100)^2) %>%
  group_by(species) %>%
  summarize(mean_bmi = mean(bmi, na.rm = TRUE)) %>%
  arrange(desc(mean_bmi))
```

---

## Practice: Probability

A researcher finds that 30% of students suffer from test anxiety. In a random sample of 20 students:

1. What is the probability that exactly 6 have test anxiety?
2. What is the probability that fewer than 4 have test anxiety?
3. What is the expected number with test anxiety?

--

```{r echo=TRUE, eval=FALSE}
# 1. P(X = 6)
dbinom(6, size = 20, prob = 0.3)

# 2. P(X < 4) = P(X <= 3)
pbinom(3, size = 20, prob = 0.3)

# 3. E(X) = n * p
20 * 0.3
```

---

## Practice: Hypothesis Testing

A psychology professor claims the mean score on a memory test is 75. A sample of 25 students has a mean of 72 and standard deviation of 8.

1. State the hypotheses (two-sided)
2. Calculate the test statistic
3. Find the p-value
4. What do you conclude at $\alpha = 0.05$?

--

$$t = \frac{72 - 75}{8/\sqrt{25}} = \frac{-3}{1.6} = -1.875$$

```{r echo=TRUE, eval=FALSE}
# p-value for two-sided test
2 * pt(-1.875, df = 24)  # ≈ 0.073
# Fail to reject H0; insufficient evidence at α = 0.05
```

---

## Practice: Confidence Interval

Using the same data (n = 25, $\bar{x}$ = 72, s = 8):

Construct a 95% confidence interval for the population mean.

--

$$72 \pm t^*_{24} \cdot \frac{8}{\sqrt{25}}$$

```{r echo=TRUE, eval=FALSE}
t_star <- qt(0.975, df = 24)  # ≈ 2.064
72 + c(-1, 1) * t_star * (8 / sqrt(25))
# [68.69, 75.31]
```

--

We are 95% confident that the true population mean is between 68.69 and 75.31.

Note: 75 (the null value) **is** inside this interval — consistent with failing to reject $H_0$.

---

## Key Takeaways

.pull-left[
**Wrangling & Viz:**
- Master the 6 `dplyr` verbs
- Use `%>%` to chain operations
- Joins add columns; filtering joins filter rows
- Tidy data: long format for analysis
- `ggplot2`: data + aes + geom
]

.pull-right[
**Inference:**
- CLT justifies normal-based methods
- CIs: plausible range for parameter
- p-value: probability under $H_0$
- Type I ($\alpha$) vs Type II ($\beta$) errors
- Power = $1 - \beta$
- Effect size $\neq$ statistical significance
]

---

## Test Tips

`r emo::ji("check")` Show your work on calculation questions

`r emo::ji("check")` State hypotheses in terms of the problem, not just symbols

`r emo::ji("check")` Interpret results in context

`r emo::ji("check")` Watch for "independent vs. paired" samples

`r emo::ji("check")` Know when to use z vs. t (known vs. unknown $\sigma$)

`r emo::ji("check")` Check whether assumptions are met before running tests

---

class: middle, center, inverse

# Wrapping Up...
