---
title: "Hypothesis Testing and Significance"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Hypothesis Testing and Significance

---

# Road Map

.pull-left[
**What is hypothesis testing?**
]
.pull-right[
Error rates
]

--

- By the end of today, you should be able to:
  - Define hypothesis testing and significance testing
  - Describe the 4 steps of hypothesis testing
  - Explain Type I and Type II errors
  - Define power and describe ways to influence it

---

# What is Hypothesis Testing?

- A statistical procedure that uses sample data to evaluate a hypothesis about a population

--

- Also called **significance testing**

--

- The goal: determine whether the observed results are likely due to chance or reflect a real effect

---

# Significance Testing

- We use data from a sample to draw inferences about a population

--

- Key question: Is the observed result **statistically significant**?

--

- Statistical significance means the result is unlikely to have occurred by chance alone

---

# The 4 Steps of Hypothesis Testing

1. **State the hypotheses** (null and alternative)

--

2. **Formulate an analysis plan** (choose test, set alpha)

--

3. **Analyze sample data** (compute the test statistic)

--

4. **Decide and interpret** (reject or fail to reject H0)

---

# The Null Hypothesis

- The **null hypothesis** (H0) is the default assumption — typically that there is no effect or no difference

--

- The **alternative hypothesis** (Ha or H1) is what the researcher expects to find

--

- We never "prove" the null hypothesis — we either reject it or fail to reject it

---

# The Jurisprudence Model

.pull-left[
- **Jury** = skeptical scientific community
- **Prosecutor** = researcher
- **Defendant** = null hypothesis
- **Detective** = statistician
]
.pull-right[
<!-- TODO: add image of courtroom/lawyer cat -->
]

--

Just as a jury starts with "innocent until proven guilty," we start by assuming H0 is true

---

# Definitions

- **Effect size**: the magnitude of the difference or relationship in the population

--

- **Alpha level** (α): the threshold for statistical significance (typically .05)
  - The probability of rejecting H0 when it is actually true

--

- **Power**: the probability of correctly rejecting H0 when it is false
  - Aim for power ≥ .80

---

# Consequences of Our Decisions

|  | H0 is True | H0 is False |
|--|-----------|------------|
| **Reject H0** | Type I Error (α) | Correct! (Power) |
| **Fail to Reject H0** | Correct! | Type II Error (β) |

---

# Type I Error

- Rejecting H0 when it is actually **true**

--

- A "false positive" — concluding there is an effect when there isn't one

--

- Probability of Type I error = **alpha (α)**
  - If α = .05, there is a 5% chance of a Type I error

---

# Type II Error

- Failing to reject H0 when it is actually **false**

--

- A "false negative" — missing a real effect

--

- Probability of Type II error = **beta (β)**

--

- Power = 1 − β
  - Higher power means lower chance of a Type II error

---

# Power and Type II Error

- **Power** = 1 − Type II Error (β)

--

- We want power to be high (aim for ≥ .80)

--

- Low power means we are likely to miss real effects

---

# Ways to Influence Power

- **Increase sample size** → increases power

--

- **Decrease standard deviation** (reduce noise) → increases power

--

- **Set a less stringent significance level** (e.g., α = .10 instead of .05) → increases power (but also increases Type I error risk)

--

- **Increase effect strength** (study stronger manipulations) → increases power

---

# A Brief History of the Null Hypothesis

- **R.A. Fisher (1935)** introduced the concept of the null hypothesis

--

- Fisher argued the null hypothesis can never be *proved*, only *disproved*

--

- If the data are sufficiently improbable under H0, we reject it

---

# Lady Tasting Tea

<!-- TODO: add image of tea cups -->

- Fisher's famous example: Can a lady tell whether milk or tea was poured first?

--

- She is given **8 cups**: 4 with milk first, 4 with tea first

--

- If she is just guessing, the probability of correctly identifying all 4 milk-first cups is:
  - 1/70 ≈ **1.4%**

--

- This is below α = .05, so we would reject H0 (that she is just guessing)

---

class: middle, center, inverse

# Mind Reading Activity

---

class: middle, center, inverse

# Wrapping Up...

- Hypothesis testing uses sample data to evaluate claims about populations
- The 4 steps: state hypotheses → analysis plan → compute test statistic → decide
- Type I error (false positive) = α; Type II error (false negative) = β
- Power = 1 − β; aim for ≥ .80
- Increase power via larger N, reduced SD, stronger effects
