---
title: "Experimental Design"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---


```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle

# Experimental Design

---

## Roadmap

- Review: Experiments with one independent variable
- Experiments with two independent variables can show interactions
- Factorial variations
- Identifying factorial designs in your readings

--

**By the end of this lecture, you should be able to:**

- Explain what a factorial design is and identify its components
- Define an interaction and describe what it means for findings
- Distinguish main effects from interactions
- Interpret 2x2 factorial results in table and graph form
- Describe independent-groups, within-groups, and mixed factorial designs
- Identify factorial designs in empirical and popular press articles

---

class: middle, center, inverse

# Review: Experiments with One Independent Variable

---

## Review: Experiments with One Independent Variable

<!-- TODO: add diagram or summary of one-IV experiment structure (IV levels → DV) -->

---

class: middle, center, inverse

# Experiments with Two Independent Variables Can Show Interactions

---

## Interaction: Definition

**Interaction =**

- A difference in differences =
- The effect of one independent variable **depends on the level** of the other independent variable

---

## Intuitive Interactions

<!-- TODO: add figure or examples of intuitive interactions (e.g., sun + rain = rainbow; sun + no rain = sunny day) -->

---

## Reconsidering the Cell Phone and Driving Study

<!-- TODO: add figure or diagram of cell phone and driving study setup showing the 2x2 design -->

---

class: middle, center, inverse

# Factorial Designs Study Two Independent Variables

---

## Factorial Designs: Key Terms

.pull-left[
**Factorial design**

A study that includes two or more independent variables, each with multiple levels, combined so that every level of one IV occurs with every level of the other IV(s).
]

.pull-right[
**Cell**

Each combination of levels of the IVs in a factorial design.

**Participant variable**

A variable that represents a pre-existing characteristic of participants (e.g., age, sex, diagnosis).
]

---

## Factorial Designs Can Test Limits

Sometimes, researchers conduct factorial designs to test whether an IV affects:

- different kinds of **people**, or
- people in different **situations**

in the same way.

---

## Factorial Design Results in Table and Graph Formats

<!-- TODO: add table and graph showing factorial design results (e.g., cell phone x driver age on reaction time) -->

---

## Factorial Designs Can Test Limits (continued)

- A form of **external validity**

--

- Interactions show **moderators**
  - A moderator is a variable that changes the relationship between the IV and DV

---

## Factorial Designs Can Test Theories

- Using a factorial design to test a theory of alcohol cues
- Using an interaction to test a memory theory

---

## Using a Factorial Design to Test a Theory of Alcohol Cues

| | **IV 1: Photo type** | |
|---|---|---|
| **IV 2: Word type** | Alcohol photo | Plant photo |
| Aggressive | | |
| Neutral | | |

DV: Reaction time (ms)

---

## Using a Factorial Design to Test a Theory of Alcohol Cues: Results

| | **IV 1: Photo type** | |
|---|---|---|
| **IV 2: Word type** | Alcohol photo | Plant photo |
| Aggressive | 551 | 559 |
| Neutral | 562 | 552 |

DV: Reaction time (ms)

---

## Using an Interaction to Test a Memory Theory

| | **IV 1: Participant type** | |
|---|---|---|
| **IV 2: Type of item** | Child chess experts | Adult novices |
| Chess pieces | 9.3 | 5.9 |
| Digits | 6.1 | 7.8 |

---

class: middle, center, inverse

# Interpreting Factorial Results: Main Effects and Interactions

---

## Interpreting Factorial Results

.pull-left[
**Main effect:** Is there an overall difference?
]

.pull-right[
**Interactions**

- Is there a difference in differences?
- More important than main effects
]

--

Possible main effects and interactions in a 2 x 2 factorial design.

---

## Main Effect: Is There an Overall Difference?

.pull-left[
**DV: Reaction time (ms)**

| | **IV 1: Photo type** | | |
|---|---|---|---|
| **IV 2: Word type** | Alcohol photo | Plant photo | **Marginal mean** |
| Aggressive | 551 | 559 | 555 |
| Neutral | 562 | 552 | 557 |
| **Marginal mean** | 556.5 | 555.5 | |

Main effect for IV2 (word type): 555 vs. 557

Main effect for IV1 (photo type): 556.5 vs. 555.5
]

.pull-right[
**DV: Shock intensity (1–10 scale)**

| | **IV 1: Body weight** | | |
|---|---|---|---|
| **IV 2: Drinking condition** | Light men (151 lbs) | Heavy men (215 lbs) | **Marginal mean** |
| Placebo group | 5.09 | 4.72 | 4.91 |
| Drunk group | 6.00 | 7.19 | 6.60 |
| **Marginal mean** | 5.55 | 5.96 | |

Main effect for IV2 (drinking condition): 4.91 vs. 6.60

Main effect for IV1 (body weight): 5.55 vs. 5.96
]

---

## Interactions: Is There a Difference in Differences? Options 1 and 2

.pull-left[
**Alcohol cues study**

| | Alcohol photo | Plant photo |
|---|---|---|
| Aggressive | 551 | 559 |
| Neutral | 562 | 552 |

Difference for alcohol photos: 551 − 562 = −11

Difference for plant photos: 559 − 552 = +7
]

.pull-right[
**Cell phone and driving study**

| | On cell phone | Not on phone |
|---|---|---|
| Younger drivers | 912 | 780 |
| Older drivers | 1,086 | 912 |

Difference for younger drivers: 912 − 780 = 132

Difference for older drivers: 1,086 − 912 = 174
]

---

## Interactions: Is There a Difference in Differences? Line and Bar Graphs

<!-- TODO: add line graphs and bar graphs showing interactions and non-interactions for the factorial design examples -->

---

## Interactions: Is There a Difference in Differences? Words

**Describing interactions in words:**

- "The effect of [IV1] on [DV] **depends on** the level of [IV2]."
- "There was an interaction between [IV1] and [IV2]."
- "[IV1] had a larger effect on [DV] when [IV2 level] than when [IV2 other level]."

---

## Interactions Are More Important Than Main Effects

When a study shows both a main effect and an interaction,
- the interaction is **almost always more important.**

--

There may be real differences in marginal means,
- but the more exciting part is the interaction.

---

## Possible Main Effects and Interactions in a 2 x 2 Factorial Design

<!-- TODO: add figure showing all 8 possible patterns of main effects and interactions in a 2x2 factorial (Pattern 1: no main effects, no interaction) -->

---

## Possible Main Effects and Interactions in a 2 x 2 Factorial Design

<!-- TODO: add figure showing patterns 2-3: one main effect only (IV1 or IV2), no interaction -->

---

## Possible Main Effects and Interactions in a 2 x 2 Factorial Design

<!-- TODO: add figure showing pattern 4: two main effects, no interaction -->

---

## Possible Main Effects and Interactions in a 2 x 2 Factorial Design (continued)

<!-- TODO: add figure showing pattern 5: interaction only, no main effects -->

---

## Possible Main Effects and Interactions in a 2 x 2 Factorial Design (continued)

<!-- TODO: add figure showing patterns 6-7: interaction plus one main effect -->

---

## Possible Main Effects and Interactions in a 2 x 2 Factorial Design (continued)

<!-- TODO: add figure showing pattern 8: interaction plus two main effects -->

---

class: middle, center, inverse

# Wrapping Up...

---

class: middle, center, inverse

# This Time: Factorial Variations

---

## Factorial Variations

- Independent-groups factorial designs
- Within-groups factorial designs
- Mixed factorial designs
- Increasing the number of levels of an independent variable
- Increasing the number of independent variables

---

## Independent-Groups Factorial Designs

.pull-left[
Both IVs are studied as **independent groups**.

A 2 x 2 independent-groups factorial design has **four groups or cells**.

This is also known as a **between-subjects factorial design**.
]

---

## Within-Groups Factorial Designs

.pull-left[
Both IVs are studied as **within-groups** (repeated measures).

All participants experience all conditions.

Fewer participants needed, but order effects must be managed.
]

.pull-right[
<!-- TODO: add diagram illustrating within-groups factorial design structure -->
]

---

## Mixed Factorial Designs

One IV is manipulated as **independent-groups** and the other is manipulated **within-groups**.

--

This design is intermediate between the within-groups design and the independent-groups design in terms of number of participants.

---

## Increasing the Number of Levels of an Independent Variable

| | **IV 1: Cell phone condition** | | |
|---|---|---|---|
| **IV 2: Driver age** | Hands-free cell phone | Handheld cell phone | No phone |
| Younger drivers | 1,100 | 1,200 | 850 |
| Older drivers | 1,450 | 1,500 | 1,050 |

---

## Increasing the Number of Independent Variables

- Main effects and interactions from a three-way design
- Why worry about all these interactions?

---

## Main Effects and Interactions from a Three-Way Design

**Main effects:** Is there a difference?

- A three-way design (A x B x C) has **three main effects**:
  - Main effect of A
  - Main effect of B
  - Main effect of C

---

## Main Effects and Interactions from a Three-Way Design: Two-Way Interaction

**Two-way interactions:** Is there a difference in differences?

- A three-way design has **three two-way interactions**:
  - A × B interaction
  - A × C interaction
  - B × C interaction

---

## Main Effects and Interactions from a Three-Way Design: Three-Way Interaction

**Three-way interactions:** Are the two-way interactions different?

- A three-way design has **one three-way interaction**:
  - A × B × C interaction

<!-- TODO: add figure or diagram illustrating a three-way interaction -->

---

## Main Effects and Interactions from a Three-Way Design (continued)

**Three-way interactions:** Are the two-way interactions different?

- The A × B interaction is different at different levels of C.

<!-- TODO: add graphs illustrating the three-way interaction from different slices -->

---

## Why Test All These Interactions?

Most outcomes in psychological studies are **not main effects**, but rather **interactions**.

---

class: middle, center, inverse

# Identifying Factorial Designs in Your Reading

---

## Identifying Factorial Designs

.pull-left[
**IDENTIFYING FACTORIAL DESIGNS IN EMPIRICAL JOURNAL ARTICLES**
]

.pull-right[
**IDENTIFYING FACTORIAL DESIGNS IN POPULAR PRESS ARTICLES**
]

---

## Identifying Factorial Designs in Empirical Journal Articles

- The **Method section** will describe the design of the study.
  - Factorial notation: _____ x _____ x _____

--

- The **Results section** will examine whether the main effects and interactions were significant.
  - Reference to "significance" or a p value and information about MANOVA and F

---

## Identifying Factorial Designs in Popular Media Articles

.pull-left[
- Look for phrases such as **"it depends"** or **"only when"** used to highlight an interaction.

- Look for **participant variables** (e.g., age, gender, ethnicity).
]

---

class: middle, center, inverse

# Wrapping Up...
