---
title: "F-tests, Regression, and ANOVA Pt. 2"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# F-tests, Regression, and ANOVA Pt. 2

---

class: middle

# Picking up where we left off...

<!-- TODO: add image of "picking up where we left off" graphic -->

---

class: middle

# This Time...

**The idea of analysis of variance**

<!-- TODO: add image illustrating analysis of variance concept -->

---

# The Idea of ANOVA

- What matters is **relative variability**

--

- ANOVA tests whether multiple populations have the **same mean**

--

- Key insight: we compare variability *between* groups to variability *within* groups

---

# ANOVA Tests

- ANOVA (Analysis of Variance) tests whether populations have the same mean

--

- H0: μ1 = μ2 = μ3 = ... (all group means are equal)
- Ha: At least one mean differs

--

- We use the **F statistic** to make this determination

---

# Variation Within and Between

<!-- TODO: add image of box plots comparing conditions a vs b -->

.pull-left[
**Scenario A**: Large overlap between groups → groups look similar
]
.pull-right[
**Scenario B**: Little overlap between groups → groups look different
]

--

The question is always: how much do groups differ *relative to* how much individuals vary within groups?

---

# Interpretation

- When groups overlap a lot → variation between groups is small relative to variation within groups → **not significant**

--

- When groups are well-separated → variation between groups is large relative to variation within groups → **significant**

---

# The ANOVA F Statistic

- ANOVA provides a general method for testing equality of means

--

- In **one-way ANOVA**, we have one grouping variable (factor)

--

$$F = \frac{\text{variation among sample means}}{\text{variation among individuals within the same sample}}$$

---

# Between-Group Variability

<!-- TODO: add image of overlapping distributions for multiple groups -->

- When distributions overlap heavily, between-group variability is **small**

--

<!-- TODO: add image of well-separated distributions for multiple groups -->

- When distributions are well-separated, between-group variability is **large**

--

- **Between-group variability** is measured by the **Mean Square for Groups (MSG)**

--

- MSG is based on the variability of each group mean around the **grand mean** (overall mean)

---

# Between-Group Variability: Grand Mean

<!-- TODO: add image of grand mean diagram showing group means relative to overall mean -->

- The **grand mean** $\bar{x}$ is the mean of all observations across all groups

--

- MSG captures how far each group mean deviates from this grand mean

---

# Within-Group Variability

<!-- TODO: add image of three overlapping samples illustrating within-group spread -->

- **Within-group variability** is the spread of individual scores within each group

--

<!-- TODO: add image of three well-separated samples to contrast -->

- Even when groups differ, there is always some within-group variability

--

- **Within-group variability** is measured by the **Mean Square Error (MSE)**

---

# Measuring Within-Group Variability

<!-- TODO: add image of sum of squares diagram for within-group variability -->

- MSE is based on the **sum of squared deviations** of each observation from its group mean

--

- Also called the **error** or **residual** variability — the part we can't explain with group membership

---

# Some Details of ANOVA

$$MSG = \frac{\sum n_i(\bar{x}_i - \bar{x})^2}{k - 1}$$

--

$$MSE = \frac{\sum (n_i - 1)s_i^2}{N - k}$$

Where:
- $k$ = number of groups
- $N$ = total sample size
- $n_i$ = sample size of group $i$
- $\bar{x}_i$ = mean of group $i$
- $\bar{x}$ = grand mean
- $s_i^2$ = variance of group $i$

---

# Putting the Pieces Together

$$F = \frac{MSG}{MSE}$$

<!-- TODO: add diagram showing how MSG and MSE combine into F statistic -->

--

- Large F → between-group variability >> within-group variability → evidence against H0

--

- Small F (≈ 1) → between-group variability ≈ within-group variability → consistent with H0

---

<!-- TODO: add image illustrating within vs. between group variance side by side -->

---

# Putting the Pieces Together: Formulas

$$F = \frac{MSG}{MSE} = \frac{\text{Between-group variability}}{\text{Within-group variability}}$$

--

$$MSG = \frac{SSG}{k-1}, \quad MSE = \frac{SSE}{N-k}$$

--

$$F = \frac{SSG/(k-1)}{SSE/(N-k)}$$

---

# ANOVA and the F Statistic

<!-- TODO: add image of F distribution with critical region shaded -->

- The F distribution is right-skewed

--

- We reject H0 when the observed F exceeds the **critical value**

--

- The critical value depends on:
  - Significance level α
  - Degrees of freedom: df1 = k − 1, df2 = N − k

---

# ANOVA and the F Statistic: F-Table

- **df1** (numerator) = k − 1 (between groups)
- **df2** (denominator) = N − k (within groups)

--

- Use an F-table or software to find the critical value or p-value

---

# The F-Test

$$F = \frac{\text{Signal}}{\text{Noise}} = \frac{\text{Between-group variability}}{\text{Within-group variability}}$$

--

- "Signal" = systematic differences between groups
- "Noise" = random variation within groups

---

# If the Null Hypothesis Is True...

- When H0 is true, all groups come from populations with the same mean

--

- MSG ≈ MSE, so **F ≈ 1**

--

- Any observed F value should be close to 1 (with random variation)

---

# Why Use ANOVA Instead of Multiple t-Tests?

- If we run many t-tests (one per pair of groups), the probability of a Type I error accumulates

--

- With 3 groups and α = .05:
  - Each test has a 5% chance of a false positive
  - Running 3 tests: actual Type I error rate ≈ 1 − (.95)³ ≈ **14%**

--

- ANOVA controls the Type I error rate at α across all comparisons simultaneously

---

# If the Null Can Be Rejected...

- When H0 is false and groups truly differ, MSG >> MSE

--

- **F = BIG number**

--

- A large F suggests the group differences are too large to attribute to chance

---

# What ANOVA Can Tell Us

- **Main effects**: whether a factor (grouping variable) has an overall effect

--

- **Interaction effects** (in factorial ANOVA): whether the effect of one factor depends on another factor

--

- ANOVA tells us *that* at least one group differs — but not *which* groups differ

--

- For that, we need **post-hoc tests** (e.g., Tukey HSD, Bonferroni)

---

class: middle, center, inverse

# Wrapping Up...

- ANOVA compares between-group variability (MSG) to within-group variability (MSE)
- $F = MSG/MSE$; large F → reject H0
- When H0 is true, $F \approx 1$
- ANOVA controls Type I error better than multiple t-tests
- ANOVA tells us *that* groups differ; post-hocs tell us *which* ones
