---
title: "Lying with Statistics: Research Ethics"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Lying with Statistics

<!-- TODO: add image of multiple editions of "How to Lie with Statistics" by Darrell Huff (blue cover and orange cover versions) -->

---

## Tuskegee Syphilis Study

- **1926:** Syphilis is a major health problem.
  - "one of medicine's most intricate problems"
    - (Jones, 1944)
  - Prevalence of 35 percent in adults

<!-- TODO: add image of chart showing "Probable Outcome of Acute Untreated Syphilis" with stages: Acute Disease, Secondary Stage, Early Relapse, Benign Syphilis (Tertian), Neuro-Syphilis, Cardio-Vascular Syphilis, and Spontaneous Cure 25-35% -->

---

## The Syphilis Studies

- **1926:** Syphilis is a major health problem.
  - "one of medicine's most intricate problems" (Jones, 1944)
  - Prevalence of 35 percent in adults

<!-- TODO: add image showing severe syphilis skin symptoms on a patient's back -->

---

## The Syphilis Studies

- **1926:** Syphilis is a major health problem.
  - "one of medicine's most intricate problems" (Jones, 1944)
  - Prevalence of 35 percent in adults

<!-- TODO: add image of a skull/death imagery representing fatal consequences of untreated syphilis -->

---

## Tuskegee Syphilis Study

- **1929:** Aggressive treatment approach
  - initiated with mercury and bismuth.
- Cure rate is less than 30 percent;
  - treatment requires months and
  - side effects are toxic, sometimes fatal.
- **1931:** Funding is cut from project
  - decide to follow men left untreated due to lack of funds in order to show need for treatment program.
- **1932:** Follow-up effort organized into study of 399 men with syphilis and 201 without.

<!-- TODO: add image of vintage syphilis public health poster "False Shame and Fear May Destroy Your Future - Have Your Blood Tested" -->

---

## Tuskegee Syphilis Study

- The men were impoverished, African-American sharecroppers from Alabama
- The men were given periodic physical assessments and
  - told they were being treated for "bad blood,"
    - local term used to describe several ailments, including syphilis, anemia, and fatigue
- Original study was designed to
  - last 6 months

<!-- TODO: add historical photograph of a medical professional drawing blood from a Black patient in the Tuskegee study -->

---

## Tuskegee Syphilis Study

- **1936:** Major paper published.
  - Study criticized because it is not known if men are being treated.
  - Local physicians asked to assist with study and not to treat men.
  - Decision was made to follow the men until death.

- **World War II:** Researchers attempted to prevent drafted men in the study from getting treatment
  - "So far, we are keeping the known positive patients from getting treatment."
  - Despite this, 96% of the 90 original test subjects reexamined in 1963
    - had received treatments from another health provider

---

## Tuskegee Syphilis Study

- **1945:** Penicillin accepted as treatment of choice for syphilis.

- **1947:** USPHS establishes "Rapid Treatment Centers" to treat syphilis;
  - men in study are not treated

<!-- TODO: add images of four vintage syphilis public health posters: "Six Out of Ten Cured", "Syphilis Strikes One Out of Ten Adults", "Shame May Be Fatal", "Syphilis is a Dangerous Disease - but it can be cured" -->

---

## Tuskegee Syphilis Study

- **1968:** 60% of Men in Study are Dead
  - Concerns raised about ethics of study by
    - Peter Buxtun, the eventual "whistleblower", as well as others

- **1969** CDC reaffirms need for study and
  - gains official support from
    - American Medical Association and
    - National Medical Association

- **Early 1970s** Buxtun leaks story to media
  - The story first reported by
  - Washington Star on July 25, 1972

<!-- TODO: add image of Classification of Cases in Tuskegee Study table showing Controls/Syphilitic/Total breakdown with known dead percentages -->

---

## Tuskegee Syphilis Study

- **July 1972:** Ad Hoc Advisory Panel created by US Gov
  - nine members from medicine, law, religion, labor, education, health administration, and public affairs.

- **October 1972:** The advisory panel concluded that the Tuskegee Study was "ethically unjustified"

- **November 1972:** Study Officially ends.
  - 74 men in study alive. Of the original 399 infected men,
  - 128 had died of syphilis or related complications,
  - 40 of their wives had been infected, and
  - 19 of their children were born with congenital syphilis.

- **1973** Congress holds hearings and
  - a class-action lawsuit is filed on behalf of the study participants.

---

## Tuskegee Syphilis Study

- **1974** Congress passed the National Research Act
  - and created a commission to study and write regulations governing studies involving human participants.
  - Within the US Department of Health and Human Services,
    - the Office for Human Research Protections (OHRP) was established to oversee clinical trials.

- Now studies require **informed consent**,
  - communication of diagnosis,
  - accurate reporting of test results, and
  - approval from **Institutional review boards (IRBs)**,
    - including laypeople, are established in scientific research groups and hospitals to review study protocols and protect patient interests, to ensure that participants are fully informed.

---

## Tuskegee Syphilis Study

- **1974:** A $10 million out-of-court settlement is reached and the U.S. government promised to give lifetime medical benefits and burial services to all living participants.

- **1975:** Wives, widows, and offspring were added to the program.

- **1997:** On May 16th President Clinton apologizes on behalf of the Nation.

---

class: middle, center, inverse

# Wrapping Up...
