---
title: "Sampling Distributions"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

---

class: middle, center, inverse

# Sampling Distributions

---

# Roadmap

.pull-left[
**Where we've been:**
- Exploratory data analysis
- Data visualization
- Probability
]

.pull-right[
**Where we're going:**
- Sampling distributions
- Central Limit Theorem
- Inference
]

--

**By the end of this lecture, you should be able to:**

- Distinguish between a parameter and a statistic
- Explain what a sampling distribution is and how it is constructed
- Describe the Central Limit Theorem and its conditions
- Calculate and interpret the standard error of a sample mean or proportion
- Apply the normal model to answer questions about sample statistics

---

class: middle, center, inverse

# Parameters and Statistics

---

# Parameters vs. Statistics

.pull-left[
**Parameter**
- A numerical summary of a **population**
- Fixed (but usually unknown)
- Denoted with Greek letters
  - Population mean: $\mu$
  - Population proportion: $p$
  - Population SD: $\sigma$
]

.pull-right[
**Statistic**
- A numerical summary of a **sample**
- Varies from sample to sample
- Denoted with Latin letters (or hats)
  - Sample mean: $\bar{x}$
  - Sample proportion: $\hat{p}$
  - Sample SD: $s$
]

---

# Parameters vs. Statistics

<!-- TODO: add [diagram showing population with parameter mu/p and sample with statistic x-bar/p-hat] image -->

- We use **statistics** to **estimate** parameters
- Because we rarely have access to the full population

---

class: middle, center, inverse

# Sampling Variability

---

# Sampling Variability

- Different samples from the same population produce different statistics
- This is called **sampling variability** (or sampling error)

--

- Example: Take many samples of size $n$ from a population
  - Each sample gives a (slightly) different $\bar{x}$ or $\hat{p}$
  - The values cluster around the true parameter

--

- Key question: **How much** does the statistic vary from sample to sample?

---

# Sampling Variability

<!-- TODO: add [dotplot or histogram showing distribution of many sample means from repeated sampling] image -->

- Each dot represents the mean of one sample
- The distribution of these sample means forms a **sampling distribution**

---

class: middle, center, inverse

# Sampling Distributions

---

# Sampling Distribution

The **sampling distribution** of a statistic is the distribution of that statistic computed from all possible samples of size $n$ from the same population.

--

- In practice, we don't take all possible samples
- But we can use theory to describe the shape, center, and spread of the sampling distribution

--

**Three things to describe:**
1. Shape
2. Center (mean)
3. Spread (standard error)

---

# Sampling Distribution of $\hat{p}$

For a sample proportion $\hat{p}$ based on $n$ observations from a population with true proportion $p$:

--

- **Center:** $\mu_{\hat{p}} = p$
- **Spread:** $\sigma_{\hat{p}} = \sqrt{\dfrac{p(1-p)}{n}}$
- **Shape:** Approximately normal when conditions are met

--

**Conditions for normality:**
- $np \geq 10$
- $n(1-p) \geq 10$

---

# Sampling Distribution of $\bar{x}$

For a sample mean $\bar{x}$ based on $n$ observations from a population with mean $\mu$ and standard deviation $\sigma$:

--

- **Center:** $\mu_{\bar{x}} = \mu$
- **Spread:** $\sigma_{\bar{x}} = \dfrac{\sigma}{\sqrt{n}}$
- **Shape:** Depends on the population and $n$

---

class: middle, center, inverse

# The Central Limit Theorem

---

# Central Limit Theorem (CLT)

> If a sample consists of at least 30 independent observations and the data are not strongly skewed, then the distribution of the sample mean $\bar{x}$ is well approximated by a normal distribution.

--

$$\bar{x} \sim N\!\left(\mu,\ \frac{\sigma}{\sqrt{n}}\right)$$

--

- The CLT applies **regardless of the shape of the population distribution**
- As $n$ increases, the sampling distribution becomes more normal
- Works for proportions too (with success/failure condition)

---

# Central Limit Theorem — Illustrated

<!-- TODO: add [four panels showing population distributions (uniform, skewed, bimodal) and their sampling distributions converging to normal as n increases] image -->

- Even if the population is skewed or non-normal...
- ...the sampling distribution of $\bar{x}$ becomes approximately normal as $n$ grows

---

# CLT Conditions

**Independence:**
- Sampled observations must be independent
- If sampling without replacement: sample size $< 10\%$ of population

**Sample size / skew:**
- $n \geq 30$ is a useful rule of thumb
- For strongly skewed populations, larger $n$ may be needed
- For symmetric populations, smaller $n$ may suffice

---

class: middle, center, inverse

# Standard Error

---

# Standard Error

The **standard error (SE)** is the standard deviation of a sampling distribution.

--

.pull-left[
**Standard error of $\bar{x}$:**
$$SE_{\bar{x}} = \frac{\sigma}{\sqrt{n}}$$

- As $n$ increases, $SE$ **decreases**
- Larger samples give more precise estimates
]

.pull-right[
**Standard error of $\hat{p}$:**
$$SE_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}}$$

- Also decreases as $n$ increases
- Most variable when $p = 0.5$
]

---

# Standard Error — Key Insight

<!-- TODO: add [side-by-side histograms showing sampling distributions with small n vs large n, illustrating narrower spread with larger n] image -->

- Doubling $n$ does **not** halve the SE
- To halve the SE, you need to **quadruple** $n$

$$SE \propto \frac{1}{\sqrt{n}}$$

---

class: middle, center, inverse

# Applying the Normal Model

---

# Using the Normal Model

Once we know the sampling distribution is approximately normal, we can calculate probabilities.

--

**Example: Sample mean**

A population has $\mu = 70$ and $\sigma = 10$. A sample of $n = 25$ is taken.

- $\mu_{\bar{x}} = 70$
- $SE_{\bar{x}} = 10 / \sqrt{25} = 2$
- $\bar{x} \sim N(70,\ 2)$

--

What is the probability that the sample mean exceeds 73?

$$Z = \frac{73 - 70}{2} = 1.5$$

---

# Using the Normal Model — Proportions

**Example: Sample proportion**

Suppose 60% of voters support a candidate ($p = 0.60$). A poll surveys $n = 400$ people.

--

- $\mu_{\hat{p}} = 0.60$
- $SE_{\hat{p}} = \sqrt{0.60 \times 0.40 / 400} = 0.0245$
- $\hat{p} \sim N(0.60,\ 0.0245)$

--

Check conditions:
- $np = 400 \times 0.60 = 240 \geq 10$ `r emo::ji("check")`
- $n(1-p) = 400 \times 0.40 = 160 \geq 10$ `r emo::ji("check")`

---

# Normal Model — Example Continued

<!-- TODO: add [normal curve showing sampling distribution of p-hat centered at 0.60 with shaded region] image -->

What is the probability the poll shows less than 57% support?

$$Z = \frac{0.57 - 0.60}{0.0245} \approx -1.22$$

$$P(\hat{p} < 0.57) = P(Z < -1.22) \approx 0.111$$

---

class: middle, center, inverse

# Putting It All Together

---

# Sampling Distribution Summary

| | Sample Mean $\bar{x}$ | Sample Proportion $\hat{p}$ |
|---|---|---|
| **Center** | $\mu$ | $p$ |
| **Spread (SE)** | $\sigma / \sqrt{n}$ | $\sqrt{p(1-p)/n}$ |
| **Shape (approx. normal when...)** | $n \geq 30$, not strongly skewed | $np \geq 10$ and $n(1-p) \geq 10$ |

--

- Statistics are **unbiased** estimators of their parameters (center = parameter)
- SE quantifies **precision** — how much the statistic varies
- Normal model enables **probability calculations**

---

# Why Does This Matter?

- Sampling distributions are the **foundation of statistical inference**
- Confidence intervals and hypothesis tests both rely on knowing the sampling distribution
- The CLT tells us that, for large enough samples, we can always use the normal model

--

- Next: using sampling distributions to build **confidence intervals**

---

class: middle, center, inverse

# Wrapping Up...

---

# Summary

- A **parameter** describes a population; a **statistic** describes a sample
- Statistics vary across samples — this is **sampling variability**
- The **sampling distribution** describes how a statistic varies across all possible samples
- The **Central Limit Theorem** tells us the sampling distribution of $\bar{x}$ (and $\hat{p}$) is approximately normal for large $n$
- The **standard error** measures the spread of the sampling distribution: $SE_{\bar{x}} = \sigma/\sqrt{n}$

---

# References

```{r, results='asis', echo=FALSE}
PrintBibliography <- function(...) invisible(NULL)
```
