---
title: "Probability"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Probability

---

# Roadmap

- What is probability?
- Defining probability: classical, empirical, subjective
- Sample spaces and events
- Probability rules
- Conditional probability and independence
- Bayes' theorem
- Introduction to probability distributions

--

**By the end of this lecture, you should be able to:**

- Define probability and distinguish among classical, empirical, and subjective interpretations
- Identify sample spaces and events, and apply the addition and multiplication rules
- Explain conditional probability and statistical independence
- Apply Bayes' theorem to update probabilities given new information
- Describe the properties of the normal distribution and use z-scores

---

class: middle, center, inverse

# What Is Probability?

---

# Probability in Everyday Life

- "There's a 70% chance of rain tomorrow."
- "The probability of rolling a 6 is 1/6."
- "I'm 90% sure I locked the door."

--

All of these are statements about **uncertainty** — probability is the language of uncertainty.

---

# Formal Definition

- **Probability**: a number between 0 and 1 that expresses the likelihood of an event
  - 0 = impossible
  - 1 = certain
  - 0.5 = equally likely to occur or not

$$0 \leq P(A) \leq 1$$

---

# Three Interpretations of Probability

.pull-left[
**1. Classical (a priori)**
- Based on equally likely outcomes
- $P(A) = \frac{\text{# favorable outcomes}}{\text{# total outcomes}}$
- Example: fair coin, fair die

**2. Empirical (frequentist)**
- Based on long-run relative frequency
- $P(A) = \lim_{n \to \infty} \frac{\text{# times A occurs}}{n}$
- Example: proportion of heads in many coin flips
]
.pull-right[
**3. Subjective (Bayesian)**
- Based on personal degree of belief
- Updated as new information arrives
- Example: "I believe there's a 60% chance it will rain"

<!-- TODO: add [three-panel illustration of classical/empirical/subjective probability] image -->
]

---

class: middle, center, inverse

# Sample Spaces and Events

---

# Key Vocabulary

- **Experiment**: a process with an uncertain outcome
- **Outcome**: a single result of an experiment
- **Sample space (S)**: the set of all possible outcomes
- **Event (A)**: a subset of the sample space

---

# Examples of Sample Spaces

| Experiment | Sample Space |
|-----------|-------------|
| Flip one coin | {H, T} |
| Roll one die | {1, 2, 3, 4, 5, 6} |
| Flip two coins | {HH, HT, TH, TT} |
| Draw a card | {A♠, 2♠, ..., K♣} (52 outcomes) |

---

# Events

- An **event** is a collection of outcomes we care about
- Example: Roll a die; event A = "roll an even number"
  - A = {2, 4, 6}
  - P(A) = 3/6 = 0.5

--

- **Simple event**: a single outcome
- **Compound event**: two or more outcomes

---

# Venn Diagrams

- Venn diagrams illustrate relationships between events and sample spaces
- The rectangle = sample space S
- Circles = events A, B

<!-- TODO: add [Venn diagram showing sample space S with events A and B overlapping] image -->

---

class: middle, center, inverse

# Probability Rules

---

# Rule 1: Complement Rule

- The **complement** of event A (written $A^c$ or $\bar{A}$) = "A does not occur"
- **Complement Rule:**

$$P(A^c) = 1 - P(A)$$

--

- Example: P(rolling a 6) = 1/6, so P(not rolling a 6) = 1 − 1/6 = 5/6

---

# Rule 2: Addition Rule

**For any two events A and B:**

$$P(A \cup B) = P(A) + P(B) - P(A \cap B)$$

- $A \cup B$ = "A or B (or both)" — the **union**
- $A \cap B$ = "A and B both" — the **intersection**
- We subtract the intersection to avoid **double counting**

---

# Addition Rule: Mutually Exclusive Events

- Two events are **mutually exclusive** (disjoint) if they cannot both occur
  - $P(A \cap B) = 0$
- For mutually exclusive events:

$$P(A \cup B) = P(A) + P(B)$$

--

- Example: Rolling a 2 and rolling a 5 on one die — can't happen at the same time

---

# Addition Rule: Example

Draw one card from a standard deck. What is P(King or Heart)?

- P(King) = 4/52
- P(Heart) = 13/52
- P(King AND Heart) = 1/52 (King of Hearts)

$$P(\text{King or Heart}) = \frac{4}{52} + \frac{13}{52} - \frac{1}{52} = \frac{16}{52} \approx 0.308$$

---

# Rule 3: Multiplication Rule

**For any two events A and B:**

$$P(A \cap B) = P(A) \times P(B \mid A)$$

- $P(B \mid A)$ = "probability of B **given** A has occurred" — **conditional probability**

--

**For independent events:**

$$P(A \cap B) = P(A) \times P(B)$$

---

# Multiplication Rule: Example

Draw two cards **without replacement**. What is P(both Aces)?

- P(1st card is Ace) = 4/52
- P(2nd card is Ace | 1st was Ace) = 3/51

$$P(\text{both Aces}) = \frac{4}{52} \times \frac{3}{51} = \frac{12}{2652} \approx 0.0045$$

---

class: middle, center, inverse

# Conditional Probability

---

# Conditional Probability

- **Conditional probability**: the probability of event A given that event B has already occurred

$$P(A \mid B) = \frac{P(A \cap B)}{P(B)}, \quad P(B) > 0$$

- Conditioning **restricts the sample space** to only outcomes where B occurred

---

# Conditional Probability: Example

In a class of 100 students:
- 60 are women
- 40 are men
- 30 women passed the exam; 20 men passed the exam

What is P(passed | woman)?

$$P(\text{passed} \mid \text{woman}) = \frac{30}{60} = 0.50$$

---

# Conditional Probability: Contingency Table

<!-- TODO: add [2x2 contingency table with rows = passed/failed, columns = women/men, cell counts] image -->

Contingency tables (two-way tables) make conditional probability calculations straightforward:
- Row or column totals define the conditioning event
- Cell counts give the joint counts

---

# Independence

- Two events A and B are **independent** if knowing B occurred gives no information about A:

$$P(A \mid B) = P(A)$$

- Equivalently: $P(A \cap B) = P(A) \times P(B)$

--

- **Independence ≠ mutually exclusive**
  - Mutually exclusive events with P > 0 are *negatively* dependent

---

# Testing Independence

Are A and B independent?
1. Compute P(A), P(B), and P(A ∩ B)
2. Check: does P(A) × P(B) = P(A ∩ B)?
   - If yes: independent
   - If no: dependent

---

# Independence: Example

Flip a fair coin twice. Let:
- A = "first flip is Heads"
- B = "second flip is Heads"

- P(A) = 1/2, P(B) = 1/2
- P(A ∩ B) = P(HH) = 1/4
- P(A) × P(B) = 1/2 × 1/2 = 1/4 ✓ **Independent**

---

class: middle, center, inverse

# Bayes' Theorem

---

# Bayes' Theorem

- **Bayes' theorem** tells us how to update a probability given new evidence

$$P(A \mid B) = \frac{P(B \mid A) \cdot P(A)}{P(B)}$$

- $P(A)$ = **prior probability** (before seeing B)
- $P(A \mid B)$ = **posterior probability** (after seeing B)
- $P(B \mid A)$ = **likelihood** (how probable is B if A is true)

---

# Bayes' Theorem: Full Form

Using the **law of total probability** to expand P(B):

$$P(A \mid B) = \frac{P(B \mid A) \cdot P(A)}{P(B \mid A) \cdot P(A) + P(B \mid A^c) \cdot P(A^c)}$$

---

# Bayes' Theorem: Medical Testing Example

- A disease affects 1% of the population: P(Disease) = 0.01
- A test is 95% sensitive: P(Positive | Disease) = 0.95
- The test has a 5% false-positive rate: P(Positive | No Disease) = 0.05

**If someone tests positive, what is P(Disease | Positive)?**

--

$$P(\text{Dis} \mid +) = \frac{0.95 \times 0.01}{(0.95 \times 0.01) + (0.05 \times 0.99)}$$

$$= \frac{0.0095}{0.0095 + 0.0495} = \frac{0.0095}{0.059} \approx 0.161$$

--

Only ~16%! The **base rate** matters enormously.

---

# Bayes' Theorem: Base Rate Neglect

<!-- TODO: add [2x2 natural frequency table showing 1000 people: 10 with disease, 9 true positive, 1 false negative, 990 without disease, ~50 false positive, ~940 true negative] image -->

- When prevalence (base rate) is low, even good tests produce many false positives
- Visualizing with **natural frequencies** (rather than percentages) helps intuition

---

class: middle, center, inverse

# Probability Distributions

---

# Random Variables

- A **random variable** is a variable whose value is determined by a random experiment
- **Discrete** random variable: countable outcomes (e.g., number of heads in 10 flips)
- **Continuous** random variable: any value in an interval (e.g., height, reaction time)

---

# Probability Distributions

- A **probability distribution** describes all possible values of a random variable and their probabilities
- For discrete RVs: **probability mass function (PMF)**
- For continuous RVs: **probability density function (PDF)**

--

Properties:
- All probabilities ≥ 0
- All probabilities sum (or integrate) to 1

---

# Discrete Distribution: Example

Roll a fair die. Let X = outcome.

| x | 1 | 2 | 3 | 4 | 5 | 6 |
|---|---|---|---|---|---|---|
| P(X=x) | 1/6 | 1/6 | 1/6 | 1/6 | 1/6 | 1/6 |

<!-- TODO: add [bar chart of uniform discrete distribution for a fair die] image -->

---

# Expected Value

- The **expected value** (mean) of a random variable:

$$E(X) = \mu = \sum_x x \cdot P(X = x)$$

- For a fair die: $E(X) = 1(1/6) + 2(1/6) + \ldots + 6(1/6) = 3.5$

--

- E(X) is the **long-run average** — not necessarily a possible outcome

---

# Variance of a Random Variable

$$\text{Var}(X) = \sigma^2 = \sum_x (x - \mu)^2 \cdot P(X = x)$$

$$\text{SD}(X) = \sigma = \sqrt{\text{Var}(X)}$$

---

# The Binomial Distribution

- Models the number of **successes** in *n* independent trials, each with success probability *p*

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$$

- Parameters: n (number of trials), p (probability of success)
- Mean: $\mu = np$
- SD: $\sigma = \sqrt{np(1-p)}$

---

# Binomial: Example

Flip a fair coin 10 times. What is P(exactly 6 heads)?

$$P(X = 6) = \binom{10}{6} (0.5)^6 (0.5)^4 = 210 \times 0.015625 \times 0.0625 \approx 0.205$$

<!-- TODO: add [binomial distribution bar chart for n=10, p=0.5 with k=6 highlighted] image -->

---

class: middle, center, inverse

# The Normal Distribution

---

# The Normal Distribution

- The most important continuous distribution in statistics
- **Bell-shaped**, symmetric, unimodal
- Characterized by two parameters:
  - $\mu$ (mean) — center of the distribution
  - $\sigma$ (standard deviation) — spread

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$$

<!-- TODO: add [normal distribution bell curve with mean and ±1, ±2, ±3 SD marked] image -->

---

# Properties of the Normal Distribution

- Symmetric about the mean: $P(X < \mu) = P(X > \mu) = 0.5$
- Mean = Median = Mode
- The **Empirical Rule** (68–95–99.7 rule):
  - ~68% of data within ±1 SD of mean
  - ~95% of data within ±2 SD of mean
  - ~99.7% of data within ±3 SD of mean

---

# The Empirical Rule

<!-- TODO: add [normal curve with shaded regions showing 68%, 95%, 99.7% areas] image -->

- 68%: $\mu \pm 1\sigma$
- 95%: $\mu \pm 2\sigma$
- 99.7%: $\mu \pm 3\sigma$

--

These percentages are **probabilities** — e.g., a randomly drawn observation has a 95% chance of falling within 2 SDs of the mean.

---

# The Standard Normal Distribution

- The **standard normal** distribution has $\mu = 0$ and $\sigma = 1$
- Written: $Z \sim N(0, 1)$
- Any normal distribution can be converted to standard normal using a **z-score**

$$z = \frac{x - \mu}{\sigma}$$

---

# Z-Scores

- A **z-score** tells you how many standard deviations a value is above or below the mean
- $z > 0$: above the mean
- $z < 0$: below the mean
- $z = 0$: at the mean

--

- Example: IQ scores have $\mu = 100$, $\sigma = 15$. An IQ of 130:
$$z = \frac{130 - 100}{15} = \frac{30}{15} = 2.0$$
  IQ of 130 is 2 SDs above the mean.

---

# Using Z-Scores to Find Probabilities

- After computing z, use a **z-table** (or software) to find $P(Z < z)$
- The table gives the **cumulative probability** — area to the left

<!-- TODO: add [standard normal curve with shaded left tail up to z, z-table excerpt] image -->

--

- $P(Z < 2.0) \approx 0.977$
- So P(IQ < 130) ≈ 97.7%

---

# Normal Distribution: Finding Areas

| Goal | Formula |
|------|---------|
| P(X < a) | Find z, look up table |
| P(X > a) | 1 − P(X < a) |
| P(a < X < b) | P(X < b) − P(X < a) |

---

# Normal Approximation to the Binomial

- When *n* is large and *p* is not too extreme, the binomial distribution is approximately normal:

$$X \sim B(n, p) \approx N(\mu = np,\ \sigma = \sqrt{np(1-p)})$$

- Rule of thumb: use normal approximation when $np \geq 10$ and $n(1-p) \geq 10$

<!-- TODO: add [binomial bar chart overlaid with normal curve showing approximation] image -->

---

# Why the Normal Distribution Is Everywhere

- **Central Limit Theorem (CLT)**: the sampling distribution of the mean approaches normal as sample size increases, regardless of the population distribution
- This is why so many statistical tests assume normality
- Even non-normal populations produce approximately normal sample means for large *n*

---

# Summary: Probability Rules

| Rule | Formula |
|------|---------|
| Complement | $P(A^c) = 1 - P(A)$ |
| Addition (general) | $P(A \cup B) = P(A) + P(B) - P(A \cap B)$ |
| Addition (mutually exclusive) | $P(A \cup B) = P(A) + P(B)$ |
| Multiplication (general) | $P(A \cap B) = P(A) \cdot P(B \mid A)$ |
| Multiplication (independent) | $P(A \cap B) = P(A) \cdot P(B)$ |
| Conditional probability | $P(A \mid B) = P(A \cap B) / P(B)$ |
| Bayes' theorem | $P(A \mid B) = P(B \mid A)P(A) / P(B)$ |

---

class: middle, center, inverse

# Wrapping Up...

---

# Recap

- **Probability** quantifies uncertainty on a 0–1 scale; it can be interpreted classically, empirically, or subjectively
- A **sample space** lists all possible outcomes; **events** are subsets of interest
- Core rules: **complement**, **addition**, and **multiplication**
- **Conditional probability** restricts attention to a subset; events are **independent** when conditioning doesn't change the probability
- **Bayes' theorem** updates prior probabilities given evidence — and base rates matter
- The **normal distribution** is fundamental; **z-scores** standardize values for probability lookup
