---
title: "Sampling"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Sampling

---

class: middle, center

# Roadmap

.pull-left[
**Last Week**
- Study design
- Observational studies
- Experiments
]

.pull-right[
**This Week**
- Sampling
- Sampling methods
- Sampling bias
]

--

**By the end of this lecture, you should be able to:**

- Distinguish between a population and a sample
- Identify and describe common probability sampling methods
- Identify and describe common non-probability sampling methods
- Explain how sampling bias arises and how to avoid it
- Evaluate the quality of a sample with respect to representativeness

---

class: middle, center, inverse

# Populations and Samples

---

# Populations and Samples

- **Population**: the entire group of individuals or objects we want to learn about
- **Sample**: a subset of the population that we actually observe

--

- We use data from a **sample** to make inferences about the **population**
- The goal is for the sample to be **representative** of the population

--

<!-- TODO: add population and sample diagram image -->

---

# Why Sample?

- Studying the entire population is often:
  - Too expensive
  - Too time-consuming
  - Logistically impossible

--

- A well-chosen sample can give us accurate information about the population at a fraction of the cost

--

- **Key question**: How do we choose a good sample?

---

class: middle, center, inverse

# Sampling Methods

---

# Sampling Methods

Two broad categories:

--

- **Probability sampling**: every member of the population has a known, non-zero chance of being selected
  - Simple random sample
  - Stratified sample
  - Cluster sample
  - Multistage sample

--

- **Non-probability sampling**: selection is not based on random chance
  - Convenience sample
  - Voluntary response sample

---

class: middle, center, inverse

# Probability Sampling

---

# Simple Random Sample (SRS)

- Every member of the population has an **equal** chance of being selected
- Each possible sample of size $n$ is equally likely

--

- How to take an SRS:
  1. Number all members of the population
  2. Use a random number generator (or table) to select $n$ individuals

--

<!-- TODO: add simple random sample diagram image -->

---

# Simple Random Sample (SRS)

**Advantages**
- Easy to understand
- Eliminates systematic bias in selection

--

**Disadvantages**
- Requires a complete list of the population (sampling frame)
- May not represent small subgroups well
- Can be expensive if population is spread out geographically

---

# Stratified Sample

- Divide the population into **strata** (groups) based on a shared characteristic
- Take a simple random sample from **each stratum**

--

- Strata are often defined by:
  - Age
  - Sex/gender
  - Race/ethnicity
  - Geographic region

--

<!-- TODO: add stratified sample diagram image -->

---

# Stratified Sample

**Advantages**
- Guarantees representation of all strata
- Can give more precise estimates than SRS if strata are homogeneous

--

**Disadvantages**
- Must know population strata in advance
- More complex to implement than SRS

---

# Cluster Sample

- Divide the population into **clusters** (naturally occurring groups)
- Randomly select some clusters
- Sample **all** (or most) individuals within selected clusters

--

- Clusters are often defined by:
  - Geographic area (e.g., city blocks, schools)
  - Natural groupings (e.g., households)

--

<!-- TODO: add cluster sample diagram image -->

---

# Cluster Sample

**Advantages**
- Cost-effective when population is spread over a large area
- Does not require a complete list of all individuals — only a list of clusters

--

**Disadvantages**
- Individuals within a cluster tend to be similar, which reduces precision
- Higher sampling error than SRS of the same size

---

# Stratified vs. Cluster Sampling

.pull-left[
**Stratified**
- Sample from **every** stratum
- Strata should be **internally homogeneous**
- Reduces variability
]

.pull-right[
**Cluster**
- Sample only **some** clusters
- Clusters should be **internally heterogeneous**
- Convenient and cost-effective
]

--

<!-- TODO: add stratified vs cluster comparison diagram image -->

---

# Multistage Sample

- Combines multiple sampling methods
- Example: randomly select clusters, then randomly select individuals within those clusters

--

- Used in large, complex surveys (e.g., national health surveys)

--

<!-- TODO: add multistage sample diagram image -->

---

class: middle, center, inverse

# Non-Probability Sampling

---

# Convenience Sample

- Select individuals who are **easy to reach** or **readily available**

--

**Examples**
- Surveying students in your own class
- Asking people at a mall
- Using Amazon Mechanical Turk

--

**Problem**: The sample may not be representative of the population of interest

---

# Voluntary Response Sample

- Individuals **choose** to participate (self-select)

--

**Examples**
- Online polls
- Call-in surveys
- Customer reviews

--

**Problem**: People with strong opinions (especially negative ones) are more likely to respond — **voluntary response bias**

---

class: middle, center, inverse

# Sampling Bias

---

# Sampling Bias

**Sampling bias** occurs when the method of selecting the sample produces a sample that is not representative of the population

--

The sample **systematically** differs from the population in some way

--

Result: conclusions drawn from the sample will not generalize to the population

---

# Sources of Sampling Bias

- **Undercoverage**: some groups in the population are underrepresented or excluded from the sampling frame

--

- **Voluntary response bias**: people with strong opinions self-select into the sample

--

- **Convenience sampling**: easy-to-reach individuals may differ systematically from the population

--

- **Non-response bias**: individuals selected for the sample don't respond, and non-responders differ from responders

---

# The Literary Digest Poll (1936)

- *Literary Digest* magazine predicted Alf Landon would beat Franklin Roosevelt in the 1936 presidential election

--

- They mailed **10 million** ballots and received **2.4 million** responses

--

- Roosevelt won in a **landslide**

--

**What went wrong?**
- The mailing list came from telephone directories and car registrations — wealthier, more Republican-leaning individuals
- **Undercoverage** + **voluntary response bias**

<!-- TODO: add Literary Digest 1936 poll image -->

---

# The 1948 Dewey vs. Truman Election

- Polls predicted **Thomas Dewey** would defeat Harry Truman

--

- **Truman won**

--

**What went wrong?**
- Quota sampling was used — interviewers had flexibility in who they interviewed
- Interviewers tended to interview people like themselves (more affluent, Republican-leaning)
- Polling stopped too early — late-deciding voters broke for Truman

<!-- TODO: add Dewey defeats Truman newspaper image -->

---

# Non-response Bias

- Even with a good sampling method, **non-response** can introduce bias

--

- If the people who don't respond differ systematically from those who do, the sample will be biased

--

**Example**: Survey about exercise habits — people who exercise regularly may be more likely to respond

--

- **Solutions**:
  - Follow up with non-respondents
  - Compare respondents to known population characteristics
  - Weight responses to adjust for non-response

---

class: middle, center, inverse

# Sample Size and Variability

---

# Sample Size

- Larger samples tend to produce more precise estimates

--

- But size alone doesn't guarantee a representative sample
  - A large biased sample is still biased
  - A small well-chosen sample can be very informative

--

**"It's not the size of the sample relative to the population that matters — it's the absolute size of the sample."**

---

# Sampling Variability

- Different samples from the same population will give different results

--

- **Sampling variability** (sampling error): the natural variation between samples

--

- With probability sampling, we can **quantify** this variability using statistics
  - Margin of error
  - Confidence intervals

--

- With non-probability sampling, we **cannot** reliably quantify sampling variability

---

# Sampling Distribution

- If we took many samples and computed the sample mean each time, the distribution of those means is the **sampling distribution**

--

- The sampling distribution is centered at the **population mean**

--

- The spread of the sampling distribution decreases as **sample size increases**

--

<!-- TODO: add sampling distribution diagram image -->

---

class: middle, center, inverse

# Toward Good Sampling Practice

---

# What Makes a Good Sample?

- **Random selection**: reduces systematic bias
- **Adequate size**: reduces sampling variability
- **Complete sampling frame**: reduces undercoverage
- **High response rate**: reduces non-response bias
- **Appropriate method**: matches the structure of the population

---

# Putting It All Together

| Method | Random? | Representative? | Cost |
|---|---|---|---|
| Simple random sample | Yes | Usually | Medium |
| Stratified sample | Yes | Yes (by design) | Medium-High |
| Cluster sample | Yes | Often | Low-Medium |
| Multistage sample | Yes | Usually | Medium |
| Convenience sample | No | Often not | Low |
| Voluntary response | No | Often not | Very Low |

---

class: middle, center, inverse

# Wrapping Up...

---

# Summary

- A **sample** is used to make inferences about a **population**
- **Probability sampling** methods use random selection and allow us to quantify uncertainty
  - Simple random, stratified, cluster, multistage
- **Non-probability sampling** methods do not use random selection and are prone to bias
  - Convenience, voluntary response
- **Sampling bias** occurs when the sample systematically differs from the population
- Larger samples reduce variability, but don't fix bias

---

# References

```{r refs, echo=FALSE, results="asis"}
# bibliography will be printed here
```
