---
title: "ANOVA -> Regression <br> `r emo::ji('r')`"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
#       slideNumberFormat: ""  
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
# Installs library if missing
if (!require("HistData")) install.packages("HistData") 
library(emo)
knitr::opts_chunk$set(echo = FALSE,out.width = "90%", fig.align = "center")

```

class: middle

# f-tests and their applications

---

## Roadmap: Last Week 
.large[
- t-tests
- t-tests
- t-tests
]

---

## Roadmap: This Week
.large[
- multiple comparisons
- f-test logic
- analysis of variance
- regression
]

---

class: middle

# Regression


---

# Anova Example



Flower_Lengths <- read.delim("dat/data_anova2.dat")

# describe the data

summary(Flower_Lengths)

# by groups?

library(psych)
describeBy(Flower_Lengths,group=Flower_Lengths$Type)

# Plots
library(ggplot2)
a <- ggplot(Flower_Lengths, aes(factor(Type)))
a + geom_bar()

a + geom_bar(aes(fill=factor(Type)))

a + geom_boxplot(aes(y=Length))

a + geom_density(aes(fill=factor(Type),x=Length), alpha=0.8)

a + geom_density(aes(fill=factor(Type),x=Length), alpha=0.8) +
  labs(title="Density plot",
       subtitle="FLOWER LENGTHS (MM) \n FOR THREE HELICONIA VARIETIES",
       caption="Source: Basic Practice Data",
       x="Length (mm)",
       fill="Types")

a + geom_boxplot(aes(y=Length))+
  geom_dotplot(aes(y=Length),
               binaxis='y',
               stackdir='center',
               dotsize = .5,
               fill="red") +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
  labs(title="Box plot + Dot plot",
       subtitle="FLOWER LENGTHS (MM) FOR THREE HELICONIA VARIETIES",
       caption="Source: Basic Practice Data",
       x="Length (mm)",
       fill="Types")

a + geom_violin(aes(y=Length)) +
  labs(title="Violin plot",
       subtitle="FLOWER LENGTHS (MM) FOR THREE HELICONIA VARIETIES",
       caption="Source: Basic Practice Data",
       x="Length (mm)",
       fill="Types")
# Non-ggplot
plot(Length ~ Type, data = Flower_Lengths)


## multiple t.tests? (NO!!!!!!!!!!)

## Red versus BIHAI
t.test(Flower_Lengths$Length[Flower_Lengths$Type == "RED"],Flower_Lengths$Length[Flower_Lengths$Type == "BIHAI"])

## Red versus YELLOW
t.test(Flower_Lengths$Length[Flower_Lengths$Type == "RED"],Flower_Lengths$Length[Flower_Lengths$Type == "YELLOW"])

## BIHAI versus YELLOW
t.test(Flower_Lengths$Length[Flower_Lengths$Type == "BIHAI"],Flower_Lengths$Length[Flower_Lengths$Type == "YELLOW"])

## Type 1 error rate
.05*3


# REGRESSION
model_1=lm(Length~Type,data=Flower_Lengths)
summary(model_1)

# ANOVA
anova(lm(Length~Type,data=Flower_Lengths))


