---
title: "Surveys"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Surveys

---

class: middle, center

# Roadmap

.pull-left[
**Last Time**
- Observational studies
- Experiments
- Sampling methods
]

.pull-right[
**Today**
- Survey design
- Question wording
- Response bias
- Survey modes
]

--

**By the end of this lecture, you should be able to:**

- Explain how question wording and order affect survey responses
- Identify common sources of response bias in surveys
- Describe different survey modes and their tradeoffs
- Evaluate the quality of survey questions
- Distinguish between sampling bias and response bias

---

class: middle, center, inverse

# What Is a Survey?

---

# What Is a Survey?

- A **survey** is a method of collecting data by asking people questions

--

- Surveys are used in:
  - Political polling
  - Public health research
  - Marketing research
  - Academic social science
  - Government data collection (e.g., Census)

--

- Even with a perfect sample, a poorly designed survey can produce **misleading data**

---

# Two Major Sources of Survey Error

.pull-left[
**Sampling Error**
- Error due to studying a sample rather than the whole population
- Reduced by larger, well-chosen samples
- Can be quantified with probability sampling
]

.pull-right[
**Non-Sampling Error**
- Error not related to sampling
- Includes measurement error, response bias, non-response
- Often harder to detect and correct
]

---

class: middle, center, inverse

# Question Wording

---

# Question Wording Matters

- The way a question is worded can **dramatically** affect responses

--

**Example**: "Should the US spend more money on assistance to the poor?"
- 63% said yes

--

**Example**: "Should the US spend more money on welfare?"
- 23% said yes

--

Same underlying concept — very different responses `r emo::ji("thinking")`

---

# Question Wording: Leading Questions

- A **leading question** pushes respondents toward a particular answer

--

**Example (leading)**: "Don't you agree that the death penalty is barbaric and should be abolished?"

--

**Better**: "Do you support or oppose the death penalty?"

--

- Leading questions introduce **response bias** — responses don't reflect true opinions

---

# Question Wording: Loaded Language

- **Loaded language** uses emotionally charged words that bias responses

--

**Examples**:
- "pro-life" vs. "anti-abortion"
- "undocumented immigrants" vs. "illegal aliens"
- "estate tax" vs. "death tax"

--

- Word choice can shift response rates by 10–20 percentage points or more

---

# Question Wording: Double-Barreled Questions

- A **double-barreled question** asks about two things at once

--

**Example**: "Do you support stricter gun control laws and expanded mental health services?"

--

- Respondent may agree with one part but not the other — what does "yes" mean?

--

**Fix**: Ask as two separate questions

---

# Question Wording: Double Negatives

- **Double negatives** are confusing and lead to inaccurate responses

--

**Example**: "Do you disagree that the policy should not be repealed?"

--

- Even careful readers can misinterpret these

--

**Fix**: Rephrase to avoid negatives when possible

---

# Question Wording: Ambiguous Questions

- **Ambiguous questions** mean different things to different people

--

**Example**: "How often do you exercise regularly?"
- What counts as "exercise"? What counts as "regularly"?

--

**Better**: "In the past week, how many times did you engage in physical activity for at least 30 minutes?"

--

- Specificity reduces ambiguity

---

# Question Wording: Sensitive Questions

- People may not answer honestly about sensitive topics:
  - Drug use
  - Sexual behavior
  - Income
  - Illegal activity
  - Socially undesirable opinions

--

- **Social desirability bias**: tendency to give answers that make you look good

---

# Reducing Social Desirability Bias

- **Anonymous surveys**: respondents feel safer being honest

--

- **Randomized response technique**: respondent flips a coin (privately) — if heads, answer a harmless question; if tails, answer the sensitive question. Researcher can estimate true rate without knowing any individual's answer.

--

- **List experiments**: respondents count the number of items on a list they agree with; the sensitive item is included in one version of the list

--

- **Forgiving wording**: "Some people feel... others feel... What do you think?"

---

class: middle, center, inverse

# Question Order and Format

---

# Question Order Effects

- The order of questions can influence responses — **order effects**

--

- **Context effects**: earlier questions prime respondents to think about a topic in a particular way

--

**Example**: Asking "How happy are you with your life overall?" after asking about recent stressful events produces lower happiness ratings than asking it first

--

- **Recency and primacy effects**: for lists of items, people tend to choose options at the beginning (primacy) or end (recency) of the list

---

# Response Scale Design

- The **format of response options** affects answers

--

- **Number of categories**: 5-point vs. 7-point vs. 11-point scales
  - More points = more fine-grained information, but may overwhelm respondents

--

- **Labeled vs. unlabeled**: labeling only endpoints vs. all points affects interpretation

--

- **Agree/disagree scales** can introduce **acquiescence bias**: tendency to agree regardless of content

--

- **Forced choice** (no "neutral" option) vs. **middle option** inclusion

---

# Acquiescence Bias

- **Acquiescence bias** (yes-saying): tendency to agree with survey statements regardless of content

--

- More common among some demographic groups and in some cultural contexts

--

**Solution**: Balance questions so that agreeing sometimes reflects one opinion and sometimes the opposite

--

**Example**:
- "The government should do more to help the poor" (agree = liberal)
- "People should take responsibility for their own lives" (agree = conservative)

---

class: middle, center, inverse

# Non-Response

---

# Non-Response

- **Unit non-response**: a selected individual doesn't participate at all

--

- **Item non-response**: a participant skips particular questions

--

- Non-response is a problem when non-respondents differ from respondents

--

- Survey response rates have been **declining for decades**
  - Pew Research Center: telephone response rates fell from ~36% in 1997 to ~6% in 2018

<!-- TODO: add response rate decline chart image -->

---

# Non-Response Bias

- If non-responders differ systematically from responders, estimates will be biased

--

**Example**: A survey about health behaviors may attract healthier, more health-conscious people — overestimating healthy behaviors in the population

--

- **Solutions**:
  - Follow up with non-respondents
  - Weight data to match population demographics
  - Use mixed-mode surveys to reach hard-to-reach groups

---

class: middle, center, inverse

# Survey Modes

---

# Survey Modes

Different ways to administer surveys:

--

- **In-person (face-to-face)**
- **Telephone** (landline and cell)
- **Mail (paper)**
- **Online / web-based**
- **Mixed mode**

--

Each has different tradeoffs for cost, coverage, response rates, and data quality

---

# In-Person Surveys

.pull-left[
**Advantages**
- High response rates
- Interviewer can clarify questions
- Can use visual aids
- Can observe non-verbal cues
]

.pull-right[
**Disadvantages**
- Very expensive
- Time-consuming
- Interviewer effects (social desirability)
- Geographic limitations
]

---

# Telephone Surveys

.pull-left[
**Advantages**
- Moderate cost
- Relatively fast
- Can reach geographically dispersed samples
- Random Digit Dialing (RDD) can reach unlisted numbers
]

.pull-right[
**Disadvantages**
- Declining response rates
- Cell phones complicate RDD
- Cannot use visual aids
- Interviewer effects
- Cell-only households may be missed
]

---

# Mail Surveys

.pull-left[
**Advantages**
- Low cost
- No interviewer effects
- Respondents can take their time
- Good for sensitive topics
]

.pull-right[
**Disadvantages**
- Low response rates
- Slow
- No opportunity to clarify questions
- Literacy requirements
- May not reach all groups
]

---

# Online Surveys

.pull-left[
**Advantages**
- Very low cost
- Fast
- No interviewer effects
- Easy to include skip patterns and multimedia
- Can reach large samples
]

.pull-right[
**Disadvantages**
- Coverage issues (not everyone is online)
- Hard to get probability samples
- Self-selection
- Satisficing (rushing through)
- Cannot verify respondent identity
]

---

# Mode Effects

- The **mode** of survey administration can affect responses

--

- **Social desirability**: higher in interviewer-administered surveys (phone, in-person)

--

- **Coverage**: online surveys miss those without internet access; telephone surveys miss cell-only households

--

- **Response rates** vary substantially by mode

--

- **Mixed-mode** surveys try to combine advantages of multiple modes

---

class: middle, center, inverse

# Probability vs. Non-Probability Online Panels

---

# Online Survey Panels

- Many surveys now use **online panels**: pools of pre-recruited respondents

--

- **Probability-based panels**: recruited via probability sampling (e.g., address-based sampling)
  - AmeriSpeak (NORC), KnowledgePanel (Ipsos/formerly Knowledge Networks)

--

- **Non-probability panels**: recruited via opt-in (e.g., ads, websites)
  - Much cheaper, faster
  - But unknown representativeness

--

- Research shows probability-based panels tend to produce more accurate estimates, but non-probability panels can perform well with careful weighting

---

# Weighting

- **Weighting** adjusts survey data to better represent the population

--

- **Design weights**: account for unequal probabilities of selection

--

- **Post-stratification weights**: adjust sample to match population on key demographics (age, sex, race, education, etc.)

--

- Weighting can reduce bias but cannot fix all problems
  - If important characteristics are not weighted, bias remains
  - Extreme weights increase variance

---

class: middle, center, inverse

# Evaluating Survey Quality

---

# Evaluating Survey Quality

When reading about survey results, ask:

--

- **Who** was surveyed? Is the sample representative of the population of interest?

--

- **How** were they selected? Probability or non-probability?

--

- **How many** responded? What was the response rate?

--

- **How** were questions worded? Could wording have influenced responses?

--

- **Who** sponsored the survey? Could there be a conflict of interest?

---

# Red Flags in Survey Reporting

- No information about sampling methodology
- Very low or unreported response rate
- Leading or loaded question wording
- Results based on voluntary response (online polls, call-in surveys)
- Sponsor with a stake in the outcome
- Small sample sizes for subgroup analyses

--

**"Garbage in, garbage out"** — even sophisticated analysis cannot fix a fundamentally flawed survey

---

# A Good Survey Question Is...

- **Clear**: unambiguous, one interpretation
- **Specific**: concrete, not vague
- **Neutral**: no leading or loaded language
- **Single-barreled**: asks one thing at a time
- **Answerable**: respondents actually know the answer
- **Relevant**: respondents have an opinion or experience to report

---

class: middle, center, inverse

# Wrapping Up...

---

# Summary

- Surveys are a powerful data collection tool — but design matters enormously
- **Question wording** can dramatically affect responses: avoid leading questions, loaded language, double-barreled questions, and ambiguity
- **Question order** creates context effects that influence responses
- **Non-response** is a major challenge; declining response rates require careful attention to non-response bias
- **Survey mode** affects cost, coverage, response rates, and data quality
- **Weighting** can reduce bias but cannot fix all problems
- Always critically evaluate survey methodology before interpreting results

---

# References

```{r refs, echo=FALSE, results="asis"}
# bibliography will be printed here
```
