---
title: "Excel Lab: Z-Scores and Normal Distribution"
author: "S. Mason Garrison"
bibliography: library.bib
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(openxlsx)
library(magick)  # For image annotations
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle

# Excel Lab: Z-Scores and Normal Distribution

---

## Download the Dataset

We will use the famous **Iris** dataset for this lab. You’ll use the `Sepal.Length` column for Z-scores and percentile calculations.

```{r}
# Load Iris dataset and select Sepal.Length
iris_data <- iris %>% select(Sepal.Length)

# Save the dataset as an Excel file
write.xlsx(iris, "../slides/slides/t08_lab/IrisData.xlsx")
```

[Download the Excel file here](IrisData.xlsx)

---

## Step 1: Calculate the Mean in Excel

1. Open the Excel file `IrisData_ZScores.xlsx`.
2. **Calculate the Mean** for the `Sepal.Length` column using the following formula in Excel:
   - `=AVERAGE(A2:A151)`
   
![Insert Screenshot of Mean Calculation here]

```{r, eval=FALSE}
image_path <- "path_to_mean_screenshot.png"
img <- image_read(image_path) %>%
  image_annotate("Calculate Mean: =AVERAGE(A2:A151)", size = 20, color = "red", location = "+50+50")
img
```

---

## Step 2: Calculate the Standard Deviation in Excel

1. **Calculate the Standard Deviation** for the `Sepal.Length` column in Excel:
   - Use the formula:  
   `=STDEV.S(A2:A151)`

![Insert Screenshot of Std Dev Calculation here]

```{r, eval=FALSE}
image_path_sd <- "path_to_stddev_screenshot.png"
img_sd <- image_read(image_path_sd) %>%
  image_annotate("Calculate Std Dev: =STDEV.S(A2:A151)", size = 20, color = "blue", location = "+50+50")
img_sd
```

---

## Step 3: Calculate Z-Scores in Excel

1. **Calculate Z-Scores** for each value in the `Sepal.Length` column using:
   - Formula: `=(A2 - [Mean]) / [Standard Deviation]`
   - Drag down to fill for all rows.

![Insert Screenshot of Z-Score Calculation here]

```{r, eval=FALSE}
image_path_zscore <- "path_to_zscore_screenshot.png"
img_zscore <- image_read(image_path_zscore) %>%
  image_annotate("Calculate Z-Scores: =(A2 - Mean) / Std Dev", size = 20, color = "green", location = "+50+50")
img_zscore
```

---

## Step 4: Calculate Percentiles in Excel

1. **Use the `NORM.S.DIST()` function** to calculate cumulative probabilities:
   - Formula: `=NORM.S.DIST([Z-Score Cell], TRUE)`
   - Drag down to fill for all Z-scores.
   
2. **Interpret the result**: These values represent the percentile rank of each `Sepal.Length` value.

![Insert Screenshot of Percentile Calculation here]

```{r, eval=FALSE}
image_path_percentile <- "path_to_percentile_screenshot.png"
img_percentile <- image_read(image_path_percentile) %>%
  image_annotate("Calculate Percentile: =NORM.S.DIST(Z-Score, TRUE)", size = 20, color = "red", location = "+50+50")
img_percentile
```

---

## Conclusion

In this section, you:
- Calculated Z-scores in Excel.
- Used Excel functions to calculate cumulative probabilities and percentiles.

Be sure to save your Excel file with Z-scores and percentiles for future reference in this lab.

---

class: center, middle

# Wrapping Up
``` 
