---
title: "Research Process and Way of Thinking"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---


```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle

# Research Process and Way of Thinking

---

## Roadmap

.pull-left[
`r emo::ji("heartbeat")` **Joy of Statistics**
]

.pull-right[
`r emo::ji("mag")` **Research Process**

- Steps in a Research Study
- Asking Good Questions
- Operationalize
]

--

**By the end of this lecture, you should be able to:**

- Explain why statistics matter for critical thinking
- List the key steps in a research study
- Distinguish a research problem from a research question
- Define operationalization and write a testable hypothesis

---

class: middle, center, inverse

# The Joy of Statistics

---

## What's the Joy in Statistics, Anyway?

.pull-left[
`r emo::ji("globe_with_meridians")`

**Statistics help us understand and interpret the world around us**
]

.pull-left[
`r emo::ji("brain")`

**They provide insights into complex phenomena**
]

.pull-right[
`r emo::ji("bar_chart")`

**Statistics empower us to make informed decisions**
]

---

## The Power of Visual Statistics

.pull-left[
**The Power of Visual Statistics**
]

.pull-right[
- **Crimespotting (Stamen Design)**
  - Visualizes relationship between topography and crime
  - Empowers citizens with information for accountability
  - Showed the "promise of visualizations of public data"
- **Polar Area Graph (Florence Nightingale)**
  - Illustrated deaths from preventable diseases in hospitals
  - Sparked a revolution in hospital hygiene
- **Billion Pound-O-Gram (David McCandless)**
  - Helps comprehend large numbers in news and government reports
  - Puts financial figures into perspective
- **"We Feel Fine" Project (Jonathan Harris & Sep Kamvar)**
  - Explores human emotion on a global scale
  - Uncovers trends in collective feelings
]

---

## The Power of Visual Statistics

.pull-left[
**The Power of Visual Statistics**
]

.pull-right[
- **Crimespotting (Stamen Design)**
  - Visualizes relationship between topography and crime
  - Empowers citizens with information for accountability
  - Showed the "promise of visualizations of public data"

<!-- TODO: add Crimespotting screenshot -->
]

---

## The Power of Visual Statistics

.pull-left[
<!-- TODO: add Florence Nightingale Polar Area diagram -->
]

.pull-right[
- **Crimespotting (Stamen Design)**
- **Polar Area Graph (Florence Nightingale)**
  - Illustrated deaths from preventable diseases in hospitals
  - Sparked a revolution in hospital hygiene
- **Billion Pound-O-Gram (David McCandless)**
  - Helps comprehend large numbers in news and government reports
  - Puts financial figures into perspective
- **"We Feel Fine" Project (Jonathan Harris & Sep Kamvar)**
  - Explores human emotion on a global scale
  - Uncovers trends in collective feelings
]

---

## The Power of Visual Statistics

.pull-left[
<!-- TODO: add Billion Pound-O-Gram screenshot -->

[informationisbeautiful.net/visualizations/the-billion-pound-o-gram](https://informationisbeautiful.net/visualizations/the-billion-pound-o-gram/)
]

.pull-right[
- **Crimespotting (Stamen Design)**
- **Polar Area Graph (Florence Nightingale)**
- **Billion Pound-O-Gram (David McCandless)**
  - Helps comprehend large numbers in news and government reports
  - Puts financial figures into perspective
- **"We Feel Fine" Project (Jonathan Harris & Sep Kamvar)**
  - Explores human emotion on a global scale
  - Uncovers trends in collective feelings
]

---

## The Power of Visual Statistics

.pull-left[
[jjh.org/we-feel-fine](https://jjh.org/we-feel-fine)

<!-- TODO: add "We Feel Fine" screenshot -->
]

.pull-right[
- **Crimespotting (Stamen Design)**
  - Visualizes relationship between topography and crime
  - Empowers citizens with information for accountability
  - Showed the "promise of visualizations of public data"
- **Polar Area Graph (Florence Nightingale)**
  - Illustrated deaths from preventable diseases in hospitals
  - Sparked a revolution in hospital hygiene
- **Billion Pound-O-Gram (David McCandless)**
  - Helps comprehend large numbers in news and government reports
  - Puts financial figures into perspective
- **"We Feel Fine" Project (Jonathan Harris & Sep Kamvar)**
  - Explores human emotion on a global scale
  - Uncovers trends in collective feelings
]

---

## Data Are Everywhere

.pull-left[
<!-- TODO: add code/data background image -->

**DATA ARE EVERYWHERE**

- We're in a "data-driven" age, right?
- But how do we "consume" these data correctly?
  - How do we get meaning from these data?
- **Statistic**
  - a number computed from a sample of data
]

---

## Statistics Are Everywhere

.pull-left[
<!-- TODO: add blue sidebar image -->

**STATISTICS ARE EVERYWHERE**
]

.pull-right[
- Education
  - SAT, GPA
- Medicine
  - Blood Cell Count, Cholesterol, Weight
- Public Health
  - Positivity Rates
  - Case Counts
- Sports
  - Baseball, Cricket
- Politics
  - Polling
]

---

## Statistics Can "Lie"

.pull-left[
<!-- TODO: add "Where We Donate vs. Diseases That Kill Us" infographic (Vox) -->
]

.pull-right[
**STATISTICS CAN "LIE"**

- The average individual has one testicle and one ovary
- Statistics can be used to mislead people
  - Money and happiness
- "Lie?"
]

---

class: middle

.pull-left[
# Steven Pinker
]

.pull-right[
<br><br>
> "..if narratives without statistics are blind, statistics without narratives are empty."

— Steven Pinker
]

---

class: middle

.pull-left[
<!-- TODO: add question marks / magnifying glass image -->

# QUESTION
]

.pull-right[
**What do you gain from being a critical consumer of research?**
]

---

## Critical Thinking Starts with Statistics

.pull-left[
<!-- TODO: add colorful question marks sticky notes image -->
]

.pull-right[
What I, in my complacency, chose to ignore is just how much of the
persuasion now aimed at the average citizen comes in the form of
numbers, specifically numbers that tell us about the future, about how
likely something is to happen (or not happen) based on how much it
happened (or didn't) in the past. These numbers sing to us the siren
song of cause and effect, humanity's favorite tune. Why do we like it so
much? **Because knowing what causes events and conditions is
the first step toward controlling them, and we human beings
are all about controlling our environments.** That's how we ended
up ruling this planet, and it's how some of us hope to save it.

.small[Laura Miller, *Slate*, August 31st 2015]
]

---

class: middle, center, inverse

# Psychology! `r emo::ji("brain")`

---

class: middle

# Psychology as a Way of Thinking

---

## Why Is Psychology So Popular?

.pull-left[
<!-- TODO: add brain/head icon image -->

# WHY IS PSYCHOLOGY SO POPULAR?
]

.pull-right[
<br><br><br>
Concerned with understanding the way people behave, feel, and think
]

---

## Popular Themes

.pull-left[
**POPULAR THEMES**
]

.pull-right[
- free will and determinism
- nature and nurture
- love and loss
- personality (growth and disorder)
- justice and violations
]

---

## What Do Scientists Do?

.pull-left[
Scientists are **empiricists.**

`r emo::ji("arrow_right")`

Scientists **tackle applied and basic problems.**

`r emo::ji("arrow_right")`
]

.pull-right[
Scientists **test theories:** the theory-data cycle.

`r emo::ji("arrow_right")`

Scientists **make their work public.**

`r emo::ji("arrow_right")`

Scientists **work in a community.**

`r emo::ji("arrow_down")`

Scientists **talk to the world:** from journal to journalism.
]

---

class: middle, center

<!-- TODO: add "The Science News Cycle" PhD Comics image (Figure 1.9, PHD Comics / W.W. Norton) -->

.footnote[PHD Comics. Copyright © 2021 W. W. Norton & Co., Inc.]

---

class: middle, center, inverse

# Wrapping Up...

---

class: middle

.pull-left[
<!-- TODO: add clock/time image -->

**THIS TIME...**

**RESEARCH PROCESS**
]

---

class: middle, center, inverse

# Research Process

---

## Steps in a Research Study

.pull-left[
<!-- TODO: add "Every baby knows the scientific method!" illustration -->
]

.pull-right[
**STEPS IN A RESEARCH STUDY**

- Question
- Design
- Sampling
- Measurement
- Summary
]

---

class: middle

# Science Starts With a Question `r emo::ji("question")`

<!-- TODO: add curious cats with question marks image -->

---

## Asking a Good Question

.pull-left[
`r emo::ji("mag_right")` **Research Problem**
]

.pull-right[
`r emo::ji("question")` **Research Question**

`r emo::ji("gear")` **Operationalize**

`r emo::ji("droplet")` **Hypothesize**
]

---

## Good Questions Begin With Problems

.pull-left[
<!-- TODO: add magnifying glass / yellow background image -->

**GOOD QUESTIONS BEGIN WITH PROBLEMS**

- Research problems are broad issues or situations that we'd like to
  - describe,
  - understand, or
  - change.
]

---

## Research Problems

Broadly, these include:

- Areas of concern,
- Conditions that could be improved,
- Difficulties that need to be eliminated, and
- Questions seeking answers.

---

## 'Top Ten' Problems in Social Science

- How can we induce people to look after their health?
- How do societies create effective and resilient institutions, such as governments?
- How can humanity increase its collective wisdom?
- How do we reduce the 'skill gap' between black and white people in America?

.footnote[Giles (2011) *Nature*]

---

## 'Top Ten' Problems in Social Science

- How can we aggregate information possessed by individuals to make the best decisions?
- How can we understand the human capacity to create and articulate knowledge?
- Why do so many female workers still earn less than male workers?
- How and why does the 'social' become 'biological'?

.footnote[Giles (2011) *Nature*]

---

## 'Top Ten' Problems in Social Science

- How can we be robust against 'black swans' — rare events that have extreme consequences?
- Why do social processes, in particular civil violence, either persist over time or suddenly change?

.footnote[Giles (2011) *Nature*]

---

## Research Questions

.pull-left[
`r emo::ji("mag_right")`

**A research problem leads to a research question**
]

.pull-right[
`r emo::ji("teacher")`

**Questions should in some way...**

- Be worth investigating,
- Contribute knowledge to the field,
- Improve educational practice, or
- Improve the human condition.
]

---

## Asking Good Questions

.pull-left[
**ASKING GOOD QUESTIONS**
]

.pull-right[
| Step | Action |
|------|--------|
| **Begin by** | Asking several questions about a topic. |
| **Eliminate** | Questions that cannot be answered by direct observation or by gathering evidence. |
| **Break** | Broad questions into smaller questions that can be investigated one at a time. |
]

---

## Research Questions

.pull-left[
<!-- TODO: add cat doing research cartoon ("Curiosity didn't kill this cat, but doing on-line research used up most of her 9 lives") -->

**RESEARCH QUESTIONS**
]

.pull-right[
Characteristics of a good research question:

- The question is **feasible**.
- The question is **clear**.
- The question is **meaningful**.
- The question is **ethical**.
]

---

## Asking Scientific Questions

.pull-left[
**ASKING SCIENTIFIC QUESTIONS**
]

.pull-right[
`r emo::ji("busts_in_silhouette")` "What is the relationship between . . . ?"

`r emo::ji("brain")` "What factors cause . . ?"

`r emo::ji("brain")` "What is the effect of . . ?"
]

---

## Operationalize

**Operationalization** is a process of defining the fuzzy concepts; and making the concept(s)
- distinguishable,
- measurable, and
- understandable in terms of observable data.

- Determine which variables you would like to understand.
- You will want to be specific.

---

## Example: Personal Space

.pull-left[
- **Personal Space**
  - Public Space (25 ft)
  - Social Space (12 ft)
  - Personal Space (4 ft)
  - Intimate Space (1.5 ft)
- We'll delve deeper into measuring later in the class
]

.pull-right[
<!-- TODO: add concentric circles diagram of personal space zones -->
]

---

## Hypothesizing

.pull-left[
`r emo::ji("speech_balloon")`

**A hypothesis is a**

precise, testable statement of what the researchers predict will be the outcome of the study.
]

.pull-right[
`r emo::ji("question")`

**In other words:**

What do you think the answer is,
- given what we already know?
]

---

## Example

.pull-left[
**Research Problem:**

What are the causes of health inequity?
]

.pull-right[
**Research Question:**

Specifically, what is the effect of social stress on health, among naturally anxious individuals?
]

---

## Example

- **Operationalize:**
  - Stress: Entering Person's Intimate Space (1.5 ft)
  - Health: Self-report ailment checklist
  - Naturally Anxious: Self-reported scores of Neuroticism (Top 20%)
- **Hypothesis:**
  - Entering a person's intimate space will cause individuals to check more ailments on the self-report checklist.

---

class: middle, center, inverse

# Wrapping Up... `r emo::ji("tada")`
