---
title: "Statistical Thinking: Back to Basics"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Statistical Thinking

### Back to Basics

---

## Learning Goals

- Define basic elements of a statistical investigation.

- Describe the role of p-values and confidence intervals in statistical inference.

- Describe the role of random sampling in generalizing conclusions from a sample to a population.

- Describe the role of random assignment in drawing cause-and-effect conclusions.

- Critique statistical studies.

---

## Roadmap

.pull-left[
**Introduction**
- Key Components to a Statistical Investigation

**Distributional Thinking**

**Statistical Significance**
- Control, Probability, Level of Significance

**Generalizability**
- Samples and Populations, Random Sample, Margin of Error

**Cause and Effect**
- Statistical Tendency, Random Assignment
]

--

**By the end of this lecture, you should be able to:**
- Identify the key components of a statistical investigation
- Explain distributional thinking and interpret data displays critically
- Describe what statistical significance means and how p-values are used
- Distinguish between generalizability and cause-and-effect conclusions
- Evaluate the role of random sampling and random assignment in research

---

## Introduction

.pull-left[
- **6 cups of coffee:**
  - Men `r emo::ji("arrow_right")` 10% lower chance of dying
  - Women `r emo::ji("arrow_right")` 15% lower chance of dying

- Does this mean you should start drinking coffee or increase your own coffee habit?
]

.pull-right[
<!-- TODO: add image of coffee beans and espresso cup -->
]

---

## Key Components to a Statistical Investigation

<!-- TODO: add image of four-step flow diagram: Planning the study -> Examining the data -> Inferring from the data -> Drawing Conclusions -->

- Planning the study
- Examining the data
- Inferring from the data
- Drawing Conclusions

---

## Roadmap

**Introduction**
- Key Components to a Statistical Investigation

**.orange[Distributional Thinking]**

**Statistical Significance**
- Control, Probability, Level of Significance

**Generalizability**
- Samples and Populations, Random Sample, Margin of Error

**Cause and Effect**
- Statistical Tendency, Random Assignment

---

## Distributional Thinking

.pull-left[
- Analyzing the pattern of data variation, called the *distribution of the variable*, often reveals insights.

- Develop an example of a distribution, of one variable, that you often encounter in your life.

- What is the variable and how does it vary?
]

.pull-right[
<!-- TODO: add image of six close-up face photos showing a range of expressions (distributional variation in human faces) -->
]

---

## Distributional Thinking

<!-- TODO: add image of frequency tables comparing patient reading levels (n=63) and pamphlet readability levels (n=30), and a bar chart (Figure 1) showing overlapping distributions of patients vs pamphlets by reading level -->

**Table 1.** Frequency tables of patient reading levels and pamphlet readability levels.

**Figure 1.** Comparison of patient reading levels and pamphlet readability levels.

---

## Collect and Display Data

.pull-left[
- **How many concerts did you go to this past year?**
  - Write your answer down and turn it in

- **What is the best way to visually represent the information?**

- **What single score best represents the data?**

- **How much variability is in the data?**
]

.pull-right[
<!-- TODO: add image of concert crowd with hands raised and lights -->
]

---

## Distributional Thinking

**Tables and graphs are often presented in mass media; sometimes accurately, sometimes not.**

<!-- TODO: add two bar charts side by side: "WRONG WAY" (y-axis starting at 6.00, Group A=7.50 vs Group B=6.50 looks exaggerated) and "RIGHT WAY" (y-axis starting at 0, same values look more comparable) -->

1. What should you look for when examining this data to interpret it correctly?
2. How are people able to manipulate visual depictions of statistics to skew conclusions?

---

## Distributional Thinking

.pull-left[
- Tables and graphs are often presented in mass media; sometimes accurately, sometimes not.
  - What should you look for when examining this data to interpret it correctly?
  - How are people able to manipulate visual depictions of statistics to skew conclusions?
]

.pull-right[
<!-- TODO: add bar charts comparing WRONG WAY (truncated y-axis 6.00-8.00) vs RIGHT WAY (y-axis from 0) for Group A (7.50) and Group B (6.50) -->
]

---

## Roadmap

**Introduction**
- Key Components to a Statistical Investigation

**Distributional Thinking**

**.orange[Statistical Significance]**
- Control, Probability, Level of Significance

**Generalizability**
- Samples and Populations, Random Sample, Margin of Error

**Cause and Effect**
- Statistical Tendency, Random Assignment

---

## Statistical Significance

.pull-left[
- **Control**
  - Are there other variables in the infant study that the researchers missed?

- **Probability / p-value**
  - P-Value extravaganza

- **Level of Significance**
  - $p < .05$
]

.pull-right[
<!-- TODO: add image of a mother bear with cub nestled in rocks -->
]

---

## Roadmap

**Introduction**
- Key Components to a Statistical Investigation

**Distributional Thinking**

**Statistical Significance**
- Control, Probability, Level of Significance

**.orange[Generalizability]**
- Samples and Populations, Random Sample, Margin of Error

**Cause and Effect**
- Statistical Tendency, Random Assignment

---

class: middle, center

# Generalizability:
# Samples and Populations

<!-- TODO: add image of a diverse group of smiling young people/students representing a sample from a population -->

---

## Generalizability

.pull-left[
<!-- TODO: add image of a city street/skyline representing a large population -->
]

.pull-right[
- How similar does a sample need to be to the population?

- Can you generalize from one class to the whole grade?

- Random Sample

- Margin of Error
]

---

## Roadmap

**Introduction**
- Key Components to a Statistical Investigation

**Distributional Thinking**

**Statistical Significance**
- Control, Probability, Level of Significance

**Generalizability**
- Samples and Populations, Random Sample, Margin of Error

**.orange[Cause and Effect]**
- Statistical Tendency, Random Assignment

---

## Cause and Effect

.pull-left[
**Statistical Tendency**

<!-- TODO: add image of a large crowd of diverse people packed closely together -->
]

.pull-right[
**Random Assignment**

<!-- TODO: add image of a large conference crowd/group photo showing many people standing together -->
]

If the p-value is small (under .05), the observed results were (probably) not coincidental

---

## CAT: The Muddiest Point `r emo::ji("cat")`

- What was the muddiest point about today's class?

- Write down what concept you are still struggling to understand.

<!-- TODO: add image of black cat silhouette sitting -->

---

class: middle, center, inverse

# Wrapping Up...
