---
title: "More Applications in R"
date: '`r format(Sys.time(), "%B %d, %Y")`'
bibliography: library.bib
output:
  slidy_presentation:
    css: teaching.css
    fig_caption: yes
    font_adjustment: 2
    highlight: pygments
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,out.width="75%", fig.align = "center",fig.retina = 2,
                      tidy.opts=list(width.cutoff = 120),  # For code
                      options(width = 120), # For output)
                      message=FALSE, warning=FALSE)  
options(digits = 3)
```

# Announcement
- The Final Exam will be take home.
    - Open Notes, Open Book, Open Internet, Open Computer
    - You can't consult with classmates.
    - But you can ask me questions.
        - I will have extended office hours.
- The exam will be made available on Friday December 7th at 5pm (through brightspace)
- It will be due on brightspace on Saturday December 15th at 7:00 pm.
    - Grades are due on Dec 17, 2018 — 11:59 p.m.


# Design
- We have selected our question 
    - and operationalized it.
    
- Now we have to decide on our design
- Whether we have our data already influences this stage.
- If we're collecting data,
    - We have to decide on a broad design:
        - Will is be experimental? 
        - Quasi-experimental? 
        - Observational?
        
***

- If we're not collecting our own data,
    - It's a bit trickier.
        - We're limited by the data we have.
    - It's a bit of an art...
        - We can approach multiple ways...
            - Start with your idealized design 
                - and see how close you can get to that design with data that are available
            - Start with the data you have, and develop your design
            - Start with a statistical test in mind... etc
            - Start with a model...
            
# Sampling

- If we're collecting our own data,
    - then we need to think about how we select participants.
    - And how those participants represent the population we're interested in.
    
***
- If we're using pre-existing data,
    - then we need to figure out how these data were selected
    - and how those data represent the population we're interested in.
    - Consider:
        - Using a subset of the data
    
# Measurement
- If we're collecting our own data,
    - then we have control over how are data are collected
        - which measures
        - what order they're presented in, etc

***
- If we're using pre-existing data,
    - then we need to figure out what was collected, 
    - how it was collected
    - are we limited to scale scores 
        - or can we have item-level information

# Summary
- Regardless of whether we collected our own data or used pre-existing data -- the process is similar (if not identical)
- We must decide how to summarize our findings
    - Which descriptive statistics do we use?
        - Robust or Classic?
    - Which plots?
        - scatter, pie, time, etc
    - Which statistical tests?
        - t-test? anova? regression? etc


# Illustration
- Question: Does seatbelt legislation reduce car-related deaths?
- Data: Seatbelts dataset from the datasets library
    - Source: Harvey, A. C. and Durbin, J. (1986). The effects of seat belt legislation on British road casualties: A case study in structural time series modelling. Journal of the Royal Statistical Society series A, 149, 187–227. doi: 10.2307/2981553.
    

- Operationalization:
    - Seatbelt legislation: The presence or absence of compulsory wearing of seatbelts in the UK (started 31 Jan 1983) by month
    - Deaths:
        - Drivers Killed per month
        - Van Drivers Killed per month
        - (Passengers are recorded as killed or seriously injured)

- Hypothesis:
    - Null: Average monthly deaths don't vary whether the law is in effect
        - $H_{0}$: $\mu_{prelaw}=\mu_{postlaw}$
    - Alternative: The average monthly deaths are lower when the law is in effect.
        - $H_{1}$: $\mu_{prelaw}>\mu_{postlaw}$

- Statistical Test
    - two sample t-test

```{r,include=TRUE, echo=TRUE}

library(datasets)
data("Seatbelts")

df=as.data.frame(Seatbelts)

# Summary Statistics
summary(df)

# by groups?

library(psych)
describeBy(df,group=df$law)

# t-test

## directional
t.test(DriversKilled~law, data=df,alternative="greater",var.equal = TRUE)

## directional (wrongway
t.test(DriversKilled~law,data=df,alternative="less",var.equal = TRUE)

## basic plot
plot(DriversKilled ~ law, data = df)
```

- Verdict: Reject the Null
- Question: Is it causal?
    - Why or why not?
- Question: What third variables might be involved?


# Additional Analyses

- Findings from the paper (Harvey and Durbin, 1986)
- "Overall, for the twenty three month period from February 1983 to December 1984, 
- we estimate the reduction due to the law in numbers [of killed or severely injured] for drivers and front seat passengers as 15,600. 
- The estimated reduction in numbers killed is 879."

```{r jarss, echo=TRUE}

## work with pre-seatbelt period to identify a model, use logs
work <- window(log10(UKDriverDeaths), end = 1982+11/12)
#par(mfrow = c(3, 1))
plot(work)#; acf(work); pacf(work) # estimates autocovariance across time
par(mfrow = c(1, 1))
(fit <- arima(work, c(1, 0, 0), seasonal = list(order = c(1, 0, 0))))
z <- predict(fit, n.ahead = 24)
ts.plot(log10(UKDriverDeaths), z$pred, z$pred+2*z$se, z$pred-2*z$se,
        lty = c(1, 3, 2, 2), col = c("black", "red", "blue", "blue"))

## now see the effect of the explanatory variables
X <- Seatbelts[, c("kms", "PetrolPrice", "law")]
X[, 1] <- log10(X[, 1]) - 4
arima(log10(Seatbelts[, "drivers"]), c(1, 0, 0),
      seasonal = list(order = c(1, 0, 0)), xreg = X)
```

# In class examples
- Using a dataset from the dataset library
- We will document these processes in class, 
    - and I'll upload the resulting .R files afterwards

# R Example
```{r echo=TRUE, include=TRUE}

library(HistData)
data(DrinksWages)

DrinksWages[1,]

summary(DrinksWages)

DrinksWages$trade

#cor(DrinksWages)
## Error in cor(DrinksWages) : 'x' must be numeric

cor(DrinksWages$drinks,DrinksWages$wage)

lm(data=DrinksWages, wage~drinks)
```