---
title: "Causation"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---


```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle

# Causation

---

## Roadmap

.pull-left[
`r emo::ji("warning")` **Correlation vs. Causation**

- Post hoc fallacy
- Necessary vs. sufficient

`r emo::ji("microscope")` **Criteria for Causation**

- Hill's Criteria
- Cook and Campbell (1979)
]

.pull-right[
`r emo::ji("shield")` **Minimizing Threats to Validity**

- By argument
- By measurement or observation
- By analysis
- By preventive action
- By design

`r emo::ji("pencil")` **Design Construction**

- Four basic elements
- Expanding designs
]

--

**By the end of this lecture, you should be able to:**

- Explain why correlation does not imply causation and name the types of alternative explanations
- Distinguish between necessary and sufficient conditions
- List and apply Hill's Criteria of Causation
- Describe Cook and Campbell's three criteria for causation
- Explain the five approaches for minimizing threats to validity
- Identify the four basic elements of research design and explain how to expand each

---

class: middle, center, inverse

# Correlation `r emo::ji("right_arrow")` Causation

---

## Correlation `r emo::ji("right_arrow")` Causation

The adage that "Correlation does not imply Causation" is a warning against

**Post hoc ergo propter hoc**
- "After this, therefore because of this"
- Specific case of the Questionable Cause Fallacy (Logical Fallacy)

--

Sometimes called:
- causal fallacy
- false cause, or
- 'non causa pro causa' — "Not the cause for the cause"

---

## Correlation `r emo::ji("right_arrow")` Causation

We cannot assume that correlation means causation. Possible explanations include:

--

- **A causes B** (direct causation)

--

- **B causes A** (reverse causation)

--

- **A and B are consequences of a common cause**, but do not cause each other

--

- **A causes B and B causes A** (bidirectional or cyclic causation)

--

- **A causes C which causes B** (indirect causation)

--

- **There is no connection between A and B**; the correlation is a coincidence

---

## Correlation `r emo::ji("right_arrow")` Causation

.pull-left[
> "Correlation doesn't imply causation, but it does waggle its eyebrows suggestively and gesture furtively while mouthing *look over there*."

— [xkcd.com/552](https://xkcd.com/552/)
]

.pull-right[
<!-- TODO: add xkcd #552 "Correlation" comic -->

Correlations are **necessary** for causation, but not **sufficient**.
]

---

class: middle, center, inverse

# Necessary Versus Sufficient (Logic)

---

## Necessary Versus Sufficient (Logic)

**Necessary** (if A, then B; denoted A `r emo::ji("right_arrow")` B)

The assertion that B is necessary for A is equivalent to:
- "A cannot be true unless B is true" or
- "if B is false, then A is false"
- (By contraposition: "whenever A is true, so is B")

--

**Example:**

Being at least 30 years old is necessary for serving in the U.S. Senate.
- If you are under 30, it is impossible to be a senator.
- If you are a senator, it follows that you are at least 30 years old.

---

## Necessary Versus Sufficient (Logic)

**Sufficient**

If X is sufficient for Y, then knowing X to be true means you can conclude that Y is true.
- However, knowing X to be false does **not** mean Y is false.

--

**Example:**

If the U.S. Congress passes a bill, the president's signature is sufficient to make it law.
- But if the president did not sign the bill (e.g., through a presidential veto), that does not mean the bill has not become law.
- (It could still have become law through a congressional override.)

---

## Correlation `r emo::ji("right_arrow")` Causation

.pull-left[
Although correlation isn't **sufficient**, it is a **necessary** component for causation.
]

.pull-right[
**Hill's Criteria of Causation**

Austin Bradford Hill, "The Environment and Disease: Association or Causation?" *Proceedings of the Royal Society of Medicine*, 58 (1965): 295–300.

**Cook and Campbell's Three Criteria (1979)**

Cook, T. D. and Campbell, D. T. (1979). *Quasi-experimentation: Design & analysis issues for field settings.* Boston: Houghton Mifflin.
]

---

class: middle, center, inverse

# Hill's Criteria

---

## Hill's Criteria

.pull-left[
**Strength:**
A relationship is more likely to be causal if the correlation coefficient is large and statistically significant.
]

.pull-right[
**Consistency:**
A relationship is more likely to be causal if it can be replicated.
]

---

## Hill's Criteria

.pull-left[
**Specificity:**
A relationship is more likely to be causal if there is no other likely explanation.
]

.pull-right[
**Temporality:**
A relationship is more likely to be causal if the effect always occurs **after** the cause.
]

---

## Hill's Criteria

.pull-left[
**Gradient:**
A relationship is more likely to be causal if a greater exposure to the suspected cause leads to a greater effect.
]

.pull-right[
**Plausibility:**
A relationship is more likely to be causal if there is a plausible mechanism between the cause and the effect.
]

---

## Hill's Criteria

**Coherence:** A relationship is more likely to be causal if it is compatible with related facts and theories.

**Experiment:** A relationship is more likely to be causal if it can be verified experimentally.

**Analogy:** A relationship is more likely to be causal if there are proven relationships between similar causes and effects.

---

## Hill's Criteria — Full List

1. Strength
2. Consistency
3. Specificity
4. Temporality
5. Gradient
6. Plausibility
7. Coherence
8. Experiment
9. Analogy

---

## Hill's Criteria in Practice: Subluxation

.pull-left[
**Subluxation**

An orthopedic subluxation — a true vertebral misalignment, or a mechanical joint dysfunction that affects mobility in the spine — is

**not the same as** a "chiropractic subluxation" that is alleged to cause disease by interfering with nerve supply to organs.
]

.pull-right[
Mirtz, T. A., Morgan, L., Wyatt, L. H., & Greene, L. (2009). An epidemiological examination of the subluxation construct using Hill's criteria of causation. *Chiropractic & osteopathy*, 17(1), 13.

[chiromt.biomedcentral.com/articles/10.1186/1746-1340-17-13](https://chiromt.biomedcentral.com/articles/10.1186/1746-1340-17-13)
]

---

class: middle, center, inverse

# Cook and Campbell (1979)

---

## Cook and Campbell (1979)

**Covariation:**
Changes in the assumed cause (X) are related to changes in the assumed effect (Y). Changing X results in a predictable change in Y.

--

**Temporal Precedence:**
The assumed cause must occur **before** the assumed effect.

--

**No Plausible Alternative Explanations:**
The assumed cause must be the only reasonable explanation for changes in the measured assumed effect.

---

## No Plausible Alternative Explanations

The third criterion is the **hardest to satisfy**.

Plausible alternative explanations are **"threats to validity"**

--

**Goal: Minimizing Threats to Validity**

Multiple approaches (can be combined):
- Argument
- Measurement or Observation
- Analysis
- Preventive Action
- Design

---

## Goal: Minimizing Threats to Validity

The five categories should not be considered mutually exclusive.

- Measurements designed to minimize threats will obviously be related to the design structure and is likely to be a factor in the analysis.
- A good research plan should, where possible, **make use of multiple methods** for reducing threats.
- In general, reducing a particular threat by **design or preventive action** will probably be stronger than by using one of the other approaches.
- The choice of which strategy to use depends on the cost and the potential seriousness of the threat.

---

class: middle, center, inverse

# Minimizing Threats to Validity

---

## By Argument

The most straightforward way to rule out a potential threat to validity is to **argue that the threat in question is not a reasonable one**.

Such an argument may be made either:
- *a priori* — more convincing because you make the argument **before** testing your results
- *a posteriori* — less convincing because you make the argument **after** testing your results

--

**Example:**
An instrumentation threat is not likely because the same test is used for pre- and post-test measurements, and did not involve observers who might improve.

--

Ruling out a threat by argument alone is **weaker** than other approaches. The most plausible threats should not, except in unusual cases, be ruled out by argument only.

---

## By Measurement or Observation

In some cases, it will be possible to rule out a threat by **measuring it** and demonstrating that either:
- it does not occur at all, or
- it occurs so minimally as to not be a strong alternative explanation.

--

**Example:**

In a study of the effects of an advertising campaign on product sales, *history* (other events causing increased purchase desire) would be a plausible alternative explanation.

- Measure local economic indicators and availability of competing products.
- If there is no change in these measures coincident with the advertising campaign, these threats are considerably minimized.

---

## By Analysis

There are a number of ways to rule out alternative explanations using **statistical analysis**.

**Example 1 (Jurs and Glass, 1971):**
Study the plausibility of an attrition threat by conducting an ANOVA with two factors:
- Factor 1: original treatment group designations (program vs. comparison)
- Factor 2: attrition (dropout vs. non-dropout)
- A main effect on attrition = threat to external validity; interaction = threat to internal validity.

--

**Example 2 (ANCOVA):**
In a study of "workfare" programs on social welfare caseloads, use a measure of economic conditions as a covariate.
- Causal assertions are strengthened by demonstrating treatment effects occur even after adjusting on good covariates.
- Note: "perfect" covariates do not exist; imperfect covariates will not completely adjust.

---

## By Preventive Action

When potential threats are **anticipated**, they can often be ruled out by some type of **preventive action**.

Auditing methods and quality control can be used to:
- Track potential experimental dropouts
- Ensure the standardization of measurement

--

**Example:**

If the program is desirable, the comparison group may feel jealous or demoralized. Actions to minimize:
- Offer the program to the comparison group upon completion of the study
- Use program and comparison groups with little opportunity for contact and communication

---

## By Design

The major emphasis is on ruling out alternative explanations by **adding treatment or control groups, waves of measurement**, and the like.

This is the **largest area of emphasis**.

--

**Basic Design Elements**

Most research designs can be constructed from four basic elements:

1. **Time**
2. **Program(s) or Treatment(s)**
3. **Observation(s) or Measure(s)**
4. **Groups or Individuals**

---

## Design Construction

<!-- TODO: add design notation diagram showing basic elements (X for treatment, O for observation, R/N/C for group assignment) -->

---

## Time (Pt 1)

A causal relationship, by its very nature, implies that some **time has elapsed** between the occurrence of the cause and the consequent effect.

- For some phenomena the elapsed time might be measured in microseconds.
- We normally assume that the cause and effect in social science arenas do **not occur simultaneously**.

---

## Time (Pt 2)

In design notation, we indicate the temporal element **horizontally**:
- The symbol for the presumed cause is placed to the **left** of the symbol for the measurement of the effect.
- Reading left to right = reading across time.
- Complex designs might involve a lengthy sequence of observations and programs across time.

---

## Program(s) or Treatment(s)

The presumed cause may be:
- A program or treatment under the explicit **control of the researcher**, or
- The occurrence of some **natural event** not explicitly controlled.

--

In design notation:
- Presumed cause = **"X"**
- Multiple programs: use subscripts **"X₁"** or **"X₂"**
- Comparison group (no program) = **no "X"**

---

## Observation(s) or Measure(s)

Measurements are depicted in design notation with the symbol **"O"**.

- If the same measurement is taken at every point in time → a single "O" suffices.
- If different measures are given at different times → subscript the "O" to indicate which measurement is given at which time.

---

## Groups or Individuals

The final design element consists of the intact groups or individuals who participate in various conditions.

In design notation, each group is indicated on a **separate line**.

Group assignment symbols:
- **R** = randomly assigned
- **N** = nonrandomly assigned (nonequivalent group or cohort)
- **C** = assigned using a cutoff score on a measurement

---

## Illustrative Example: Two-Group Pre-Post Design

One of the most used designs in social research is the **two-group pre-post design**:

```
N  O  X  O
N  O     O
```

- **Two lines** = two groups
- **N** = nonrandomly assigned
- First **O** in each line = pre-program measurement
- **X** = program/treatment (first group only)
- Final **O** = post-program measurement for both groups

---

## Posttest-Only Randomized Experiment

Another common design is the **posttest-only randomized experiment**:

```
R  X  O
R     O
```

- Two groups are **randomly selected** (R)
- One group receives the program; one acts as comparison
- Both groups are measured **after** the program is administered

---

## Expanding a Design

We can combine the four basic design elements in several ways to arrive at a specific design appropriate for the setting at hand.

One strategy begins with the basic causal relationship:

```
X  O
```

--

This is the **simplest design** in causal research and serves as a starting point for better strategies.

When we add to this design, we are expanding one of the four basic elements. Each expansion has implications for:
- Cost of the study
- Threats that can be ruled out

---

## Expanding Across Time

We can add additional observations **before or after** the program.

Adding pretests provides a **"baseline"**:

```
O  O  X  O
```

--

This helps assess the potential of a **maturation or testing threat**:
- If change occurs between the first and second pre-program measures → expect similar change between second pretest and posttest even without the program.
- If **no change** occurs between the two pretests → more confidence that maturation or testing is not a likely alternative explanation.

---

## Expanding Across Time

Additional **post-program measures** can be added:

```
X  O  O  O
```

Useful for determining whether:
- An immediate program effect **decays over time**
- There is a **lag** between program initiation and effect occurrence

--

We might also add and remove the program over time — the **ABAB design**:

```
X  O  O  X  O  O
```

- Frequently used in clinical psychology and psychiatry
- Particularly strong against a **history threat**: when the program is repeated, it is less likely that unique historical events are responsible for replicated outcome patterns.

---

## Expanding Across Programs

We can **partition the program** into different levels of treatment.

Example — study the effect of a drug at multiple dosages:

```
X₁  O
X₂  O
```

--

This is an example of a **simple factorial design** with one factor having two levels.

This is a common strategy in a "sensitivity" or "parametric" study where the focus is on the effects obtained at various program levels.

We can also expand by varying **specific components** across groups, or by using **theoretically polarized ("opposite") treatments** to demonstrate that the outcome measure is sensitive enough to differentiate between different programs.

---

## Expanding Across Observations

At any point in time, it is desirable to collect **multiple measurements**:

--

- Add similar measures to determine whether results **converge** (convergent validity)
- Add measures theoretically unaffected by the program to demonstrate **discriminant validity**

--

**Proxy measurements:**
If pre-program measurements were neglected, use existing records (e.g., GPA records) as a proxy measure of pre-program status.

--

**Recollected measures:**
Ask participants to **recall** how they felt or behaved prior to the study.

The key principle: use **multiple measurements** rather than relying on only a single strategy.

---

## Expanding Across Groups

Adding additional groups can rule out specific threats to validity.

Consider a pre-post two-group randomized design in a **single institution** where groups are in contact — this risks threats from:
- Intergroup communication
- Group rivalry
- Demoralization of a group denied a desirable treatment

--

Solution: Add a **nonequivalent group from a similar institution** whose members are unaware of the original groups:

```
R  O  X  O
R  O     O
N  O     O
```

---

## Expanding Across Groups

When nonequivalent groups are used, it is advantageous to have **multiple replications** of each group.

Using many nonequivalent groups minimizes the potential of a particular **selection bias** affecting results.

--

We can also include the **norm group** as an additional nonequivalent control group.
- Norming group averages are available for most standardized achievement tests.

**Cohort groups** might also be used to rule out a **testing threat**:

```
N  O  X  O
N     X  O  (cohort — no pretest exposure)
```

---

## Next Time...

**Lab**

---

class: middle, center, inverse

# Wrapping Up...
