---
title: "Replication Crisis & Open Science"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Replication Crisis & Open Science

---

# Roadmap

- What is the replication crisis?
- Why does it happen?
  - p-hacking
  - HARKing
  - Publication bias
- What are the consequences?
- What can we do about it?
  - Open Science practices

--

**By the end of this lecture, you should be able to:**

- Define the replication crisis and explain its scope across scientific disciplines
- Identify questionable research practices (p-hacking, HARKing) and explain how they inflate false-positive rates
- Explain how publication bias contributes to a distorted scientific literature
- Describe open science solutions such as pre-registration, open data, and registered reports
- Distinguish between Type I and Type II errors and explain the role of statistical power in replication

---

class: middle, center, inverse

# What Is the Replication Crisis?

---

# The Replication Crisis

- A **replication crisis** (or reproducibility crisis) is an ongoing methodological crisis
- Findings from many scientific studies are **difficult or impossible to reproduce**
- Affects psychology, medicine, economics, and other fields
- Has prompted major reforms in how science is conducted and reported

---

# The Open Science Collaboration (2015)

- Replicated **100 published psychology studies**
- Only **~36–39%** of replications showed a significant effect in the same direction
- Original studies had a 97% success rate; replications: ~36%
- Published in *Science* — sparked enormous debate

<!-- TODO: add [bar chart comparing original vs. replication success rates from OSC 2015] image -->

---

# Why Does This Matter?

- Science is supposed to be **self-correcting**
- If results don't replicate, the knowledge base is unreliable
- Has real consequences:
  - Clinical treatments based on flawed findings
  - Policy decisions built on shaky evidence
  - Wasted research resources

---

class: middle, center, inverse

# Why Does the Crisis Happen?

---

# Questionable Research Practices (QRPs)

.pull-left[
**Common QRPs:**
- p-hacking (data dredging)
- HARKing (Hypothesizing After Results are Known)
- Selective reporting of outcomes
- Stopping data collection when *p* < .05
- Dropping conditions or participants post-hoc
- Running multiple analyses and reporting only significant ones
]
.pull-right[
<!-- TODO: add [illustration of researcher at branching garden-of-forking-paths decision tree] image -->
]

---

# p-Hacking

- **p-hacking**: manipulating analyses until *p* < .05
- Researchers make many "researcher degrees of freedom" decisions:
  - Which outliers to exclude
  - Which covariates to include
  - Which dependent variable to report
  - When to stop collecting data
- Each decision inflates the false-positive rate

---

# The Garden of Forking Paths

- Gelman & Loken (2013): the **garden of forking paths**
- Even without deliberate cheating, researchers face many analytic choices
- Each path that could have been taken (but wasn't) inflates Type I error
- The final reported analysis is just one of many possible analyses

<!-- TODO: add [garden of forking paths diagram] image -->

---

# Simmons, Nelson & Simonsohn (2011)

- Demonstrated that flexible data collection and analysis can produce *p* < .05 for **impossible effects**
  - e.g., listening to "When I'm Sixty-Four" made participants *younger*
- With just a few researcher degrees of freedom, false-positive rate jumps from 5% to **>60%**

---

# HARKing

- **HARKing** = Hypothesizing After Results are Known
- Presenting post-hoc findings as if they were predicted a priori
- Makes exploratory findings look like confirmatory ones
- Inflates the apparent evidential value of results

--

- Example: you measure 20 variables, find that one is significant, and write your paper as if that was the hypothesis all along

---

# Publication Bias

.pull-left[
- **Publication bias**: significant results are more likely to be published than null results
- Creates a **file drawer problem**: null results sit unpublished
- The published literature is a biased sample of all studies conducted
]
.pull-right[
<!-- TODO: add [funnel plot illustrating publication bias / missing studies in bottom-left] image -->
]

---

# The File Drawer Problem

- Rosenthal (1979) coined the term **file drawer problem**
- For every published significant result, how many null results are sitting in file drawers?
- If 20 labs run the same study and only the 1 that gets *p* < .05 publishes, that result is meaningless

---

# Funnel Plots & Publication Bias

- **Funnel plots** plot effect size against sample size (or standard error)
- In the absence of bias: symmetric funnel shape
- Publication bias produces **asymmetric funnels** — small studies with null results are missing

<!-- TODO: add [symmetric vs. asymmetric funnel plot comparison] image -->

---

# Statistical Power

- **Power** = probability of detecting a true effect if it exists
  - Power = 1 − β (where β = Type II error rate)
- Low-powered studies:
  - Miss real effects (high Type II error)
  - When they do get significant results, **effect size estimates are inflated** (winner's curse)

---

# Type I and Type II Errors

|  | H₀ True | H₀ False |
|--|---------|---------|
| **Reject H₀** | Type I Error (α) | Correct (Power) |
| **Fail to Reject H₀** | Correct | Type II Error (β) |

- **Type I error** (false positive): finding an effect that isn't there
- **Type II error** (false negative): missing a real effect
- Conventional thresholds: α = .05, power = .80

---

# The Winner's Curse

- When power is low, only the most extreme (largest) estimates cross the significance threshold
- Published effects from underpowered studies are **inflated** relative to the true effect
- Replications with adequate power often find smaller effects — looks like a failure to replicate

<!-- TODO: add [winner's curse diagram showing distribution of estimates that achieve significance] image -->

---

class: middle, center, inverse

# Scope of the Crisis

---

# Beyond Psychology

The replication crisis spans many fields:

- **Medicine**: Ioannidis (2005) — "Why Most Published Research Findings Are False"
- **Cancer biology**: Reproducibility Project: Cancer Biology replicated only a fraction of landmark findings
- **Economics**: replication rates ~50–60%
- **Neuroscience**: small samples, many false positives
- **Social priming**: many classic effects failed to replicate

---

# Many Labs Projects

- **Many Labs 1** (Klein et al., 2014): 36 labs, 13 effects — some replicated robustly, others did not
- **Many Labs 2**: 28 effects across 125 samples — high variability in replicability
- **Many Labs 3**: tested whether time of semester moderates effect sizes — mostly no
- These projects show which findings are **robust** vs. **fragile**

<!-- TODO: add [Many Labs replication success rate bar chart] image -->

---

# Estimating the Scale

- Survey of 1,500 scientists (Baker, 2016, *Nature*):
  - >70% have tried and failed to reproduce another scientist's experiments
  - >50% have failed to reproduce their own experiments
  - 52% said there is a "significant crisis"

---

class: middle, center, inverse

# What Can We Do About It?

---

# Open Science Solutions

.pull-left[
**Transparency practices:**
- Pre-registration
- Registered Reports
- Open data & materials
- Open access publishing
- Replication studies
]
.pull-right[
<!-- TODO: add [Open Science Framework (OSF) logo or open science badge graphic] image -->
]

---

# Pre-Registration

- **Pre-registration**: publicly committing to hypotheses, design, and analysis plan *before* data collection
- Separates **confirmatory** (hypothesis-testing) from **exploratory** (hypothesis-generating) research
- Prevents p-hacking and HARKing
- Platforms: OSF, AsPredicted, ClinicalTrials.gov

---

# What Pre-Registration Does (and Doesn't) Do

.pull-left[
**Does:**
- Lock in primary hypotheses and analyses
- Reduce researcher degrees of freedom for confirmatory tests
- Increase credibility of significant findings
]
.pull-right[
**Doesn't:**
- Prevent exploratory analyses (just label them)
- Guarantee a good study design
- Eliminate all QRPs
]

---

# Registered Reports

- A publishing format where **peer review happens before data collection**
- Stage 1: introduction, methods, and analysis plan reviewed; journal gives **in-principle acceptance**
- Stage 2: completed study reviewed for adherence to plan
- Results are published **regardless of outcome**
- Directly eliminates publication bias

<!-- TODO: add [registered reports two-stage pipeline diagram] image -->

---

# Open Data & Open Materials

- **Open data**: sharing raw data so others can verify and reuse it
- **Open materials**: sharing stimuli, code, and procedures
- Enables **independent replication** and error-checking
- Platforms: OSF, GitHub, Zenodo, Dataverse

--

- Many journals now offer **open science badges** for data sharing, materials sharing, and pre-registration

---

# Effect Sizes

- Move beyond binary significant/non-significant
- **Effect sizes** quantify the *magnitude* of an effect
  - Cohen's *d*, Pearson's *r*, η², odds ratio
- Allow comparison across studies and meta-analyses
- Small *p*-value ≠ large effect

---

# Cohen's Conventions

| Label | Cohen's *d* | *r* |
|-------|------------|-----|
| Small | 0.2 | .10 |
| Medium | 0.5 | .30 |
| Large | 0.8 | .50 |

- These are **arbitrary conventions** — interpret in context
- A "small" effect can be practically important (e.g., a cheap public health intervention)

---

# Confidence Intervals

- Report **confidence intervals** alongside point estimates
- A 95% CI that barely excludes 0 ≠ strong evidence
- Wide CIs indicate imprecise estimates (usually low N)
- Overlapping CIs do not necessarily mean non-significant difference

---

# Meta-Analysis

- **Meta-analysis**: statistically combining results across multiple studies
- Provides more precise estimates of true effect sizes
- Can detect publication bias (funnel plot asymmetry, Egger's test, p-curve)
- Systematic reviews + meta-analyses sit at the top of the **evidence hierarchy**

<!-- TODO: add [evidence pyramid / hierarchy of evidence diagram] image -->

---

# p-Curve Analysis

- **p-curve** (Simonsohn et al., 2014): distribution of significant p-values
- If a true effect exists: p-curve is **right-skewed** (many p-values near .05 are unlikely)
- If p-hacking occurred: p-curve is **left-skewed** (suspiciously many p-values just below .05)
- Can diagnose whether a literature has evidential value

<!-- TODO: add [p-curve right-skewed vs. left-skewed comparison plot] image -->

---

# Increasing Statistical Power

Ways to increase power:
- **Larger samples** (most direct)
- Larger effect sizes (use stronger manipulations)
- Within-subjects designs (reduce error variance)
- Reliable measures (reduce measurement error)
- One-tailed tests (only if justified a priori)

--

- **A priori power analysis**: determine required N before data collection

---

# Replication as Science

- **Direct replication**: same procedure, same measures — tests reliability
- **Conceptual replication**: different procedure, same construct — tests generalizability
- Both are valuable; neither is "just checking someone's homework"
- Replication failures are **informative**, not attacks

---

# Cultural & Incentive Changes

The crisis is also a **culture problem**:

- Publish-or-perish rewards novelty, not rigor
- Journals reward significant results, not careful science
- Replications and null results are undervalued

--

**Proposed changes:**
- Reward rigor in hiring/promotion
- Fund replication studies
- Train researchers in open science from the start

---

# Many Labs & Crowdsourced Science

- Large-scale collaborations increase power and generalizability
- **Psychological Science Accelerator**: global network running high-powered studies
- Adversarial collaborations: proponents and skeptics of an effect design a study together
- **Multi-site studies** increase external validity

---

# Summary: The Problem

- Many published findings don't replicate
- Root causes: p-hacking, HARKing, publication bias, low power
- These aren't just individual failings — they're **systemic incentive problems**

---

# Summary: The Solutions

| Problem | Solution |
|---------|----------|
| p-hacking | Pre-registration |
| HARKing | Pre-registration; distinguish confirmatory vs. exploratory |
| Publication bias | Registered Reports; open results |
| Low power | Larger samples; power analysis |
| Non-transparency | Open data; open materials |
| Effect inflation | Report effect sizes & CIs |

---

# The Glass Half Full

- The replication crisis is a **sign of a self-correcting science**
- We are identifying problems and developing solutions
- Open science practices are becoming standard
- Early-career researchers have an opportunity to **lead this change**

--

> "The goal of science is not to produce findings; it is to produce knowledge."

---

class: middle, center, inverse

# Wrapping Up...

---

# Recap

- The **replication crisis** revealed that many published findings are unreliable
- Key causes: **p-hacking**, **HARKing**, **publication bias**, and **low statistical power**
- **Open science** practices — pre-registration, registered reports, open data — address these problems directly
- **Effect sizes**, **confidence intervals**, and **meta-analysis** provide richer evidence than *p*-values alone
- The field is actively reforming; you can be part of the solution

---

# References

- Open Science Collaboration (2015). Estimating the reproducibility of psychological science. *Science*, 349(6251).
- Simmons, J. P., Nelson, L. D., & Simonsohn, U. (2011). False-positive psychology. *Psychological Science*, 22(11), 1359–1366.
- Ioannidis, J. P. A. (2005). Why most published research findings are false. *PLOS Medicine*, 2(8), e124.
- Nosek, B. A., et al. (2015). Promoting an open research culture. *Science*, 348(6242), 1422–1425.
- Gelman, A., & Loken, E. (2013). The garden of forking paths. Unpublished manuscript.
- Rosenthal, R. (1979). The file drawer problem and tolerance for null results. *Psychological Bulletin*, 86(3), 638–641.
