---
title: "Probability"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

---

class: middle, center, inverse

# Probability

---

# Roadmap

.pull-left[
**Where we've been:**
- Exploratory data analysis
- Data visualization
- Summarizing distributions
]

.pull-right[
**Where we're going:**
- Defining probability
- Probability rules
- Conditional probability
- Random variables
]

--

**By the end of this lecture, you should be able to:**

- Define probability and describe the sample space and events
- Apply the addition, complement, and multiplication rules correctly
- Distinguish between independent and dependent events
- Calculate conditional probabilities and apply Bayes' theorem
- Define a random variable and compute its expected value and variance

---

class: middle, center, inverse

# What Is Probability?

---

# Defining Probability

**Probability** is a number between 0 and 1 that describes how likely an event is to occur.

$$0 \leq P(A) \leq 1$$

--

- $P(A) = 0$: the event **never** occurs
- $P(A) = 1$: the event **always** occurs
- $P(A) = 0.5$: the event occurs **half the time**

--

**Frequentist interpretation:**

$$P(A) = \lim_{n \to \infty} \frac{\text{number of times } A \text{ occurs}}{n}$$

---

# The Law of Large Numbers

> As the number of trials increases, the proportion of times an event occurs approaches the true probability.

<!-- TODO: add [line plot showing running proportion of heads in coin flips converging to 0.5 as number of flips increases] image -->

--

- With only a few trials, results can vary widely
- With many trials, the proportion stabilizes near $P(A)$
- This is why casinos always win in the long run `r emo::ji("slot_machine")`

---

class: middle, center, inverse

# Sample Spaces and Events

---

# Sample Space

The **sample space** $S$ is the set of all possible outcomes of a random process.

--

**Examples:**

| Random process | Sample space |
|---|---|
| Flip a coin | $S = \{H, T\}$ |
| Roll a die | $S = \{1, 2, 3, 4, 5, 6\}$ |
| Draw a card | $S = \{A\spadesuit, 2\spadesuit, \ldots, K\heartsuit\}$ |
| Record a person's height | $S = (0, \infty)$ |

---

# Events

An **event** is any collection of outcomes from the sample space — a subset of $S$.

--

**Example:** Roll a standard die.
- $S = \{1, 2, 3, 4, 5, 6\}$
- Event $A$ = roll an even number = $\{2, 4, 6\}$
- Event $B$ = roll a number greater than 4 = $\{5, 6\}$

--

**Venn diagrams** are useful for visualizing relationships between events.

<!-- TODO: add [Venn diagram showing two overlapping circles A and B within rectangle S] image -->

---

class: middle, center, inverse

# Probability Rules

---

# Rule 1: Complement Rule

The **complement** of event $A$, written $A^c$ (or $\bar{A}$), is the event that $A$ does **not** occur.

$$P(A^c) = 1 - P(A)$$

--

**Example:** If the probability of rain tomorrow is 0.30, then:

$$P(\text{no rain}) = 1 - 0.30 = 0.70$$

---

# Rule 2: Addition Rule

**For any two events $A$ and $B$:**

$$P(A \text{ or } B) = P(A) + P(B) - P(A \text{ and } B)$$

<!-- TODO: add [Venn diagram illustrating addition rule with shaded union of A and B] image -->

--

**Special case — Mutually Exclusive Events:**

If $A$ and $B$ cannot both occur ($P(A \text{ and } B) = 0$):

$$P(A \text{ or } B) = P(A) + P(B)$$

---

# Addition Rule — Example

Roll a fair die. What is $P(\text{even or} > 4)$?

- $A$ = even = $\{2, 4, 6\}$, so $P(A) = 3/6$
- $B$ = greater than 4 = $\{5, 6\}$, so $P(B) = 2/6$
- $A \text{ and } B$ = $\{6\}$, so $P(A \text{ and } B) = 1/6$

--

$$P(A \text{ or } B) = \frac{3}{6} + \frac{2}{6} - \frac{1}{6} = \frac{4}{6} \approx 0.667$$

---

class: middle, center, inverse

# Independence and the Multiplication Rule

---

# Independent Events

Two events $A$ and $B$ are **independent** if knowing that one occurred does not change the probability of the other.

$$P(A | B) = P(A) \quad \text{and} \quad P(B | A) = P(B)$$

--

**Examples of independent events:**
- Two separate coin flips
- Drawing two cards **with replacement**
- Two different people's blood types (unrelated)

--

**Examples of dependent events:**
- Drawing two cards **without replacement**
- Weather on consecutive days

---

# Rule 3: Multiplication Rule

**For independent events $A$ and $B$:**

$$P(A \text{ and } B) = P(A) \times P(B)$$

--

**Example:** Flip a fair coin twice.

$$P(HH) = P(H) \times P(H) = \frac{1}{2} \times \frac{1}{2} = \frac{1}{4}$$

--

**General multiplication rule (any events):**

$$P(A \text{ and } B) = P(A) \times P(B | A)$$

---

# Multiplication Rule — Example

A bag contains 5 red and 3 blue marbles. Draw two marbles **without replacement**.

$P(\text{both red}) = ?$

--

- $P(\text{1st red}) = 5/8$
- $P(\text{2nd red} | \text{1st red}) = 4/7$

$$P(\text{both red}) = \frac{5}{8} \times \frac{4}{7} = \frac{20}{56} \approx 0.357$$

---

class: middle, center, inverse

# Conditional Probability

---

# Conditional Probability

The **conditional probability** of $A$ given $B$ is the probability that $A$ occurs, given that $B$ has already occurred.

$$P(A | B) = \frac{P(A \text{ and } B)}{P(B)}, \quad P(B) > 0$$

--

- We "restrict" the sample space to only outcomes where $B$ occurred
- Then ask: in that restricted space, how likely is $A$?

<!-- TODO: add [Venn diagram showing P(A|B) as the intersection A∩B relative to the area of B] image -->

---

# Conditional Probability — Example

A standard deck of 52 cards. Draw one card.

$P(\text{King} | \text{Face card}) = ?$

--

- There are 12 face cards (J, Q, K in each of 4 suits)
- There are 4 Kings

$$P(\text{King} | \text{Face card}) = \frac{P(\text{King and Face card})}{P(\text{Face card})} = \frac{4/52}{12/52} = \frac{4}{12} = \frac{1}{3}$$

---

# Conditional Probability — Table Method

<!-- TODO: add [two-way table showing counts of events to illustrate conditional probability calculation] image -->

Two-way (contingency) tables are a handy tool for computing conditional probabilities:

1. Find the row or column corresponding to the conditioning event
2. Divide the count of the joint event by the row/column total

---

class: middle, center, inverse

# Bayes' Theorem

---

# Bayes' Theorem

Sometimes we know $P(B | A)$ but want $P(A | B)$.

$$P(A | B) = \frac{P(A) \times P(B | A)}{P(B)}$$

--

Where the **law of total probability** gives:

$$P(B) = P(B | A) \times P(A) + P(B | A^c) \times P(A^c)$$

---

# Bayes' Theorem — Example

A medical test for a disease:
- Prevalence: $P(\text{disease}) = 0.01$
- Sensitivity: $P(+ | \text{disease}) = 0.99$
- False positive rate: $P(+ | \text{no disease}) = 0.05$

--

$P(\text{disease} | +) = ?$

--

$$P(B) = (0.99)(0.01) + (0.05)(0.99) = 0.0099 + 0.0495 = 0.0594$$

$$P(\text{disease} | +) = \frac{0.99 \times 0.01}{0.0594} \approx 0.167$$

--

Even with a positive test, only ~17% chance of actually having the disease! `r emo::ji("thinking")`

---

# Bayes' Theorem — Tree Diagram

<!-- TODO: add [probability tree diagram for the medical test Bayes' example, showing disease/no disease branches and positive/negative test branches with probabilities labeled] image -->

Tree diagrams help organize the joint probabilities needed for Bayes' theorem.

---

class: middle, center, inverse

# Random Variables

---

# Random Variables

A **random variable** is a variable whose value is determined by the outcome of a random process.

--

.pull-left[
**Discrete random variable**
- Takes on a countable number of values
- Examples: number of heads in 10 flips, number of students absent
]

.pull-right[
**Continuous random variable**
- Takes on any value in an interval
- Examples: height, weight, time
]

---

# Probability Distribution

The **probability distribution** of a discrete random variable $X$ lists all possible values and their probabilities.

| $x$ | $P(X = x)$ |
|---|---|
| 0 | 0.25 |
| 1 | 0.50 |
| 2 | 0.25 |

--

**Requirements:**
- All probabilities $\geq 0$
- Probabilities sum to 1: $\sum P(X = x) = 1$

---

# Expected Value

The **expected value** (mean) of a random variable is the long-run average value.

$$E(X) = \mu_X = \sum x \cdot P(X = x)$$

--

**Example:** Number of heads in 2 fair coin flips.

$$E(X) = 0(0.25) + 1(0.50) + 2(0.25) = 0 + 0.50 + 0.50 = 1$$

--

- Interpretation: if we repeat the experiment many times, the average number of heads will be close to 1

---

# Variance and Standard Deviation

The **variance** of a random variable measures spread around the mean.

$$\text{Var}(X) = \sigma^2_X = \sum (x - \mu_X)^2 \cdot P(X = x)$$

$$SD(X) = \sigma_X = \sqrt{\text{Var}(X)}$$

--

**Example continued:** ($\mu_X = 1$)

$$\sigma^2_X = (0-1)^2(0.25) + (1-1)^2(0.50) + (2-1)^2(0.25) = 0.25 + 0 + 0.25 = 0.50$$

$$\sigma_X = \sqrt{0.50} \approx 0.707$$

---

class: middle, center, inverse

# Rules for Random Variables

---

# Linear Transformations

If $Y = a + bX$:

$$E(Y) = a + b \cdot E(X)$$
$$\text{Var}(Y) = b^2 \cdot \text{Var}(X)$$
$$SD(Y) = |b| \cdot SD(X)$$

--

**Example:** Convert temperature $X$ in Celsius to Fahrenheit: $Y = 32 + 1.8X$

If $E(X) = 20°C$ and $SD(X) = 5°C$:
- $E(Y) = 32 + 1.8(20) = 68°F$
- $SD(Y) = 1.8 \times 5 = 9°F$

---

# Combining Random Variables

For **any** two random variables $X$ and $Y$:

$$E(X + Y) = E(X) + E(Y)$$
$$E(X - Y) = E(X) - E(Y)$$

--

For **independent** random variables $X$ and $Y$:

$$\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y)$$
$$\text{Var}(X - Y) = \text{Var}(X) + \text{Var}(Y)$$

--

Note: variances **add** for both sums and differences (when independent).

---

# Combining Random Variables — Example

A coffee shop sells lattes and cappuccinos.
- Lattes: $E(L) = \$4.50$, $SD(L) = \$0.50$
- Cappuccinos: $E(C) = \$4.00$, $SD(C) = \$0.40$
- $L$ and $C$ are independent

--

Expected total revenue from one of each:

$$E(L + C) = 4.50 + 4.00 = \$8.50$$

--

SD of total:

$$SD(L + C) = \sqrt{0.50^2 + 0.40^2} = \sqrt{0.25 + 0.16} = \sqrt{0.41} \approx \$0.64$$

---

class: middle, center, inverse

# Continuous Random Variables

---

# Continuous Distributions

For a **continuous** random variable:
- Probabilities are represented by **areas under a curve**
- The curve is called a **probability density function (PDF)**
- $P(X = \text{exact value}) = 0$

<!-- TODO: add [smooth bell-curve or other density curve with a shaded region representing P(a < X < b)] image -->

--

$$P(a < X < b) = \int_a^b f(x)\, dx$$

---

# The Normal Distribution

The **normal distribution** is the most important continuous distribution in statistics.

$$X \sim N(\mu, \sigma)$$

<!-- TODO: add [normal distribution bell curve labeled with mean mu at center and sigma showing spread, possibly with 68-95-99.7 rule marked] image -->

--

- Symmetric, bell-shaped
- Completely described by $\mu$ and $\sigma$
- **68-95-99.7 rule:**
  - 68% of data within $\pm 1\sigma$
  - 95% within $\pm 2\sigma$
  - 99.7% within $\pm 3\sigma$

---

# Standardizing — Z-scores

Any normal variable can be converted to the **standard normal** $Z \sim N(0, 1)$:

$$Z = \frac{X - \mu}{\sigma}$$

--

- $Z$ tells us how many standard deviations $X$ is from the mean
- We can use a Z-table (or R) to find probabilities

--

**Example:** Heights are $N(65, 3.5)$ inches. What proportion are taller than 70 inches?

$$Z = \frac{70 - 65}{3.5} \approx 1.43$$

$$P(X > 70) = P(Z > 1.43) \approx 0.076$$

---

class: middle, center, inverse

# Wrapping Up...

---

# Summary

- **Probability** measures likelihood; ranges from 0 to 1
- The **sample space** contains all possible outcomes; **events** are subsets
- Key rules:
  - **Complement:** $P(A^c) = 1 - P(A)$
  - **Addition:** $P(A \text{ or } B) = P(A) + P(B) - P(A \text{ and } B)$
  - **Multiplication:** $P(A \text{ and } B) = P(A) \times P(B|A)$
- **Independent** events: knowing one gives no information about the other
- **Conditional probability:** $P(A|B) = P(A \text{ and } B) / P(B)$
- **Bayes' theorem** lets us "flip" the conditioning
- **Random variables** have expected values and variances that follow algebraic rules

---

# References

```{r, results='asis', echo=FALSE}
PrintBibliography <- function(...) invisible(NULL)
```
