---
title: "Validity"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---


```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle

# Validity

---

## Roadmap

.pull-left[
**Ways to Organize Validity**

**Causal Inference Perspective**

- Cooke & Campbell
  - External Validity
  - Internal Validity
- Hill's Criteria
]

.pull-right[
**Education Measurement Perspective**

- Content-related Validity
- Criterion-related Validity
- Construct-related Validity
]

--

**By the end of this lecture, you should be able to:**

- Define validity and distinguish it from reliability
- Explain the two major frameworks for organizing validity (causal inference vs. education measurement perspectives)
- Describe internal and external validity and apply rules of thumb for generalizability
- Distinguish content-related, criterion-related, and construct-related evidence of validity
- Identify threats to validity including construct underrepresentation and construct-irrelevant variance
- Explain the relationship between validity and reliability, including the concept of maximum validity

---

## Validity Definitions

.pull-left[
**The extent to which a test measures what it was designed to measure.**

**Agreement between a test score or measure and the quality it is believed to measure.**
]

.pull-right[
Proliferation of definitions led to a dilution of the meaning of the word into all kinds of "validities"
]

---

## Validity That Isn't

**Face Validity**

- The mere appearance that a test has validity.
- Does the test look like it measures what it is supposed to measure?
- Do the items seem to be reasonably related to the perceived purpose of the test.

> Does a depression inventory ask questions about being sad?

- Not a "real" measure of validity, but one that is commonly seen in the literature.
- Not considered legitimate form of validity by the Joint Committee.

---

## Ways to Organize Validity

- **Causal Inference Perspective**
  - Shadish, Cook, and Campbell
  - Different types of validity
- **Education Measurement Perspective**
  - Standards for Educational and Psychological Testing
  - Different types of evidence

---

class: middle, center, inverse

# Causal Inference Perspective

---

class: middle

# Causal Inference Perspective

---

## Internal Validity

- Differs by kind of study
  - For experiments....
  - For quasi-experiments...
  - For observational studies...

- General idea is that your finding is "real"

---

## External Validity

.pull-left[
**To what extent one may safely generalize?**

- the (internally valid) finding from the sample studied to the
  - defined target population and
  - to other populations
    - (i.e. across time and space).

Generalize to other people
]

.pull-right[
<!-- TODO: add image illustrating generalization to other people -->
]

---

## Types of External Validity

**Population validity**

- can the sample results be generalized to the target population

**Ecological validity**

- whether the results can be applied to real life situations.
- Generalize to other (real) situations

---

## Generalizability

**The burden of proof**

- Avoid simplistic generalizations to members of other cultures and people in different times (including differences)
- Those who question the generalizability of a study
  - should propose when, how, and why it is not generalizable

--

Personally, my default assumption is that the study doesn't generalize

- And that it is up to the authors of the study to convince the reader (*me*)
- But, that is not the norm...

---

## Thinking About Generalizability

Slightly less than arbitrary rules of thumb:

- **My quick rule of thumb:**
  - if you could have participated in the study,
    - then the results mostly likely generalize to you
- **My thoughtful rule of thumb:**
  - If the reasons you couldn't have participated is unrelated to findings of the study,
    - then the results probably generalize to you.

*Assumption: The study is internally valid.*

This approach can be applied to groups other than yourself...

---

## Example: British Doctors' Study

.pull-left[
- Does smoking cause cancer?
  - Wasn't obvious or accepted in the medical community before this study
  - Doctors were often associated with tobacco ads
    - (like this one from 1946)
- Doll and Hill (1954, 1956) tracked 40,701 Male British Doctors
  - (2/3 of all registered UK physicians in Oct 1951)

<!-- TODO: add image of 1946 tobacco advertisement featuring doctors -->
]

.pull-right[
**It does.**
]

---

## British Doctors' Study (1951-2004)

- Doll and Hill (1954, 1956) tracked 40,701 Male British Doctors
  - (2/3 of all registered UK physicians in Oct 1951)
  - They examined whether there was a relationship between smoking and cancer/respiratory disease.
  - They compared non-smokers, ex-smokers, and current smokers by age.
- Their studies changed the conversation about cancer and smoking

---

## British Doctors' Study (1951-2004)

<!-- TODO: add image/figure from the British Doctors' Study showing smoking and cancer data -->

---

## British Doctors' Study and Those Slightly Less Than Arbitrary Rules of Thumb

.pull-left[
- Could I have participated in the study?
  - **No**
  - Why Not?:
    - Not British,
    - Not a Man,
    - Not an MD,
    - Born after this study was published

- Are those reasons related to the study's finding?
  - **Probably not.**

- **Verdict: Don't Smoke**
]

.pull-right[
<!-- TODO: add image related to smoking/British Doctors' Study -->
]

---

## If You Want to Learn More...

If you want to learn more about smoking, statistics, causation, or related history...

- [The study that helped spur the U.S. stop-smoking movement](https://www.cancer.org/latest-news/the-study-that-helped-spur-the-us-stop-smoking-movement.html)

- [Cigarette ads featuring doctors](https://www.history.com/news/cigarette-ads-doctors-smoking-endorsement)

- [Tobacco Advertising Archive](http://tobacco.stanford.edu)

- [Tobacco Industry Archives](https://www.industrydocuments.ucsf.edu/tobacco/)

---

class: middle, center, inverse

# Wrapping Up... Causal Inference Perspective

---

class: middle, center, inverse

# Education Measurement Perspective

---

## Key Ideas

.pull-left[
<!-- TODO: add image illustrating accumulated evidence / measurement -->
]

.pull-right[
- The validity is based on accumulated evidence
- Validity of past inferences changes based on the current available evidence
- Evidentiary bases for interpretations can be derived through a variety of methods
]

---

## Education Measurement Perspective

There is a total mess of "validities" and their definitions, what to do?

.pull-left[
**1985 - Joint Committee of:**

- AERA: American Education Research Association
- APA: American Psychological Association
- NCME: National Council on Measurement in Education

developed **Standards for Educational and Psychological Testing** (revised in 1999).
]

---

## Standards for Educational and Psychological Testing

According to the Joint Committee:

> **Validity is the evidence for inferences made about a test score.**

Three types of evidence:

- Content-related
- Criterion-related
- Construct-related

*Different from the notion of "different types of validity"*

---

## Three Types of Evidence

**Content-related evidence**
- Content Validity

--

**Criterion-related evidence**
- Criterion Validity

--

**Construct-related evidence**
- Construct Validity

---

class: middle, center, inverse

# Wrapping Up... Education Measurement Perspective

---

class: middle, center, inverse

# Content Validity

---

## Big Idea: Content Validity

.pull-left[
**Content validity**

- When trying to measure a domain are all sub-domains represented?
- Complimentary to
  - domain sampling theory and
  - reliability
- However, increasing content validity lowers internal consistency reliability
]

.pull-right[
**Main steps to establishing evidence**

1. Define the content domain...
2. Outline...
3. Subject matter experts evaluate...
]

---

## Content-Related Evidence

Does the test adequately sample the content or behavior domain that it is designed to measure?

- If items are not a good sample, results of testing will be misleading.

--

- Usually developed during test development.
- Not generally empirically evaluated.
- Judgment of subject matter experts.

---

## Content-Related Evidence

To develop a test with high content-related evidence of validity, you need:

- good logic
- intuitive skills
- Perseverance

Must consider:

- wording
- reading level

---

## Other Content-Related Evidence Terms

**Threats to content validity**

--

**Construct underrepresentation:**

- failure to capture important components of a construct.
- Test is designed for chapters 1-10 but
  - only chapters 1-8 show up on the test.

--

**Construct-irrelevant variance:**

- occurs when scores are influenced by factors irrelevant to the construct.
- Test is well-intentioned, but problems secondary to the test
  - negatively influence the results
  - (e.g., reading level, vocabulary, unmeasured secondary domains)

---

class: middle, center, inverse

# Wrapping Up... Content Validity

---

class: middle, center, inverse

# Criterion-Related Validity

---

## Criterion-Related Validity

**The success of measures used for prediction or estimation.**

Two types:

--

.pull-left[
**Concurrent validity**

<!-- TODO: add image illustrating concurrent validity -->
]

.pull-right[
**Predictive validity**

- the degree to which a test predicts (correlates with)
  - an external criteria that is
  - measured some time in the future
]

---

## Criterion-Related Evidence

Tells us how well a test corresponds with a particular criterion

- **criterion:** behavioral or measurable outcome

Examples:
- SAT predicting GPA (GPA is criterion)
- BDI scores predicting suicidality (suicide is criterion)

*Used to "predict the future" or "predict the present."*

---

## Predictive Validity Evidence

.pull-left[
**Forecasting the future**

- How well does a test predict future outcomes?
- Example: SAT predicting 1st year GPA
- Most tests don't have great predictive validity
  - Decrease due to time & method variance
]

.pull-right[
<!-- TODO: add diagram illustrating predictive validity timeline -->
]

---

## Concurrent Validity Evidence

**Forecasting the present**

- How well does a test predict current similar outcomes?
- Job samples, alternative tests used to demonstrate concurrent validity evidence
- Generally higher than predictive validity estimates

---

## Validity Coefficient

**Quantifying Criterion-Related Evidence**

- Correlation between the test and the criterion

However,

- recall that r² indicates explained variance.

*Sound Problematic??*

---

## Evaluating Criterion-Related Evidence

.pull-left[
**Look**
- Look for changes in the cause of relationships

**Examine**
- Examine what the criterion really means.
- Optimally the criterion should be something the test is trying to measure

**Review**
- Review the subject population in the validity study.
]

---

## Evaluating Criterion-Related Evidence

- Ensure the sample size in the validity study was adequate.

--

- Never confuse the criterion with the predictor.

--

- Watch for restricted ranges!

---

## Evaluating Criterion-Related Evidence

- Review evidence for validity generalization.
  - Tests only given in laboratory settings, then expected to demonstrate validity in classrooms?
  - Ecological validity?

--

- Consider differential prediction.
  - External validity?

---

class: middle, center, inverse

# Wrapping Up... Criterion-Related Validity

---

class: middle, center, inverse

# Construct Validity

---

## Construct Validity

**Are you measuring what you are intending to measure, overall?**

Specific types...

- Intentional validity
- Representation validity or translation validity
- "Face validity"
- Observation validity
- Convergent validity
- Discriminant validity

---

## Construct Validity: Types

.pull-left[
**Intentional validity**

- are you measuring what you are intending
  - and not something else?
- Requires ...

**Representation validity**

- aka ...
- How well have the constructs been translated into measurable outcomes?
]

.pull-right[
**Observation validity**

- How good are the measures themselves?

**Convergent validity**

- refers to the degree to which a measure is correlated ...

**Discriminant validity**

- describes the degree to which the operationalization does...
]

---

## Construct-Related Evidence

**Construct:** something constructed by mental synthesis

- What is Intelligence? Love? Depression?

**Construct Validity Evidence**

- Assembling evidence about what a test means
  - (and what it doesn't)
- Is a sequential process

---

## Construct-Related Evidence

.pull-left[
**Convergent Evidence**

Obtained when a measure correlates ...

- Self-report, collateral-report measures...
]

.pull-right[
**Discriminant Evidence**

Obtained when a measure...
]

---

class: middle, center, inverse

# Wrapping Up... Construct Validity

---

class: middle, center, inverse

# Connecting Validity to Reliability

---

## Recall

- The standard error of measurement (SE$_m$) estimates how repeated measures of a person on the same instrument tend to be distributed around their "true" score.
  - The true score is always an unknown because ...

---

## Validity & Estimates

**Standard Error of Estimate:**

$$s_{est.} = s_{Y-\hat{Y}} = s_y \sqrt{(1 - r^2) \frac{N-1}{N-2}}$$

- $s_{est.}$: standard error of estimate
- $s_y$: standard deviation of the test
- $r$: validity of the test

Essentially, regression all over again.

---

## Maximum Validity

<!-- TODO: add figure/diagram illustrating maximum validity concept -->

The maximum possible validity (correlation between test and criterion) is constrained by the reliability of both the test and the criterion.

---

## Maximum Validity Table

| Reliability of Test | Reliability of Criterion | Maximum Validity (Correlation) |
|:-------------------:|:------------------------:|:------------------------------:|
| 1.0 | 1.0 | 1.00 |
| 0.8 | 1.0 | 0.89 |
| 0.6 | 1.0 | 0.77 |
| 0.4 | 1.0 | 0.63 |
| 0.2 | 1.0 | 0.45 |
| 0.0 | 1.0 | 0.00 |
| 1.0 | 0.5 | 0.71 |
| 0.8 | 0.5 | 0.63 |
| 0.6 | 0.5 | 0.55 |
| 0.4 | 0.5 | 0.45 |
| 0.2 | 0.5 | 0.32 |
| 0.0 | 0.5 | 0.00 |
| 1.0 | 0.2 | 0.45 |
| 0.8 | 0.2 | 0.40 |
| 0.6 | 0.2 | 0.35 |
| 0.4 | 0.2 | 0.28 |
| 0.2 | 0.2 | 0.20 |
| 0.0 | 0.2 | 0.00 |
| 1.0 | 0.0 | 0.00 |
| 0.8 | 0.0 | 0.00 |
| 0.6 | 0.0 | 0.00 |
| 0.4 | 0.0 | 0.00 |
| 0.2 | 0.0 | 0.00 |
| 0.0 | 0.0 | 0.00 |

---

class: middle, center, inverse

# Wrapping Up...
