---
title: "p-values"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# p-values

---

# Roadmap

- What is a p-value?
- Common misinterpretations
- How p-values are used (and abused)
- The replication crisis
- Alternatives and complements to p-values
- Best practices

--

**By the end of this lecture, you should be able to:**

- Correctly define and interpret a p-value
- Identify common misinterpretations of p-values
- Explain how p-hacking inflates false positive rates
- Describe alternatives and complements to null hypothesis significance testing
- Apply best practices for reporting statistical evidence

---

class: middle, center, inverse

# What Is a p-value?

---

# What Is a p-value?

The **p-value** is the probability of observing a test statistic as extreme as (or more extreme than) the one you obtained, **assuming the null hypothesis is true**.

$$p = P(\text{data this extreme} \mid H_0 \text{ is true})$$

--

- It is **not** the probability that the null hypothesis is true
- It is **not** the probability that your result is due to chance
- It is **not** the probability that your finding will replicate

---

# The Logic of Null Hypothesis Significance Testing

1. State a **null hypothesis** $H_0$ (usually: no effect, no difference)
2. Collect data and compute a test statistic
3. Ask: how likely is this test statistic if $H_0$ were true?
4. If that probability (the p-value) is below a threshold $\alpha$, **reject** $H_0$

--

- The threshold $\alpha$ is conventionally set at **0.05**
- This means: if $H_0$ is true, we will incorrectly reject it 5% of the time (Type I error rate)

---

# Visualizing the p-value

<!-- TODO: add p-value visualization image (null distribution with shaded tail area) -->

- The p-value is the **area in the tail(s)** of the null distribution
- A small p-value means the observed result is far from what we would expect under $H_0$
- A large p-value means the observed result is consistent with $H_0$

---

class: middle, center, inverse

# Common Misinterpretations

---

# The Most Common Misinterpretations

.pull-left[
**What a p-value IS NOT:**

- The probability $H_0$ is true
- The probability your result is a fluke
- The probability your finding will replicate
- A measure of effect size
- Evidence that an effect is practically important
]

.pull-right[
**What a p-value IS:**

- $P(\text{data} \mid H_0)$
- A statement about the data given the null
- One piece of evidence, not a verdict
- Sensitive to sample size
- A continuous measure, not a binary pass/fail
]

---

# "p < .05 means the result is true"

- This is wrong — and dangerous
--

- Even with $p < .05$, there is a non-trivial probability the result is a false positive
- The **positive predictive value** of a significant result depends on:
  - The prior probability that $H_0$ is false (base rate of true effects)
  - The statistical power of the study
  - The significance threshold

--

- With low power and low base rates, **most** significant findings can be false positives (Ioannidis, 2005)

---

# False Positive Rate Depends on Power and Prior

<!-- TODO: add false positive rate / PPV diagram (Ioannidis-style) -->

- Suppose only 10% of hypotheses tested are true
- Study has 50% power and $\alpha = .05$
- What fraction of significant results are true positives?

$$PPV = \frac{\text{power} \times \text{prior}}{\text{power} \times \text{prior} + \alpha \times (1 - \text{prior})}$$

$$PPV = \frac{0.50 \times 0.10}{0.50 \times 0.10 + 0.05 \times 0.90} \approx 0.53$$

--

- So roughly **half** of significant results in this scenario are false positives!

---

# "p > .05 means there is no effect"

- Failure to reject $H_0$ is **not** the same as evidence for $H_0$
--

- Absence of evidence is not evidence of absence
- A non-significant result may reflect:
  - A truly null effect, OR
  - Insufficient power to detect a real effect
--

- Small studies that "fail to find an effect" often just lack power

---

class: middle, center, inverse

# How p-values Are Used (and Abused)

---

# The p < .05 Threshold

- Ronald Fisher originally proposed $p < .05$ as a **soft** heuristic — not a hard rule
- Over time it became a bright-line threshold for publication and credibility
--

- This created perverse incentives:
  - Studies with $p < .05$ get published; those with $p > .05$ often do not (**publication bias**)
  - Researchers have incentives to find $p < .05$ by any means necessary

---

# p-Hacking

**p-hacking** (also called "researcher degrees of freedom") refers to practices that inflate the false positive rate:

--

- Collecting data until $p < .05$, then stopping
- Trying many different outcome variables and only reporting the significant ones
- Trying many different model specifications and only reporting the best
- Removing outliers selectively to get $p < .05$
- Running many subgroup analyses and reporting only significant subgroups

--

Each of these can turn a 5% false positive rate into something much higher

---

# How Many Comparisons Are Too Many?

- If you run 20 independent tests, you expect **one** false positive at $\alpha = .05$ even if all nulls are true
--

- **Bonferroni correction**: divide $\alpha$ by the number of tests
  - For 20 tests: use $\alpha = .05/20 = .0025$
- Other corrections: Holm, Benjamini-Hochberg (FDR control)
--

- But corrections only work if you know (and report) how many tests you ran

---

# The Garden of Forking Paths

- Gelman & Loken (2014): even without deliberate p-hacking, researchers face many implicit decisions
- Each decision is a "fork" in the analysis path
- The chosen path is usually the one that looks most promising
--

- This inflates false positive rates even without any conscious manipulation
- Pre-registration addresses this by committing to analysis choices before seeing data

---

# Publication Bias

- Studies with $p < .05$ are more likely to be published
- Studies with $p > .05$ sit in file drawers
--

- The published literature overestimates effect sizes and underestimates false positive rates
- **Funnel plot asymmetry** in meta-analyses is a signature of publication bias

<!-- TODO: add funnel plot asymmetry image -->

---

class: middle, center, inverse

# The Replication Crisis

---

# The Replication Crisis

- Starting around 2011, researchers began systematically attempting to replicate published findings
- Results were alarming:
--

  - **Reproducibility Project: Psychology** (Open Science Collaboration, 2015): only ~36–39% of 100 psychology studies replicated
  - **Reproducibility Project: Cancer Biology**: similar story in pre-clinical cancer research
  - Many high-profile findings in social psychology, medicine, and economics failed to replicate

---

# Why Did This Happen?

- **Low statistical power** → inflated effect size estimates when significant
- **Publication bias** → file drawer problem
- **p-hacking** → inflated false positive rates
- **HARKing** (Hypothesizing After Results are Known) → post-hoc hypotheses presented as a priori
- **Small samples** → noisy estimates
- **Flexible analysis choices** → garden of forking paths

---

# The Winner's Curse

- When effect sizes are estimated with noise and only significant results are published, the published estimates are **biased upward**
- This is the **winner's curse**: the "winners" (significant results) win in part because of sampling error in the favorable direction
--

- Implication: expect replication estimates to be smaller than original estimates, even when the true effect is real

<!-- TODO: add winner's curse / effect size inflation diagram -->

---

# What Changed?

In response to the replication crisis, the field has adopted reforms:

--

- **Pre-registration**: commit to hypotheses and analysis plan before data collection
- **Registered reports**: journals accept papers before results are known
- **Open data and materials**: allow others to verify and replicate
- **Larger samples and higher power**
- **Multi-site replications**
- **Adversarial collaborations**

---

class: middle, center, inverse

# Alternatives and Complements to p-values

---

# Why Look Beyond p-values?

- p-values answer a narrow question: how consistent are the data with $H_0$?
- They do not tell you:
  - How big is the effect?
  - How precise is the estimate?
  - What is the probability the hypothesis is true?
--

- Several complementary or alternative approaches address these limitations

---

# Effect Sizes

- Always report **effect sizes** alongside p-values
- Common measures:
  - Cohen's *d*: standardized mean difference
  - Pearson's *r*: correlation coefficient
  - Odds ratio, risk ratio
  - Regression coefficients (on interpretable scale)
--

- An effect can be statistically significant but **practically trivial**
- An effect can be non-significant but **practically important** (underpowered study)
- Effect sizes communicate magnitude, not just existence

---

# Confidence Intervals

- A **95% confidence interval** contains the true parameter value in 95% of repeated samples (frequentist interpretation)
--

- CIs convey both **magnitude** and **precision**
- A narrow CI around a small effect: precise null
- A wide CI spanning zero: uncertain, could be positive or negative
--

- Many statisticians recommend reporting CIs instead of (or alongside) p-values
- Note: a CI that excludes zero corresponds to $p < .05$ (for two-tailed tests)

---

# Common CI Misinterpretation

.pull-left[
**What a 95% CI is NOT:**
- The probability the true value lies in this specific interval is NOT 95%
- The interval either contains the true value or it doesn't (frequentist view)
- Not a Bayesian credible interval
]

.pull-right[
**What a 95% CI IS:**
- A procedure that, over repeated samples, produces intervals that cover the true value 95% of the time
- A useful summary of estimation uncertainty
- Informative about both direction and precision
]

---

# Bayesian Approaches

- **Bayesian inference** inverts the logic: it gives $P(H \mid \text{data})$, not $P(\text{data} \mid H)$
- Requires specifying a **prior** distribution over hypotheses/parameters
--

- **Bayes factors**: ratio of marginal likelihoods under two hypotheses
  - $BF_{10} > 1$: data favor $H_1$ over $H_0$
  - $BF_{10} < 1$: data favor $H_0$ over $H_1$
  - Provides evidence on a continuous scale, including evidence for $H_0$
--

- Bayesian **credible intervals**: the probability the parameter lies in this range is X% (given the prior and data)

---

# The American Statistical Association on p-values

The ASA issued a statement in 2016 (Wasserstein & Lazar) with six principles:

--

1. p-values can indicate how incompatible the data are with a specified statistical model
2. p-values do **not** measure the probability that the studied hypothesis is true
3. Scientific conclusions should **not** be based only on whether a p-value passes a threshold
4. Proper inference requires **full reporting and transparency**
5. A p-value does **not** measure the size of an effect or the importance of a result
6. By itself, a p-value does **not** provide a good measure of evidence regarding a model or hypothesis

---

# "Retire Statistical Significance"?

- Amrhein, Greenland, & McShane (2019) in *Nature*: abandon the term "statistically significant" entirely
- Their argument: the binary significant/not-significant distinction causes more harm than good
--

- Instead: report and interpret effect sizes and CIs continuously
- Do not use "significant" as a synonym for "true" or "important"
--

- This view is controversial — others argue we need some threshold for decision-making
- The debate continues

---

class: middle, center, inverse

# Best Practices

---

# Best Practices for Reporting

- Report **effect sizes** with confidence intervals, not just p-values
- Use p-values as one input among many, not the sole arbiter
- Be transparent about all tests conducted (avoid selective reporting)
- Pre-register your hypotheses and analysis plan
- Report exact p-values (e.g., $p = .032$) rather than just $p < .05$
- Interpret non-significant results carefully — distinguish "no evidence of effect" from "evidence of no effect"

---

# Best Practices for Interpretation

- **Do not** dichotomize into "significant" vs. "not significant"
- **Do** think about the full distribution of plausible effect sizes
- **Do** consider power: a non-significant result from an underpowered study is uninformative
- **Do** consider practical significance alongside statistical significance
- **Do** triangulate across multiple studies and methods

---

# The New Statistics

Cumming (2014) advocates for the **"new statistics"**:

- Focus on **estimation** rather than binary hypothesis testing
- Use **effect sizes and confidence intervals** as the primary summaries
- Think in terms of **meta-analytic thinking** — each study contributes to a cumulative body of evidence
- Embrace **replication** as the gold standard

<!-- TODO: add "new statistics" estimation approach diagram -->

---

# p-values in Context: A Summary

.pull-left[
**p-values are useful when:**
- Used as one piece of evidence
- Reported alongside effect sizes and CIs
- Interpreted with attention to power
- Part of a pre-registered analysis
- Combined with replication evidence
]

.pull-right[
**p-values are misleading when:**
- Used as the sole criterion for truth
- Treated as a binary pass/fail
- The result of p-hacking or selective reporting
- Interpreted as $P(H_0 \mid \text{data})$
- Used to ignore effect sizes and precision
]

---

class: middle, center, inverse

# Wrapping Up...

---

# Summary

- A **p-value** is $P(\text{data this extreme} \mid H_0)$ — not the probability $H_0$ is true
- Common misinterpretations include treating $p < .05$ as proof and $p > .05$ as evidence of no effect
- **p-hacking** and publication bias inflate false positive rates in the literature
- The **replication crisis** revealed serious problems with standard NHST practices
- **Effect sizes and confidence intervals** are essential complements to p-values
- **Bayesian approaches** offer an alternative framework
- Best practice: pre-register, report fully, interpret carefully, and triangulate across evidence

---

# References

- Amrhein, V., Greenland, S., & McShane, B. (2019). Scientists rise up against statistical significance. *Nature*, 567, 305–307.
- Cumming, G. (2014). The new statistics: Why and how. *Psychological Science*, 25(1), 7–29.
- Gelman, A., & Loken, E. (2014). The statistical crisis in science. *American Scientist*, 102(6), 460.
- Ioannidis, J. P. A. (2005). Why most published research findings are false. *PLOS Medicine*, 2(8), e124.
- Open Science Collaboration. (2015). Estimating the reproducibility of psychological science. *Science*, 349(6251), aac4716.
- Wasserstein, R. L., & Lazar, N. A. (2016). The ASA statement on p-values: Context, process, and purpose. *The American Statistician*, 70(2), 129–133.
