---
title: "Cumulative Distributions: Worked Z-Score Examples"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# Remember to compile
#xaringan::inf_mr(cast_from = "..")
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Cumulative Distributions

### Worked Z-Score Problem

---

## Cumulative Proportions

<!-- TODO: add image of standard normal distribution table showing cumulative probability P[Z < z] -->

The cumulative proportion for a value $x$ in a distribution is the proportion of observations in the distribution that are less than or equal to $Z$.

---

## Worked Z-Score Problem

<!-- TODO: add image of standard normal table with z=0.0 row highlighted (showing 0.5000), and a red dashed box around the upper right region of the distribution curve -->

---

## Worked Z-Score Problem

<!-- TODO: add image of standard normal table with z=1.0 row highlighted (showing 0.8413 at column .00), and a blue box around upper right region of curve -->

---

## Worked Z-Score Problem

<!-- TODO: add image of standard normal table with z=1.0 and column .00 highlighted showing 0.8413, with red dashed vertical line at mean and blue vertical line to the right -->

$.8413 - .500 = .3413$

---

## Worked Z-Score Problem

<!-- TODO: add image of standard normal table with both sides of distribution marked, showing symmetric region from -1 SD to +1 SD -->

$.8413 - .500 = .3413 \times 2 = .68$

---

## Worked Z-Score Problem

**B) What would these proportions be in an exactly Normal distribution?**

$\pm$ **TWO SD**

<!-- TODO: add image of standard normal distribution table (upper portion z = 0.0 to 0.4) with red dashed box highlighting P[Z < z] area -->

---

## Worked Z-Score Problem

<!-- TODO: add image of standard normal table with z=2.0 row highlighted (showing 0.9773 at column .00), and red/blue vertical lines on the distribution curve -->

---

## Worked Z-Score Problem

<!-- TODO: add image of standard normal table (lower portion showing z = 1.5 to 2.1) with z=2.0 value 0.9773 highlighted in blue, and red dashed box + blue box on distribution curve -->

$.9773 - .500 = .4773$

---

## Worked Z-Score Problem

<!-- TODO: add image of standard normal table (lower portion) with symmetric region marked, showing -2 SD to +2 SD span -->

$.9773 - .500 = .4773 \times 2 = .9546$

---

## Starting from a Proportion

- SAT reading scores for a recent year are distributed according to an $N(504, 111)$ distribution.

- How high must a student score in order to be in the **top 10%** of the distribution?

- In order to use Table A, equivalently, what score has cumulative proportion **0.90** below it?

<!-- TODO: add image of normal curve N(504, 111) with 0.90 shaded area below cutoff and 0.10 in the right tail, with 504 and ? marked on x-axis -->

---

## Normal Calculations (2 of 3)

How high must a student score in order to be in the top 10% of the distribution?

.pull-left[
- Look up the closest probability (closest to 0.90) in the table.
- Find the corresponding **standardized score**.
- The value you seek is *that many standard deviations from the mean*.

$$z = 1.28$$
]

.pull-right[
<!-- TODO: add image of z-table excerpt showing z=1.2 row with .8997 circled at column .08, arrow pointing left to row 1.2 -->

| z | .07 | .08 | .09 |
|---|-----|-----|-----|
| 1.1 | .8790 | .8810 | .8830 |
| 1.2 | .8980 | **.8997** | .9015 |
| 1.3 | .9147 | .9162 | .9177 |

]

---

## Normal Calculations (3 of 3)

How high must a student score in order to be in the top 10% of the distribution?

$$z = 1.28$$

We need to "unstandardize" the z-score to find the observed value (x):

$$z = \frac{x - \mu}{\sigma} \implies x = \mu + z\sigma$$

$$x = 504 + z \cdot 111$$
$$= 504 + (1.28 \cdot 111)$$
$$= 504 + 142.08 = 646.08$$

A student would have to score at least **646.08** to be in the top 10% of the distribution of SAT reading scores for this particular year.

---

## "Backward" Normal Calculations

**USING TABLE A GIVEN A NORMAL PROPORTION**

**STEP 1:** **State the problem** in terms of the given proportion. **Draw a picture** that shows the Normal value, $x$, that you want in relation to the cumulative proportion.

**STEP 2:** **Use Table**, the fact that the total area under the curve is 1, and the given area under the standard Normal curve to find the corresponding $z$-value.

**STEP 3:** **Unstandardize** $z$ to solve the problem in terms of a non-standard Normal variable $x$.

---

## Functions in R

`dnorm()` gives the density,

`pnorm()` gives the cumulative density function,
- computes the probability that a normally distributed random number will be less than that number

`qnorm()` gives the quantile function, and
- (is the inverse of pnorm, give it a probability,
- it produces the number whose cumulative distribution matches the probability

`rnorm()` generates random deviates.

---

## Illustrations with R

- Throwback: [Sampling and Standard Error](https://link)
- [Robustness](https://link)

---

class: middle, center, inverse

# Wrapping Up...
