---
title: "Robustness"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

class: middle, center, inverse

# Robustness

---

# Roadmap

- What is robustness?
- Robustness checks in practice
- Types of robustness checks
  - Alternative model specifications
  - Alternative samples
  - Alternative measures
  - Placebo tests
- How to present robustness checks

--

**By the end of this lecture, you should be able to:**

- Define robustness and explain why it matters for causal inference
- Identify and implement common types of robustness checks
- Distinguish between robustness checks and specification searches
- Critically evaluate robustness evidence in published research
- Present robustness results clearly and honestly

---

class: middle, center, inverse

# What is Robustness?

---

# What is Robustness?

- A finding is **robust** if it holds up under alternative reasonable assumptions
--

- Robustness is about **credibility**
  - Does your result depend critically on one specific choice you made?
  - Or does it survive when you make different defensible choices?
--

- Robustness checks are **not** the same as p-hacking or specification searching
  - You are not looking for the specification that gives the result you want
  - You are checking whether your result is sensitive to reasonable alternative choices

---

# Why Do Robustness Checks Matter?

- Every empirical analysis requires **researcher degrees of freedom**
  - Which sample to use
  - Which controls to include
  - How to measure key variables
  - Which functional form to use
--

- These choices can matter — a lot
--

- Robustness checks help you (and readers) understand which choices drive the result
--

- A result that only holds for one very specific set of choices is **fragile**
- A result that holds across many reasonable alternatives is **robust**

---

class: middle, center, inverse

# Types of Robustness Checks

---

# Types of Robustness Checks

There are several broad categories:

--

1. **Alternative model specifications**
   - Adding or removing control variables
   - Using different functional forms

--

2. **Alternative samples**
   - Different time periods
   - Different geographic areas
   - Dropping outliers or influential observations

--

3. **Alternative measures**
   - Different ways of operationalizing your key variables

--

4. **Placebo tests**
   - Tests where you should find *no* effect if your identification is valid

---

class: middle, center, inverse

# Alternative Model Specifications

---

# Alternative Model Specifications

.pull-left[
**Adding/removing controls**
- Does the result change when you add more covariates?
- Does it change when you remove some?
- A robust result should not be very sensitive to the inclusion of plausible controls
]

.pull-right[
**Functional form**
- Linear vs. log-linear
- Quadratic terms
- Interactions
- Does the qualitative finding survive different functional form assumptions?
]

---

# How Sensitive Is Your Estimate?

- Report estimates across a **range** of specifications
- Show a coefficient plot or table with multiple columns
- Be transparent about what changes and what stays the same

--

- Key question: Is the change in the estimate **substantively** meaningful?
  - A coefficient that goes from 0.42 to 0.38 is probably not a concern
  - A coefficient that goes from 0.42 to -0.10 is a serious concern

---

# Oster (2019): Coefficient Stability

- **Oster (2019)** provides a formal framework for thinking about omitted variable bias
- The key insight: if adding observed controls moves the estimate a lot, unobserved controls might matter too
--

- The **delta** statistic asks: how much stronger would selection on unobservables need to be (relative to observables) to explain away the result?
--

- A result with a very small delta is fragile
- A result with a large delta is more credible

<!-- TODO: add Oster coefficient stability diagram/formula image -->

---

class: middle, center, inverse

# Alternative Samples

---

# Alternative Samples

**Why vary the sample?**

- Your main sample may have specific features that drive the result
- Checking other samples tests generalizability

--

**Common approaches:**
- Drop a specific year, region, or group
- Restrict to a subset (e.g., only large firms, only certain demographics)
- Extend the time period
- Use a different dataset covering the same phenomenon

---

# Outliers and Influential Observations

- A single observation (or small group) can sometimes drive a result
--

- **Checks:**
  - Drop the top/bottom 1% or 5% of the outcome or key variable
  - Run Cook's distance or leverage diagnostics
  - Jackknife: re-run dropping each observation one at a time (or each cluster)
--

- If your result disappears when you drop a handful of observations, that is important to report honestly

---

# Geographic and Temporal Robustness

.pull-left[
**Geographic**
- Does the result hold in different countries or regions?
- If you use US data, does it replicate in Europe or Asia?
- Cross-country variation can reveal whether institutional context matters
]

.pull-right[
**Temporal**
- Does the result hold in different time periods?
- Pre- vs. post-policy change
- Different economic cycles (boom vs. recession)
- Stability over time is reassuring
]

---

class: middle, center, inverse

# Alternative Measures

---

# Alternative Measures

- Variables are often **proxies** for underlying concepts
- Different proxies may capture different aspects of the concept
--

**Examples:**
- Income: wages vs. household income vs. consumption
- Education: years of schooling vs. test scores vs. credentials
- Health: self-reported vs. clinical measures vs. mortality
--

- If your result holds with multiple different operationalizations, it is more credible
- If it only holds for one specific measure, ask: what is special about that measure?

---

# Measurement Error

- All measures have error
- Classical measurement error in the independent variable **attenuates** estimates toward zero
--

- Robustness checks with alternative measures can also serve as a check on measurement error
  - If two noisy measures give similar results, the true effect is likely real
  - If they diverge sharply, understand why

---

class: middle, center, inverse

# Placebo Tests

---

# Placebo Tests

- A **placebo test** checks for an effect where you should theoretically find *none*
- Named by analogy to medical placebo trials
--

**Why are they useful?**
- They test whether your identification strategy is actually working
- A "false positive" on a placebo test undermines your main result

---

# Types of Placebo Tests

.pull-left[
**Outcome placebo**
- Use a different outcome that *should not* be affected by your treatment
- If you find an effect on the placebo outcome, something may be wrong with your design
]

.pull-right[
**Treatment placebo**
- Use a "fake" treatment that *should not* have an effect
- Pre-treatment assignment, randomly shuffled treatment, future treatment
]

---

# Placebo Tests: Example

**Example: Difference-in-differences**

- Main test: Did treated units' outcomes change after treatment relative to controls?
--

- Placebo test 1: Did treated units' outcomes change *before* treatment (pre-trend test)?
  - If yes: parallel trends assumption may be violated
--

- Placebo test 2: Assign fake treatment dates and re-run the analysis
  - If you consistently find large "effects" at fake dates, your result may be spurious

---

# Permutation / Randomization Inference

- A related idea: **permutation tests**
- Randomly re-assign treatment many times and re-estimate the effect each time
- Build a distribution of "effects" under the null
--

- If your actual estimate is in the tail of this null distribution, that provides evidence it is not a fluke
- Particularly useful when cluster sizes are small and asymptotic inference may be unreliable

---

class: middle, center, inverse

# Robustness vs. Specification Search

---

# Robustness Checks vs. Specification Search

These look similar but are fundamentally different:

.pull-left[
**Robustness check `r emo::ji("check")`**
- Pre-specified or motivated by theory
- You report all results honestly, including null or contradictory ones
- The goal is to test sensitivity to reasonable alternatives
]

.pull-right[
**Specification search `r emo::ji("x")`**
- Motivated by finding a significant or large result
- Selective reporting of favorable specifications
- Inflates false positive rates
- Sometimes called p-hacking or researcher degrees of freedom abuse
]

---

# How to Be Honest About Robustness

- **Pre-registration**: commit to your main specification and key robustness checks before seeing data
- Report the full set of results, not just the ones you like
- Use **specification curves** or **multiverse analysis** to summarize many specifications systematically
--

- The goal is to give readers an accurate picture of how sensitive your result is — not to cherry-pick

---

# Specification Curve Analysis

- **Simonsohn, Simmons, and Nelson (2020)** propose the **specification curve**
- Run all defensible specifications and plot the distribution of estimates
--

- Shows:
  - What fraction of specifications give a significant result?
  - What is the typical effect size across specifications?
  - Are there specific choices that systematically drive the result?

<!-- TODO: add specification curve example image -->

---

# Multiverse Analysis

- Related concept: **multiverse analysis** (Steegen et al., 2016)
- Define the full "multiverse" of reasonable analytical choices
- Run the analysis for every combination
--

- Summarize results as a distribution rather than a single point estimate
- Helps identify which choices matter most for the result

---

class: middle, center, inverse

# How to Present Robustness Checks

---

# How to Present Robustness Checks

**In a table:**
- Each column = one specification or sample restriction
- Rows = key coefficients
- Note what changes across columns
- Highlight the main specification

--

**In a figure:**
- Coefficient plot showing estimates and confidence intervals across specifications
- Easy to see how point estimates and precision change

---

# What to Report

- **Always** report your main specification clearly
- Show how the estimate changes across alternative specifications
- Be explicit about what you are varying and why
- Discuss results honestly — including cases where robustness checks give weaker results
--

- A paper with honest, comprehensive robustness checks is more credible than one with a few cherry-picked favorable checks

---

# Common Mistakes

.pull-left[
- Running robustness checks only after seeing results
- Only reporting checks that confirm the main result
- Treating a robustness check table as a formality rather than a genuine test
- Calling a non-robust result "robust" because most specs are significant
]

.pull-right[
- Forgetting to think about *why* a check fails
  - Understanding when and why robustness fails is often informative
- Conflating statistical significance with practical robustness
- Ignoring the magnitude of changes across specifications
]

---

# Robustness in Practice: Summary

- Robustness checks test whether your result depends on specific choices
- Key types: alternative specifications, samples, measures, and placebo tests
- Be honest: report all results, not just favorable ones
- Pre-registration and specification curves help guard against selective reporting
- A robust result survives across a wide range of reasonable alternatives

---

class: middle, center, inverse

# Wrapping Up...

---

# Summary

- **Robustness** means a finding holds under alternative reasonable assumptions
- Four main types of robustness checks:
  - Alternative model specifications
  - Alternative samples
  - Alternative measures
  - Placebo tests
- **Robustness checks ≠ specification searching** — honesty and transparency matter
- Tools like specification curves and multiverse analysis help summarize many specifications
- Present robustness results fully and honestly to build credibility

---

# Resources

- Oster, E. (2019). Unobservable selection and coefficient stability: Theory and evidence. *Journal of Business & Economic Statistics*, 37(2), 187–204.
- Simonsohn, U., Simmons, J. P., & Nelson, L. D. (2020). Specification curve analysis. *Nature Human Behaviour*, 4(11), 1208–1214.
- Steegen, S., Tuerlinckx, F., Gelman, A., & Vanpaemel, W. (2016). Increasing transparency through a multiverse analysis. *Perspectives on Psychological Science*, 11(5), 702–712.

class: middle, center, inverse

# Wrapping Up...
