---
title: "Research Designs"
author: "S. Mason Garrison"
bibliography: "library.bib"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    self_contained: TRUE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
      slideNumberFormat: ""
---

```{r child = "../setup.Rmd"}
```
```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
if (!require("emo")) devtools::install_github("hadley/emo")
library(emo)
knitr::opts_chunk$set(echo = FALSE, out.width = "90%", fig.align = "center")
```

---
class: middle, center, inverse

# Research Designs

---

# Roadmap

- Why research design matters
- Types of research designs
  - Observational studies
  - Experimental studies
- Key concepts: confounding, control, randomization
- Specific design types: cross-sectional, longitudinal, case-control, cohort
- Causal inference and its limits

--

**By the end of this lecture, you should be able to:**

- Distinguish between observational and experimental research designs
- Identify the key features and trade-offs of major design types
- Explain why randomization is critical for causal inference
- Recognize potential sources of confounding in a study
- Evaluate the strengths and limitations of a given research design

---
class: middle, center, inverse

# Why Research Design Matters

---

# The Goal of Research

The fundamental goal of empirical research is to answer questions about the world:

- Does this drug reduce blood pressure?
- Does class size affect student achievement?
- Do social media apps increase anxiety?

--

The **research design** determines whether we can actually answer these questions validly.

---

# Correlation vs. Causation

.pull-left[
**"Correlation does not imply causation"**

Just because two variables move together does not mean one causes the other.

- Ice cream sales and drowning rates are correlated
- Shoe size and reading ability are correlated in children
- Nicolas Cage movies and pool drownings...
]

.pull-right[
<!-- TODO: add spurious correlation example image -->
]

--

To establish **causation**, we need more than correlation — we need a research design that rules out alternative explanations.

---

# Three Requirements for Causation

To conclude that $X$ causes $Y$, we need:

1. **Covariation**: $X$ and $Y$ are statistically associated
2. **Temporal precedence**: $X$ comes before $Y$ in time
3. **Elimination of alternatives**: No other variable explains the relationship

--

Research designs vary in how well they satisfy all three requirements.

---
class: middle, center, inverse

# Types of Research Designs

---

# The Big Picture

```
Research Designs
├── Observational Studies
│   ├── Cross-sectional
│   ├── Case-control
│   ├── Cohort (prospective / retrospective)
│   └── Ecological
└── Experimental Studies
    ├── Randomized Controlled Trial (RCT)
    ├── Quasi-experiment
    └── Natural experiment
```

--

The key distinction: **Who controls the assignment** to conditions?

---

# Observational vs. Experimental

.pull-left[
**Observational Studies**

- Researcher observes but does **not** manipulate
- Participants are in conditions naturally
- Can establish association
- Weaker for causal claims

*Examples:* surveys, medical records, census data
]

.pull-right[
**Experimental Studies**

- Researcher **manipulates** the independent variable
- Participants are assigned to conditions
- With randomization, supports causal claims

*Examples:* clinical trials, lab experiments, A/B tests
]

---
class: middle, center, inverse

# Observational Designs

---

# Cross-Sectional Design

**Definition:** Data collected from a population at a **single point in time**

<!-- TODO: add cross-sectional design diagram image -->

--

.pull-left[
**Strengths:**
- Fast and inexpensive
- Good for prevalence estimation
- Can study many variables at once
]

.pull-right[
**Weaknesses:**
- Cannot establish temporal order
- Cannot track change over time
- Cohort effects can confound age effects
]

--

**Example:** Surveying 1,000 adults today about their sleep habits and mental health

---

# Longitudinal Design

**Definition:** The same participants are measured **multiple times** over a period

<!-- TODO: add longitudinal design timeline diagram image -->

--

.pull-left[
**Strengths:**
- Tracks change over time
- Establishes temporal order
- Can separate age, cohort, and period effects
]

.pull-right[
**Weaknesses:**
- Expensive and time-consuming
- Attrition (dropout) is a major problem
- Practice effects from repeated measurement
]

--

**Example:** Following 500 children from age 5 to 18, measuring vocabulary and reading each year

---

# Cohort Study

**Definition:** A group sharing a common characteristic (the **cohort**) is followed over time

- **Prospective cohort:** Start now, follow forward
- **Retrospective cohort:** Use existing records to look back

--

**Example:** The Nurses' Health Study — followed ~120,000 nurses over decades to study diet, lifestyle, and disease

--

.pull-left[
**Strengths:**
- Can study rare exposures
- Establishes temporal sequence
- Multiple outcomes can be studied
]

.pull-right[
**Weaknesses:**
- Expensive and slow (prospective)
- Attrition and loss to follow-up
- Confounding is a serious concern
]

---

# Case-Control Design

**Definition:** Start with **outcomes** and look backward for exposures

- **Cases:** People who have the outcome (e.g., lung cancer)
- **Controls:** People who do not have the outcome, matched to cases

<!-- TODO: add case-control design diagram image -->

--

.pull-left[
**Strengths:**
- Efficient for rare outcomes
- Relatively quick and cheap
- Good for disease etiology studies
]

.pull-right[
**Weaknesses:**
- Recall bias (cases remember differently)
- Difficult to select good controls
- Can only study one outcome at a time
- Cannot calculate incidence rates
]

---

# Comparing Observational Designs

| Design | Time Direction | Rare Outcome | Rare Exposure | Causal Strength |
|--------|---------------|--------------|---------------|-----------------|
| Cross-sectional | Snapshot | OK | OK | Weak |
| Cohort (prospective) | Forward | Poor | Good | Moderate |
| Cohort (retrospective) | Backward | OK | Good | Moderate |
| Case-control | Backward | Good | Poor | Moderate |

---

# Confounding

A **confounder** is a variable that:
1. Is associated with the **exposure**
2. Is associated with the **outcome**
3. Is **not** on the causal pathway between exposure and outcome

<!-- TODO: add confounding DAG diagram image -->

--

**Classic example:** Coffee drinking and lung cancer
- Coffee drinkers were more likely to smoke
- Smoking causes lung cancer
- Smoking is a confounder of the coffee–cancer relationship

---

# Dealing with Confounding (Observational)

Strategies to control confounding in observational studies:

--

- **Restriction:** Only include participants without the confounding variable
- **Matching:** Match cases to controls on the confounder
- **Stratification:** Analyze separately within levels of the confounder
- **Statistical adjustment:** Include the confounder in regression models
- **Propensity scores:** Model probability of exposure, then adjust

--

None of these are as clean as **randomization**.

---
class: middle, center, inverse

# Experimental Designs

---

# The Randomized Controlled Trial (RCT)

**The gold standard for causal inference**

Key features:
1. Researcher **manipulates** the independent variable (treatment)
2. Participants are **randomly assigned** to treatment or control
3. Outcomes are measured and compared

<!-- TODO: add RCT flow diagram image -->

---

# Why Randomization Works

Randomization ensures that, on average, the treatment and control groups are **equivalent on all variables** — both measured and unmeasured.

--

This means any difference in outcomes can be attributed to the treatment, not to pre-existing differences between groups.

--

**Analogy:** Randomly shuffling a deck of cards ensures neither player systematically gets better cards.

---

# The Anatomy of an RCT

```
Population
    │
    ▼
Eligible Sample
    │
    ▼ (random assignment)
┌───────────┐     ┌──────────────┐
│ Treatment │     │   Control    │
│  Group    │     │   Group      │
└───────────┘     └──────────────┘
    │                   │
    ▼                   ▼
 Outcome             Outcome
    │                   │
    └─────── Compare ───┘
```

---

# Control Groups

A **control group** is the comparison condition:

- **No treatment:** Participants receive nothing
- **Placebo:** Participants receive an inert treatment (controls for expectation effects)
- **Active control:** Participants receive the current best treatment (ethical comparison)
- **Waitlist control:** Participants receive treatment later

--

The choice of control group affects both validity and ethics.

---

# Blinding

**Blinding** prevents knowledge of condition assignment from influencing outcomes:

- **Single-blind:** Participants don't know their condition
- **Double-blind:** Neither participants nor outcome assessors know conditions
- **Triple-blind:** Participants, assessors, and analysts are all blinded

--

Why does this matter? **Placebo effects** and **expectancy effects** can produce real changes in outcomes even when treatment is inert.

---

# Internal vs. External Validity

.pull-left[
**Internal Validity**

"Did the treatment cause the outcome *in this study*?"

Threats:
- Confounding
- Selection bias
- Attrition
- Testing effects
- History effects
]

.pull-right[
**External Validity**

"Do the findings generalize beyond this study?"

Threats:
- Unrepresentative sample
- Artificial lab setting
- Demand characteristics
- Limited range of conditions
]

--

RCTs maximize internal validity but often sacrifice external validity.

---

# Quasi-Experiments

**Definition:** Studies that resemble experiments but lack full random assignment

--

**Examples:**
- **Regression discontinuity:** Students above/below a test score cutoff get different programs
- **Difference-in-differences:** Compare changes over time in treatment vs. control groups
- **Instrumental variables:** Use a third variable to isolate causal effect

--

Quasi-experiments exploit **natural variation** that approximates random assignment.

---

# Natural Experiments

**Definition:** Cases where assignment to conditions occurs through a process that is **as-if random** — not controlled by the researcher

--

**Famous examples:**
- Vietnam War draft lottery → effect of military service on earnings
- Flood/disaster proximity → effect of trauma on mental health
- Twin studies → separating genetic and environmental effects

--

These allow causal inference in situations where true experiments are impossible or unethical.

---
class: middle, center, inverse

# Threats to Validity

---

# Selection Bias

**Selection bias** occurs when the people who end up in one condition are systematically different from those in another.

--

**Examples:**
- Volunteers for a drug trial may be healthier than non-volunteers
- Students who take tutoring may be more motivated
- Customers who respond to a survey may be more satisfied (or dissatisfied)

--

Randomization eliminates selection bias in expectation.

---

# Attrition Bias

**Attrition** (dropout) becomes a problem when the people who drop out are different from those who stay.

--

**Example:** In a weight loss intervention, people who gain weight drop out. The remaining participants look like they maintained their weight — but this is misleading.

--

**Intention-to-treat (ITT) analysis:** Analyze all participants as originally assigned, regardless of whether they completed the intervention.

---

# Measurement Bias

Systematic error in how outcomes are measured:

- **Social desirability bias:** Participants give answers they think are expected
- **Recall bias:** Cases remember past exposures differently than controls
- **Observer bias:** Assessors rate outcomes differently based on known condition

--

Solutions: blinding, standardized measurement protocols, objective outcomes (e.g., biomarkers rather than self-report)

---

# Hawthorne Effect

**Definition:** Participants change their behavior simply because they know they are being observed — regardless of the treatment

--

Named after a 1920s–30s study at the Hawthorne Works factory, where lighting changes seemed to improve productivity — but the real cause was being watched.

--

**Implication:** Control groups help, but if everyone knows they're in a study, all groups may be affected.

---
class: middle, center, inverse

# Choosing a Design

---

# The Research Question Determines the Design

| Question Type | Best Design |
|---------------|-------------|
| How common is X? | Cross-sectional survey |
| What predicts X over time? | Longitudinal / cohort |
| What causes X (ethically manipulable)? | RCT |
| What causes X (not manipulable)? | Quasi-experiment / natural experiment |
| What factors led to rare outcome X? | Case-control |

---

# Practical Constraints

The ideal design is rarely the feasible design. Trade-offs include:

.pull-left[
**Resources:**
- Time
- Money
- Sample size availability

**Ethics:**
- Cannot randomize to harmful conditions
- Cannot withhold effective treatments
]

.pull-right[
**Logistics:**
- Participant recruitment
- Retention over time
- Measurement feasibility

**Context:**
- Population of interest
- Setting generalizability
]

---

# The Hierarchy of Evidence

```
Systematic Reviews & Meta-analyses
          ↑
   Randomized Controlled Trials
          ↑
     Quasi-experiments
          ↑
    Cohort Studies
          ↑
  Case-Control Studies
          ↑
  Cross-Sectional Studies
          ↑
 Case Reports / Expert Opinion
```

--

Higher = stronger causal evidence, but also typically more resource-intensive.

---

# A Design Checklist

When evaluating a study's design, ask:

- `r emo::ji("check")` What is the research question?
- `r emo::ji("check")` Was there random assignment? If not, what controls confounding?
- `r emo::ji("check")` What is the comparison group?
- `r emo::ji("check")` Was blinding used?
- `r emo::ji("check")` What is the attrition rate?
- `r emo::ji("check")` Is the sample representative of the population of interest?
- `r emo::ji("check")` What are the threats to internal and external validity?

---

# Case Study: Does Exercise Reduce Depression?

Let's apply our design knowledge to a concrete question: **Does exercise reduce depression?**

--

- **Cross-sectional:** People who exercise more report less depression — but healthy people may choose to exercise
- **Cohort study:** Non-depressed exercisers followed over time — but selection still a concern
- **Case-control:** Depressed vs. non-depressed, compare past exercise — recall bias problem
- **RCT:** Randomly assign depressed individuals to exercise vs. control — strongest evidence

--

Answer: Yes — multiple RCTs confirm exercise reduces depression symptoms `r emo::ji("muscle")`

---
class: middle, center, inverse

# Wrapping Up...

---

# What We Covered

- **Observational designs:** Cross-sectional, cohort, case-control — describe and associate but struggle with causation
- **Experimental designs:** RCTs with randomization — the gold standard for causal inference
- **Quasi-experiments and natural experiments:** Clever approximations when true experiments are impossible
- **Threats to validity:** Confounding, selection bias, attrition, measurement bias, Hawthorne effect
- **Design selection:** Driven by research question, ethics, and practical constraints

--

Good research design is the foundation of valid conclusions — no amount of sophisticated analysis can fully rescue a poorly designed study.
